### 2015-03-12 paper.tex
Add paper based on AIP template for JRSE

%% ****** Start of file aiptemplate.tex ****** %
%%
%%   This file is part of the files in the distribution of AIP substyles for
% %  REVTeX4.
%%   Version 4.1 of 9 October 2009.
%%
%
% This is a template for producing documents for use with
% the REVTEX 4.1 document class and the AIP substyles.
%
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.

\documentclass[aip,graphicx]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

\draft % marks overfull lines with a black rule on the right

\begin{document}

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{Modeling vertical-axis cross-flow turbine wakes} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{Peter Bachant}
\email[]{pxL3@unh.edu}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{Center for Ocean Renewable Energy, University of New Hampshire, Durham, NH}

\author{Martin Wosnik}
%\email[]{}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{Center for Ocean Renewable Energy, University of New Hampshire, Durham, NH}

% Collaboration name, if desired (requires use of superscriptaddress option in
% \documentclass).
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
% insert abstract here
\end{abstract}

\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands
\section{}
%\label{}
\subsection{}
\subsubsection{}

% If in two-column mode, this environment will change to single-column format so that long equations can be displayed. 
% Use only when necessary.
%\begin{widetext}
%$$\mbox{put long equation here}$$
%\end{widetext}

% Figures should be put into the text as floats. 
% Use the graphics or graphicx packages (distributed with LaTeX2e).
% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz, and Frank Mittelbach for examples. 
%
% Here is an example of the general form of a figure:
% Fill in the caption in the braces of the \caption{} command. 
% Put the label that you will use with \ref{} command in the braces of the \label{} command.
%
% \begin{figure}
% \includegraphics{}%
% \caption{\label{}}%
% \end{figure}

% Tables may be be put in the text as floats.
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
%
% \begin{table}
% \caption{\label{} }
% \begin{tabular}{}
% \end{tabular}
% \end{table}

% If you have acknowledgments, this puts in the proper section head.
%\begin{acknowledgments}
% Put your acknowledgments here.
%\end{acknowledgments}

% Create the reference section using BibTeX:
\bibliography{your-bib-file}

\end{document}
%
% ****** End of file aiptemplate.tex ******

### 2015-03-12 paper.tex
Minor additions and removing template comments

@@
 
-\title{Modeling vertical-axis cross-flow turbine wakes} %Title of paper
+\title{Modeling vertical-axis cross-flow turbine wakes}
 
@@
 %\altaffiliation{}
-\affiliation{Center for Ocean Renewable Energy, University of New Hampshire, Durham, NH}
+\affiliation{Center for Ocean Renewable Energy, University of New Hampshire, 
+Durham, NH}
 
@@
 %\altaffiliation{}
-\affiliation{Center for Ocean Renewable Energy, University of New Hampshire, Durham, NH}
+\affiliation{Center for Ocean Renewable Energy, University of New Hampshire,
+Durham, NH}
 
@@
 \begin{abstract}
-% insert abstract here
+The wake of a vertical-axis cross-flow turbine (CFT) is modeled numerically via
+several techniques.
 \end{abstract}
@@
 
-% Body of paper goes here. Use proper sectioning commands. 
-% References should be done using the \cite, \ref, and \label commands
-\section{}
-%\label{}
-\subsection{}
-\subsubsection{}
+\section{Introduction}
 
-% If in two-column mode, this environment will change to single-column format so that long equations can be displayed. 
-% Use only when necessary.
-%\begin{widetext}
-%$$\mbox{put long equation here}$$
-%\end{widetext}
+Araya et al. modeled the flow through a VAT array using leaky Rankine bodies
+\cite{Araya2014}.
 
-% Figures should be put into the text as floats. 
-% Use the graphics or graphicx packages (distributed with LaTeX2e).
-% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz, and Frank Mittelbach for examples. 
-%
-% Here is an example of the general form of a figure:
-% Fill in the caption in the braces of the \caption{} command. 
-% Put the label that you will use with \ref{} command in the braces of the \label{} command.
-%
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}%
-% \end{figure}
+\section{Numerical models}
 
-% Tables may be be put in the text as floats.
-% Here is an example of the general form of a table:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
-% \begin{tabular}{} command.
-%
-% \begin{table}
-% \caption{\label{} }
-% \begin{tabular}{}
-% \end{tabular}
-% \end{table}
+\section{Model verification}
+
+\section{Model validation}
+
+Models were validated against several open datasets.
 
@@
 % Create the reference section using BibTeX:
-\bibliography{your-bib-file}
+\bibliography{library}
 
 \end{document}
-%
-% ****** End of file aiptemplate.tex ******

### 2015-03-12 paper.tex
Add more basic things

@@
 %\documentclass[aip,reprint]{revtex4-1}
+\usepackage[draft]{todonotes}
 
@@
 \author{Peter Bachant}
-\email[]{pxL3@unh.edu}
+\email[]{pxl3@unh.edu}
 %\homepage[]{Your web page}
@@
 \maketitle %\maketitle must follow title, authors, abstract and \pacs
+\listoftodos
 
@@
 
-% If you have acknowledgments, this puts in the proper section head.
-%\begin{acknowledgments}
-% Put your acknowledgments here.
-%\end{acknowledgments}
+\section{Results}
+
+\subsection{Performance prediction}
+
+Predictions for both the power and drag coefficients are shown in Figure~\ref{fig:perf_comp}
+
+\begin{figure}
+\caption{Power (a) and drag (b) coefficient predictions from experiments and
+each numerical model.}
+\label{fig:perf_comp}
+\end{figure}
+
+\todo[inline]{Add bar graph comparing various performance results for the RVAT.}
+
+\begin{acknowledgments}
+The authors acknowledge Vincent Neary and the Sandia National Laboratories Water
+Power Program for use of their Red Mesa high performance computing cluster.
+\end{acknowledgments}
 

### 2015-03-17 makefigs.py
Modify directories for Linux PC

@@ -7,12 +7,12 @@ from __future__ import print_function, division
 import matplotlib.pyplot as plt
 import os
 
-cfd_sst_dir = ""
-cfd_sa_dir = ""
-cfd_sst_2d_dir = ""
-cfd_sd_2d_dir = ""
+cfd_sst_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-3d/mesh14"
+cfd_sa_dir = "/media/pete/Data2/OpenFOAM/pete-2.3.x/run/unh-rvat-3d/mesh14-sa"
+cfd_sst_2d_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-2d/kOmegaSST"
+cfd_sd_2d_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-2d/SpalartAllmaras"
 cfd_alm_dir = ""
-exp_dir = "C:/Users/Pete/Research/Experiments/RVAT Re dep"
+exp_dir = "/home/pete/Google Drive/Research/Experiments/RVAT Re dep"
 paper_dir = os.getcwd()
 
 save = True

### 2015-03-17 makefigs.py
Now possible to plot 3-D CFD mean contours/quivers

@@ -10,10 +10,12 @@ import os
 cfd_sst_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-3d/mesh14"
 cfd_sa_dir = "/media/pete/Data2/OpenFOAM/pete-2.3.x/run/unh-rvat-3d/mesh14-sa"
 cfd_sst_2d_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-2d/kOmegaSST"
-cfd_sd_2d_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-2d/SpalartAllmaras"
+cfd_sa_2d_dir = "/media/pete/Data1/OpenFOAM/pete-2.3.x/run/unh-rvat-2d/SpalartAllmaras"
 cfd_alm_dir = ""
 exp_dir = "/home/pete/Google Drive/Research/Experiments/RVAT Re dep"
 paper_dir = os.getcwd()
+cfd_dirs = {"kOmegaSST": cfd_sst_dir, "SpalartAllmaras": cfd_sa_dir}
+cfd_dirs_2d = {"kOmegaSST": cfd_sst_2d_dir, "SpalartAllmaras": cfd_sa_2d_dir}
 
 save = True
 savetype = ".pdf"
@@ -33,10 +35,21 @@ def plot_exp_meancontquiv():
     os.chdir(paper_dir)
     if save:
         plt.savefig("figures/meancontquiv_exp" + savetype)
+
+def plot_cfd_meancontquiv(case="kOmegaSST"):
+    """Plots wake mean velocity contours/quivers from 3-D CFD case."""
+    os.chdir(cfd_dirs[case])
+    import processing as processing_cfd
+    processing_cfd.plotwake()
+    os.chdir(paper_dir)
+    if save:
+         plt.savefig("figures/meancontquiv_" + case + savetype)
     
 def main():
     plot_exp_perf_curve()
     plot_exp_meancontquiv()
+    plot_cfd_meancontquiv("kOmegaSST")
+    plot_cfd_meancontquiv("SpalartAllmaras")
     plt.show()
     
 if __name__ == "__main__":

### 2015-03-17 makefigs.py
Use different plotting function for CFD data

@@ -40,7 +40,7 @@ def plot_cfd_meancontquiv(case="kOmegaSST"):
     """Plots wake mean velocity contours/quivers from 3-D CFD case."""
     os.chdir(cfd_dirs[case])
     import processing as processing_cfd
-    processing_cfd.plotwake()
+    processing_cfd.plot_meancontquiv()
     os.chdir(paper_dir)
     if save:
          plt.savefig("figures/meancontquiv_" + case + savetype)

### 2015-03-19 paper.tex
Add questions to introduction
Add comments on model verification

@@
 
+Questions addressed:
+
+\begin{enumerate}
+
+    \item Can 2-D RANS be used for array design?
+
+    \item Can ALM--RANS be used for array design?
+
+    \item Does 3-D RANS tell us more about what is happening, i.e, can it
+    ``interpolate'' the experimental results?
+
+\end{enumerate}
+
 \section{Numerical models}
 
+The RANS models employ a sliding mesh, where the turbine zone rotates at a mean
+tip speed ratio $\lambda=1.9$ with an unsteadiness as the blade passage
+frequency specified to mimic the experimental data.
+
+To capture the boundary layer,
+
 \section{Model verification}
+
+The $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified for
+convergence of the turbine mean power coefficient with respect to grid spacing
+and timestep. The grid topology was fixed, but the number of cells per unit
+length were scaled proportionally, maintaining the same background mesh cell
+aspect ratio. Results for this parameter sweep are shown in 
+Figure~\ref{fig:spatial-grid-dep}, from which the final number of streamwise
+grid points $N_x = 70$ was chosen.
+
+\begin{figure}[ht]
+\caption{Spatial grid dependence for the SST (a) and Spalart--Allmaras (b) 
+turbulence models for fixed timesteps.}
+\label{fig:spatial-grid-dep}
+\end{figure}
+
+Timestep dependence was studied using the $N_x=70$ grid, the results from which
+are shown in Figure~\ref{fig:timestep-dep}. It was seen that the
+Spalart--Allmaras model converged well with decreasing timestep, leading to a
+final timestep of 0.001 s. The results from the SST model show a local minimum,
+at $\Delta t = 0.002 s$, which was the final timestep chosen to run the
+simulations.
+
+\begin{figure}[ht]
+\caption{Spatial grid dependence for the SST (a) and Spalart--Allmaras (b) 
+turbulence models for a grid with $N_x=70$.}
+\label{fig:timestep-dep}
+\end{figure}
 
@@
 
-Predictions for both the power and drag coefficients are shown in Figure~\ref{fig:perf_comp}
+Predictions for both the power and drag coefficients are shown in Figure~\ref{fig:perf-comp}
 
-\begin{figure}
+\begin{figure}[ht]
 \caption{Power (a) and drag (b) coefficient predictions from experiments and
 each numerical model.}
-\label{fig:perf_comp}
+\label{fig:perf-comp}
 \end{figure}

### 2015-03-19 getbib.py
Modify getbib.py to download if necessary

@@ -5,8 +5,19 @@ This script copies my personal BibTeX file to the current directory.
 
 import os
 import shutil
+import sys
+if sys.version_info[0] == 3:
+    from urllib.request import urlretrieve
+else:
+    from urllib import urlretrieve
+    FileNotFoundError = IOError
 
 bibpath = os.path.join(os.path.expanduser("~"), "Google Drive",
                        "Library", "Library.bib")
-
-shutil.copy2(bibpath, "library.bib")
+biburl = "https://raw.githubusercontent.com/petebachant/Library/master/Library.bib"
+                       
+try:
+    shutil.copy2(bibpath, "library.bib")
+except FileNotFoundError:
+    print("Downloading BibTeX file")
+    urlretrieve(biburl, "library.bib")

### 2015-03-20 paper.tex
Add description of turbulence models chosen

@@
 
+The flow field was modeled using the Reynolds-averaged Navier--Stokes equations,
+employing two different turbulence models--Menter's $k$--$\omega$ SST
+\cite{Menter1994} and the Spalart--Allmaras one equation model
+\cite{Spalart1992}. The SST model was chosen due to its prominence in the
+literature for simulating separating flows, which we assumed to be present in
+the current problem in the form of dynamic stall. The Spalart--Allmaras model
+was shown by Ferreira et al. \cite{Ferreira2007} to match experimental particle
+image velocimetry (PIV) results for a CFT in dynamic stall, though this was a
+somewhat low Reynolds number case ($5 \times 10^4$).
+
 The RANS models employ a sliding mesh, where the turbine zone rotates at a mean

### 2015-03-21 paper.tex
Add mention of OpenFOAM
Add more details about numerical setup

@@
 
+In this study we set out to model the wake of the University of New Hampshire's
+Reference Vertical-Axis (cross-flow) Turbine (UNH-RVAT) using a
+Reynolds-averaged Navier--Stokes numerical model.
+
 Questions addressed:
@@
 
+\subsection{Computational mesh}
+
 The RANS models employ a sliding mesh, where the turbine zone rotates at a mean
 tip speed ratio $\lambda=1.9$ with an unsteadiness as the blade passage
-frequency specified to mimic the experimental data.
+frequency specified to mimic the experimental data. The 2-D mesh is shown in
+Figure~\ref{fig:mesh}.
 
 To capture the boundary layer,
+
+3-D computer aided design (CAD) files of the turbine are available from
+\cite{Bachant2014-RVAT-CAD}.
+
+The 3-D mesh was also generated with \textit{OpenFOAM}'s \textit{snappyHexMesh}
+application, with identical domain size, resolution, and boundary layer
+parameters in the $x$--$y$ plane, and \todo[inline]{Get $N_z$ for 3-D mesh.}
+cells in the vertical direction.
+
+\begin{figure}[ht]
+\caption{Snapshot of the 2-D computational mesh.}
+\label{fig:mesh}
+\end{figure}
+
+\subsection{Software}
+
+Simulations were run using the \textit{pimpleDyMFoam} solver from the
+open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
 

### 2015-07-20 paper.tex
Add to case description

@@
 % Multiple \preprint commands are allowed.
-%\preprint{}
+\preprint{}
 
@@
 
-The RANS models employ a sliding mesh, where the turbine zone rotates at a mean
-tip speed ratio $\lambda=1.9$ with an unsteadiness as the blade passage
-frequency specified to mimic the experimental data. The 2-D mesh is shown in
-Figure~\ref{fig:mesh}.
+The computational domain is a rectangular volume
 
-To capture the boundary layer,
+\todo[inline]{Add domain extent in terms of turbine diameter.}
+
+ with a cylindrical sliding mesh
+interface, where the turbine zone rotates at a mean tip speed ratio
+$\lambda=1.9$ with an unsteadiness as the blade passage frequency specified to
+mimic the experimental data. The 2-D mesh is shown in Figure~\ref{fig:mesh}.
+
+The mesh topology is kept constant, while its refinement is controlled by a
+single parameter---the number of cells in the streamwise direction, $N_x$.
+
+To capture the boundary layer, 20 layers are
+
+In the 3-D mesh, layers are not added to the struts and shaft to save
+computational expense.
 
@@
 open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
+
+\subsection{Initial and boundary conditions}
+
+Initial and boundary conditions were set to match those of the tow tank as well
+as possible.
 

### 2015-07-20 paper.tex
Add todonote

@@
 The computational domain is a rectangular volume
-
 \todo[inline]{Add domain extent in terms of turbine diameter.}
-
- with a cylindrical sliding mesh
-interface, where the turbine zone rotates at a mean tip speed ratio
-$\lambda=1.9$ with an unsteadiness as the blade passage frequency specified to
-mimic the experimental data. The 2-D mesh is shown in Figure~\ref{fig:mesh}.
+with a cylindrical sliding mesh interface, where the turbine zone rotates at a
+mean tip speed ratio $\lambda=1.9$ with a sinusoidal fluctuation at the blade
+passage frequency specified to mimic the experimental data. 
+\todo[inline]{Get parameters of TSR unsteadiness.}
+The 2-D mesh is
+shown in Figure~\ref{fig:mesh}.
 

### 2015-07-22 paper.tex
Add goals, but kind of shaky
Add to goals, remove ALM, use SA abbrev., and discuss SA free stream

@@
 
-    \item Can ALM--RANS be used for array design?
-
     \item Does 3-D RANS tell us more about what is happening, i.e, can it
     ``interpolate'' the experimental results?
+
+    \item Does 3-D RANS realize the proper wake recovery proportions? This would
+    imply it could be a very accurate model if used in DES.
+
+\end{enumerate}
+
+Goals:
+
+\begin{enumerate}
+
+    \item Evaluate modern computational fluid dynamics techniques for predicting
+    the dynamic loading on cross-flow turbine blades.
+
+    \item If possible, accurately predict the turbine from the perspective as a
+    wake generator, such that the results can be used to develop low-order
+    models.
 
@@
 employing two different turbulence models--Menter's $k$--$\omega$ SST
-\cite{Menter1994} and the Spalart--Allmaras one equation model
+\cite{Menter1994} and the Spalart--Allmaras (SA) one equation model
 \cite{Spalart1992}. The SST model was chosen due to its prominence in the
 literature for simulating separating flows, which we assumed to be present in
-the current problem in the form of dynamic stall. The Spalart--Allmaras model
-was shown by Ferreira et al. \cite{Ferreira2007} to match experimental particle
-image velocimetry (PIV) results for a CFT in dynamic stall, though this was a
-somewhat low Reynolds number case ($5 \times 10^4$).
+the current problem in the form of dynamic stall. The SA model was shown by
+Ferreira et al. \cite{Ferreira2007} to match experimental particle image
+velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
+low Reynolds number case ($5 \times 10^4$).
 
@@
 as possible.
+
+For the SA model, the turbulence transport variable $\tilde{\nu}$, is ideally
+zero in the free stream \cite{Spalart1992}, though others recommend using
+$\tilde{\nu}/\nu \approx 3$--$5$ for fully turbulent simulations
+\cite{Spalart2007}.
 

### 2015-07-22 paper.tex
Fix unit abbreviation
Write about the modeling challenge

@@
 Reynolds-averaged Navier--Stokes numerical model.
+
+Modeling the boundary layer flows on cross-flow turbine blades is very
+challenging due to the dynamically changing inflow and angle of attack---which
+often exceeds static stall values and causes dynamic stall. Furthermore, the
+ability to predict the occurrence and interdependence of boundary layer
+transition and separation can have dramatic influence on the blade loading and
+therefore the predicted turbine power output.
 
@@
 final timestep of 0.001 s. The results from the SST model show a local minimum,
-at $\Delta t = 0.002 s$, which was the final timestep chosen to run the
+at $\Delta t = 0.002$ s, which was the final timestep chosen to run the
 simulations.

### 2015-07-22 paper.tex
Add paragraph about low order modeling

@@
 
-Araya et al. modeled the flow through a VAT array using leaky Rankine bodies
-\cite{Araya2014}.
-
 In this study we set out to model the wake of the University of New Hampshire's
@@
 therefore the predicted turbine power output.
+
+It is assumed that if the turbine power output and near-wake predictions from
+the numerical model match those of the experiments, that the flow field can be
+inspected in greater detail, i.e., that the experiments will have been
+``interpolated''. This will allow us to see where the dominant flow structures
+originate and help develop low-order wake generator models for use in turbine
+array modeling. For example, Araya et al. modeled the flow through a VAT array
+using leaky Rankine bodies \cite{Araya2014}.
+Shamsoddin and Porte-Agel modeled a 2-D CFT using large eddy simulation (LES) with an actuator line approach to impart force on the flow field \cite{Shamsoddin2014}.
 

### 2015-07-23 paper.tex
Add array modeling refs
Describe the mesh quantitatively

@@
 array modeling. For example, Araya et al. modeled the flow through a VAT array
-using leaky Rankine bodies \cite{Araya2014}.
-Shamsoddin and Porte-Agel modeled a 2-D CFT using large eddy simulation (LES) with an actuator line approach to impart force on the flow field \cite{Shamsoddin2014}.
+using leaky Rankine bodies \cite{Araya2014}. Shamsoddin and Porte-Agel modeled a
+2-D CFT using large eddy simulation (LES) with an actuator line approach to
+impart force on the flow field \cite{Shamsoddin2014}.
+
+Goude and Agren used a 2-D vortex method to simulate a farm of cross-flow
+turbines \cite{Goude2010}. Durrani et al. used 2-D CFD to model a group of
+cross-flow turbines, but did not compare with experimental results
+\cite{Durrani2011}.
 
@@
 
-The computational domain is a rectangular volume
-\todo[inline]{Add domain extent in terms of turbine diameter.}
-with a cylindrical sliding mesh interface, where the turbine zone rotates at a
-mean tip speed ratio $\lambda=1.9$ with a sinusoidal fluctuation at the blade
-passage frequency specified to mimic the experimental data. 
-\todo[inline]{Get parameters of TSR unsteadiness.}
-The 2-D mesh is
-shown in Figure~\ref{fig:mesh}.
+The computational domain is a rectangular volume 3.66 m long, 3.66 m wide, and
+2.44 m tall (for 3-D), with the turbine located 1.52 m from the inlet, and
+centered vertically with a vertical axis. with a cylindrical sliding mesh
+interface, where the turbine zone rotates at a mean tip speed ratio
+$\lambda=1.9$ with a sinusoidal oscillation at the blade passage
+frequency---with and amplitude of 0.19 and the first peak at 0.7 radians---to
+mimic the experimental data. The 2-D mesh is shown in Figure~\ref{fig:mesh}.
 
-The mesh topology is kept constant, while its refinement is controlled by a
-single parameter---the number of cells in the streamwise direction, $N_x$.
-
-To capture the boundary layer, 20 layers are
-
-In the 3-D mesh, layers are not added to the struts and shaft to save
-computational expense.
+Meshes were generated using \textit{OpenFOAM}'s \textit{blockMesh} and
+\textit{snappHexMesh} utilities. Mesh topology consists of a background
+hexahedral mesh, which is refined in all three directions by a factor of 2 in a
+rectangular region containing the turbine and near-wake (0.9 m upstream, 1.3 m
+downstream and $\pm$ 0.8 m vertically). Cells adjacent to the turbine shaft and
+struts are refined by a factor of 4, while cells adjacent to the blades are
+refined by a factor of 6. To capture the boundary layer, 20 layers are added
+next to the blades with an expansion ratio of 1.2. Overall mesh refinement is
+controlled by a single parameter---the number of cells in the streamwise
+direction, $N_x$.
 
@@
 \cite{Bachant2014-RVAT-CAD}.
-
-The 3-D mesh was also generated with \textit{OpenFOAM}'s \textit{snappyHexMesh}
-application, with identical domain size, resolution, and boundary layer
-parameters in the $x$--$y$ plane, and \todo[inline]{Get $N_z$ for 3-D mesh.}
-cells in the vertical direction.
 

### 2015-07-24 paper.tex
Add cross-stream extent of the refined zone

@@
 rectangular region containing the turbine and near-wake (0.9 m upstream, 1.3 m
-downstream and $\pm$ 0.8 m vertically). Cells adjacent to the turbine shaft and
-struts are refined by a factor of 4, while cells adjacent to the blades are
-refined by a factor of 6. To capture the boundary layer, 20 layers are added
-next to the blades with an expansion ratio of 1.2. Overall mesh refinement is
-controlled by a single parameter---the number of cells in the streamwise
-direction, $N_x$.
+downstream, $\pm$ 0.9 m cross-stream, and $\pm$ 0.8 m vertically). Cells
+adjacent to the turbine shaft and struts are refined by a factor of 4, while
+cells adjacent to the blades are refined by a factor of 6. To capture the
+boundary layer, 20 layers are added next to the blades with an expansion ratio
+of 1.2. Overall mesh refinement is controlled by a single parameter---the number
+of cells in the streamwise direction, $N_x$.
 

### 2015-07-24 paper.tex
Plus-minus next to numbers

--- 2015-07-23 paper.tex
@@
 rectangular region containing the turbine and near-wake (0.9 m upstream, 1.3 m
-downstream and $\pm$ 0.8 m vertically). Cells adjacent to the turbine shaft and
-struts are refined by a factor of 4, while cells adjacent to the blades are
-refined by a factor of 6. To capture the boundary layer, 20 layers are added
-next to the blades with an expansion ratio of 1.2. Overall mesh refinement is
-controlled by a single parameter---the number of cells in the streamwise
-direction, $N_x$.
+downstream, $\pm 0.9$ m cross-stream, and $\pm 0.8$ m vertically). Cells
+adjacent to the turbine shaft and struts are refined by a factor of 4, while
+cells adjacent to the blades are refined by a factor of 6. To capture the
+boundary layer, 20 layers are added next to the blades with an expansion ratio
+of 1.2. Overall mesh refinement is controlled by a single parameter---the number
+of cells in the streamwise direction, $N_x$.
 

### 2015-07-24 paper.tex
Correct spelling of snappyHexMesh
Add statement about vertical wake recovery

@@
 
+It has been shown that the momentum and energy transport processes in the
+near-wake of a high-solidity, low aspect ratio vertical-axis turbine are
+dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
+that a 2-D simulation would not correctly predict wake recovery, but is of
+interest to what extent this is true, since the computational feasibility of 2-D
+simulations is attractive.
+
 Modeling the boundary layer flows on cross-flow turbine blades is very
@@
 Meshes were generated using \textit{OpenFOAM}'s \textit{blockMesh} and
-\textit{snappHexMesh} utilities. Mesh topology consists of a background
+\textit{snappyHexMesh} utilities. Mesh topology consists of a background
 hexahedral mesh, which is refined in all three directions by a factor of 2 in a

### 2015-07-24 paper.tex
Add subsection on wake results

@@
 
+\subsection{Wake characteristics}
+
+Mean velocity profiles at one turbine diameter downstream are shown in
+Figure~\ref{fig:mean-vel-profiles}. The 2-D results suffer from a blockage
+mismatch, i.e. keeping the proximity of the walls constant increases the
+blockage ratio.
+
+\begin{figure}
+\caption{Mean velocity profiles at $x/D=1$ from 2-D simulations, 3-D simulations
+($z/H=0$), and experiments \cite{Bachant2015-JoT}.}
+\label{fig:mean-vel-profiles}
+\end{figure}
+
+Turbulence kinetic energy profiles are shown in Figure~\ref{fig:k-profiles}.
+Note that there are two profiles for each---one computed from the components
+from the Reynolds stress tensor $R$:
+
+\begin{equation}
+k = \frac{1}{2} \left( R_{xx} + R_{yy} + R_{zz} \right);
+\label{eq:k}
+\end{equation}
+and one that includes unsteadiness in the Reynolds-averaged velocity
+\begin{equation}
+k_{\mathrm{tot}} = k + \frac{1}{2} \left( 
+\overline{U^\prime}^2 + 
+\overline{V^\prime}^2 + 
+\overline{W^\prime}^2 \right),
+\label{eq:ktot}
+\end{equation}
+where $U^\prime = U - \overline{U}$.
+
+\begin{figure}
+\caption{Turbulence kinetic energy profiles at $x/D=1$ from 2-D simulations, 3-D
+simulations ($z/H=0$), and experiments \cite{Bachant2015-JoT}.}
+\label{fig:k-profiles}
+\end{figure}
+
+We see that the side of the turbine with the upwind facing blade ($+y$), despite
+having regions of high turbulence intensity, is actually more dominated by
+unsteadiness in the Reynolds-averaged velocity, rather than the model's
+turbulence kinetic energy.
+
 \begin{acknowledgments}

### 2015-08-10 makefigs.py
Look up actual 3-D SST recovery

@@ -255,7 +255,7 @@ def make_recovery_bar_chart():
     data["SA (2-D)"] = sst2dpr.read_funky_log()
     # Load recovery terms from 3-D SST case (0.0 for now)
     os.chdir(cfd_dirs["3-D"]["kOmegaSST"])
-    data["SST (3-D)"] = {k: 0.0 for k in data["SST (2-D)"].keys()}
+    data["SST (3-D)"] = sst3dpr.read_funky_log()
     # Load recovery terms from 3-D SA case
     os.chdir(cfd_dirs["3-D"]["SpalartAllmaras"])
     data["SA (3-D)"] = sa3dpr.read_funky_log()

### 2015-08-11 paper.tex
Note that TKE is zero for SA and that variances are from downsampled time series; fixes #17 and fixes #20

@@
 
-Turbulence kinetic energy profiles are shown in Figure~\ref{fig:k-profiles}.
-Note that there are two profiles for each---one computed from the components
-from the Reynolds stress tensor $R$:
-
+Turbulence kinetic energy profiles are shown in Figure~\ref{fig:k-profiles}. The
+turbulence kinetic energy is calculated as
 \begin{equation}
-k = \frac{1}{2} \left( R_{xx} + R_{yy} + R_{zz} \right);
-\label{eq:k}
-\end{equation}
-and one that includes unsteadiness in the Reynolds-averaged velocity
-\begin{equation}
-k_{\mathrm{tot}} = k + \frac{1}{2} \left( 
+k = k_{\mathrm{RA}} + \frac{1}{2} \left( 
 \overline{U^\prime}^2 + 
@@
 \overline{W^\prime}^2 \right),
-\label{eq:ktot}
+\label{eq:k}
 \end{equation}
-where $U^\prime = U - \overline{U}$.
+where $U^\prime = U - \overline{U}$ and $k_{\mathrm{RA}}$ is the kinetic energy
+calculated by the turbulence model, which is zero for the SA model. Note that
+the statistics of the Reynolds-averaged velocity are calculated from a time
+series that has been downsampled to 50 Hz. It is assumed that this is far enough
+from the blade passage frequency that differences from the variance in the
+original velocity will be negligible
 

### 2015-08-11 paper.tex
Both profiles are on one figure

@@
 Mean velocity profiles at one turbine diameter downstream are shown in
-Figure~\ref{fig:mean-vel-profiles}. The 2-D results suffer from a blockage
+Figure~\ref{fig:profiles}. The 2-D results suffer from a blockage
 mismatch, i.e. keeping the proximity of the walls constant increases the
@@
 \begin{figure}
-\caption{Mean velocity profiles at $x/D=1$ from 2-D simulations, 3-D simulations
-($z/H=0$), and experiments \cite{Bachant2015-JoT}.}
-\label{fig:mean-vel-profiles}
+\includegraphics[width=0.95\textwidth]{figures/profiles} \caption{Mean velocity
+	(left) and turbulence kinetic energy (right) profiles at $x/D=1$ from 2-D
+	simulations, 3-D simulations ($z/H=0$), and experiments \cite{Bachant2015-JoT}.}
+\label{fig:profiles}
 \end{figure}
 
-Turbulence kinetic energy profiles are shown in Figure~\ref{fig:k-profiles}. The
-turbulence kinetic energy is calculated as
+Turbulence kinetic energy profiles are also shown in Figure~\ref{fig:profiles}.
+The turbulence kinetic energy is calculated as
 \begin{equation}
@@
 from the blade passage frequency that differences from the variance in the
-original velocity will be negligible
-
-\begin{figure}
-\caption{Turbulence kinetic energy profiles at $x/D=1$ from 2-D simulations, 3-D
-simulations ($z/H=0$), and experiments \cite{Bachant2015-JoT}.}
-\label{fig:k-profiles}
-\end{figure}
+original velocity will be negligible.
 

### 2015-08-11 paper.tex
Add momentum recovery subsubsection

@@
 
+\subsubsection{Momentum recovery}
+
+To get an overall idea of the wake recovery predicted by each model, we
+rearrange the streamwise component of the Navier--Stokes equation to isolate
+$\partial U / \partial x$---following Bachant and
+Wosnik\cite{Bachant2015-JoT}---and compute each term at $X/D = 1$ to compare
+with the experimental results.
+
+\begin{figure}[ht]
+    \includegraphics[width=0.9\textwidth]{figures/mom_bar_graph}
+    \caption{Weighted sum normalized momentum recovery terms for each CFD case
+        and experiments\cite{Bachant2015-JoT} at $x/D=1$.} \label{fig:recovery}
+\end{figure}
+
 \begin{acknowledgments}

### 2015-08-11 paper.tex
Add performance results and approximate computation times for CFD cases; closes #13

@@
 Figure~\ref{fig:spatial-grid-dep}, from which the final number of streamwise
-grid points $N_x = 70$ was chosen.
+grid points $N_x = 70$ was chosen. This corresponds to a total cell count of approximately $5 \times 10^4$ for the 2-D cases and 
+\todo[inline]{Get total cell count for 3-D case.}
 
@@
 
+Computations for the 3-D cases were run on 192 processes and took on the order
+of 1,000 CPU hours per second of simulation time. The 2-D simulations were run
+on a single processor and took on the order of one CPU hour per second.
+
 \subsection{Performance prediction}
 
-Predictions for both the power and drag coefficients are shown in Figure~\ref{fig:perf-comp}
+Predictions for both the mean rotor power and drag coefficients are shown in
+Figure~\ref{fig:perf-comp}. In general, the 2-D CFD cases both overpredict
+turbine loading, which is likely due to their increased blockage ratio,
+unresolved blade end effects, and lack of blade support struts.
+
+The 3-D simulations fair better at predicting the experimental measurements,
+with the Spalart--Allmaras model performing relatively better. The apparent
+overprediction of rotor drag coefficient could be an effect of the experimental
+procedure, where the ``tare drag'' from the turbine mounting structure was
+measured without a turbine installed, then subtracted in post-processing.
+Technically the local flow field will have changed, and hence the tare drag will
+change as well.
 
 \begin{figure}[ht]
-\caption{Power (a) and drag (b) coefficient predictions from experiments and
-each numerical model.}
-\label{fig:perf-comp}
+    \includegraphics[width=0.9\textwidth]{figures/perf_bar_chart} \caption{Power
+        (left) and drag (right) coefficient predictions from experiments and each
+        numerical model.} \label{fig:perf-comp}
 \end{figure}
-
-\todo[inline]{Add bar graph comparing various performance results for the RVAT.}
 

### 2015-08-12 makefigs.py
Move save options to bottom

@@ -46,9 +46,6 @@ U = 1.0
 D = 1.0
 nu = 1e-6
 
-save = True
-savetype = ".pdf"
-
 def load_exp_data():
     """Loads section of exp wake data for U_infty=1.0 m/s."""
     return pd.read_csv(os.path.join(exp_dir, "Data", "Processed", 
@@ -307,13 +304,16 @@ if __name__ == "__main__":
     set_sns()
     plt.rcParams["axes.grid"] = True
     
+    save = True
+    savetype = ".pdf"
+    
 #    plot_exp_perf_curve()
 #    plot_exp_meancontquiv()
 #    plot_cfd_meancontquiv("kOmegaSST")
 #    plot_cfd_meancontquiv("SpalartAllmaras")
 #    plot_cfd_u_profile()
-#    plot_verification()
+    plot_verification()
 #    plot_profiles()
 #    make_perf_bar_charts()
-    make_recovery_bar_chart()
+#    make_recovery_bar_chart()
     plt.show()

### 2015-08-12 paper.tex
Remove incorrect statement
Add verification figure and 3-D mesh count; closes #23 and closes #6

@@
 Figure~\ref{fig:spatial-grid-dep}, from which the final number of streamwise
-grid points $N_x = 70$ was chosen. This corresponds to a total cell count of approximately $5 \times 10^4$ for the 2-D cases and 
-\todo[inline]{Get total cell count for 3-D case.}
+grid points $N_x = 70$ was chosen. This corresponds to a total cell count of
+approximately $5 \times 10^4$ for the 2-D cases and 16 million for the 3-D case.
 
 \begin{figure}[ht]
-\caption{Spatial grid dependence for the SST (a) and Spalart--Allmaras (b) 
-turbulence models for fixed timesteps.}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{figures/verification}
+
+    \caption{Time step (left) and grid size (right) dependence for the 2-D case
+        with both the SST and SA turbulence models. Time step dependence was carried
+        out with $N_x=70$ and grid size dependence with the time steps annotated for
+        each turbulence model.}
+
     \label{fig:spatial-grid-dep}
@@
 
-Timestep dependence was studied using the $N_x=70$ grid, the results from which
+Time step dependence was studied using the $N_x=70$ grid, the results from which
 are shown in Figure~\ref{fig:timestep-dep}. It was seen that the
-Spalart--Allmaras model converged well with decreasing timestep, leading to a
-final timestep of 0.001 s. The results from the SST model show a local minimum,
-at $\Delta t = 0.002$ s, which was the final timestep chosen to run the
+Spalart--Allmaras model converged well with decreasing time step, leading to a
+final time step of 0.001 s. The results from the SST model show a local minimum,
+at $\Delta t = 0.002$ s, which was the final time step chosen to run the
 simulations.
 
-\begin{figure}[ht]
-\caption{Spatial grid dependence for the SST (a) and Spalart--Allmaras (b) 
-turbulence models for a grid with $N_x=70$.}
-\label{fig:timestep-dep}
-\end{figure}
+\todo[inline]{Find some examples in the literature where time steps are much
+    larger.}
 
@@
 
-We see that the side of the turbine with the upwind facing blade ($+y$), despite
-having regions of high turbulence intensity, is actually more dominated by
-unsteadiness in the Reynolds-averaged velocity, rather than the model's
-turbulence kinetic energy.
-
 \subsubsection{Momentum recovery}

### 2015-08-12 paper.tex
Some rewrapping and reorganizing

@@
 \begin{figure}[ht]
-    \includegraphics[width=0.9\textwidth]{figures/perf_bar_chart} \caption{Power
-        (left) and drag (right) coefficient predictions from experiments and each
-        numerical model.} \label{fig:perf-comp}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{figures/perf_bar_chart}
+
+    \caption{Power (left) and drag (right) coefficient predictions from
+        experiments and each numerical model.}
+
+    \label{fig:perf-comp}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{figures/profiles} \caption{Mean velocity
-	(left) and turbulence kinetic energy (right) profiles at $x/D=1$ from 2-D
-	simulations, 3-D simulations ($z/H=0$), and experiments \cite{Bachant2015-JoT}.}
+    \centering
+
+    \includegraphics[width=0.95\textwidth]{figures/profiles}
+
+    \caption{Mean velocity (left) and turbulence kinetic energy (right) profiles
+        at $x/D=1$ from 2-D simulations, 3-D simulations ($z/H=0$), and experiments
+        \cite{Bachant2015-JoT}.}
+
     \label{fig:profiles}

### 2015-08-12 paper.tex
Fix some labels

@@
 aspect ratio. Results for this parameter sweep are shown in
-Figure~\ref{fig:spatial-grid-dep}, from which the final number of streamwise
+Figure~\ref{fig:verification}, from which the final number of streamwise
 grid points $N_x = 70$ was chosen. This corresponds to a total cell count of
@@
 
-    \label{fig:spatial-grid-dep}
+    \label{fig:verification}
 \end{figure}
@@
 Time step dependence was studied using the $N_x=70$ grid, the results from which
-are shown in Figure~\ref{fig:timestep-dep}. It was seen that the
+are shown in Figure~\ref{fig:verification}. It was seen that the
 Spalart--Allmaras model converged well with decreasing time step, leading to a
@@
 \begin{figure}[ht]
+    \centering
+
     \includegraphics[width=0.9\textwidth]{figures/mom_bar_graph}
+
     \caption{Weighted sum normalized momentum recovery terms for each CFD case
-        and experiments\cite{Bachant2015-JoT} at $x/D=1$.} \label{fig:recovery}
+        and experiments\cite{Bachant2015-JoT} at $x/D=1$.}
+
+    \label{fig:recovery}
 \end{figure}

### 2015-08-12 paper.tex
Start conclusions

@@
 
+\section{Conclusions}
+
+A cross-flow turbine was modeled using the $k$--$\omega$ SST and
+Spalart--Allmaras Reynolds-averaged Navier--Stokes turbulence models to test
+their abilities to predict turbine performance and near-wake dynamics. It was
+observed that when modeled in 2-D, the performance is over-predicted compared to
+the data from the tow tank experiments, which was expected due to omission of
+blade end effects, support strut drag, and increased blockage. Vertical (or
+axial) wake dynamics were unresolved in the 2-D model, despite being identified
+as a significant contributor to streamwise wake recovery, which casts doubt on
+the 2-D model's applicability as a tool to study array spacing effects.
+
 \begin{acknowledgments}

### 2015-08-13 makefigs.py
Profiles figure looking good

@@ -147,19 +147,26 @@ def plot_profiles():
     # Load data from 2-D SA case
     os.chdir(cfd_dirs["2-D"]["SpalartAllmaras"])
     df = sa2dpr.load_u_profile()
-    ax[0].plot(df.y_R, df.u, "--", label="SA (2-D)")
+    ax[0].plot(df.y_R, df.u, "-.", label="SA (2-D)")
     df = sa2dpr.load_k_profile()
-    ax[1].plot(df.y_R, df.k_total, "--", label="SA (2-D)")
+    ax[1].plot(df.y_R, df.k_total, "-.", label="SA (2-D)")
+    # Load data from 3-D SST case
+    os.chdir(cfd_dirs["3-D"]["kOmegaSST"])
+    df = sst3dpr.load_u_profile()
+    ax[0].plot(df.y_R, df.u, "--", label="SST (3-D)")
+    df = sa3dpr.load_k_profile()
+    ax[1].plot(df.y_R, df.k_total, "--", label="SST (3-D)")
     # Load data from 3-D SA case
     os.chdir(cfd_dirs["3-D"]["SpalartAllmaras"])
     df = sa3dpr.load_u_profile()
-    ax[0].plot(df.y_R, df.u, "-.", label="SA (3-D)")
+    ax[0].plot(df.y_R, df.u, ":", label="SA (3-D)")
     df = sa3dpr.load_k_profile()
-    ax[1].plot(df.y_R, df.k_total, "-.", label="SA (3-D)")
+    ax[1].plot(df.y_R, df.k_total, ":", label="SA (3-D)")
     # Load data from experiment
     df = load_exp_data()
-    ax[0].plot(df.y_R, df.mean_u, "--o", label="Exp.")
-    ax[1].plot(df.y_R, df.k, "--o", label="Exp.")
+    df = df[df.z_H == 0]
+    ax[0].plot(df.y_R, df.mean_u, "o", label="Exp.", markerfacecolor="none")
+    ax[1].plot(df.y_R, df.k, "o", label="Exp.", markerfacecolor="none")
     # Set legend and labels
     ax[1].legend(loc="best")
     for a in ax: a.set_xlabel("$y/R$")
@@ -312,8 +319,8 @@ if __name__ == "__main__":
 #    plot_cfd_meancontquiv("kOmegaSST")
 #    plot_cfd_meancontquiv("SpalartAllmaras")
 #    plot_cfd_u_profile()
-    plot_verification()
[…]

### 2015-08-13 makefigs.py
Fix up perf figure

@@ -50,14 +50,24 @@ def load_exp_data():
     """Loads section of exp wake data for U_infty=1.0 m/s."""
     return pd.read_csv(os.path.join(exp_dir, "Data", "Processed", 
                                     "Wake-1.0.csv"))
+                                    
+def load_exp_perf_data():
+    """Loads section of exp perf data for U_infty=1.0 m/s."""
+    return pd.read_csv(os.path.join(exp_dir, "Data", "Processed", 
+                                    "Perf-1.0.csv"))
 
-def plot_exp_perf_curve():
-    os.chdir(exp_dir)
-    import Modules.plotting as plotting_exp
-    plotting_exp.plot_cp_curve(1.0, show=False)
-    os.chdir(paper_dir)
+def plot_exp_perf():
+    df = load_exp_perf_data()
+    fig, ax = plt.subplots(nrows=1, ncols=2, figsize=(7.5, 3))
+    ax[0].plot(df.mean_tsr, df.mean_cp, "-o")
+    ax[0].set_ylabel(r"$C_P$")
+    ax[1].plot(df.mean_tsr, df.mean_cd, "-o")
+    ax[1].set_ylabel(r"$C_D$")
+    for a in ax:
+        a.set_xlabel(r"$\lambda$")
+    fig.tight_layout()
     if save:
-        plt.savefig("figures/perf_curve_exp" + savetype)
+        fig.savefig("figures/exp_perf" + savetype)
     
 def plot_exp_meancontquiv():
     os.chdir(exp_dir)
@@ -314,13 +324,13 @@ if __name__ == "__main__":
     save = True
     savetype = ".pdf"
     
-#    plot_exp_perf_curve()
+    plot_exp_perf()
 #    plot_exp_meancontquiv()
 #    plot_cfd_meancontquiv("kOmegaSST")
 #    plot_cfd_meancontquiv("SpalartAllmaras")
 #    plot_cfd_u_profile()
 #    plot_verification()
-    plot_profiles()
+#    plot_profiles()
 #    make_perf_bar_charts()
 #    make_recovery_bar_chart()
     plt.show()

### 2015-08-13 paper.tex
Add perf figure

@@
 \end{enumerate}
+
+In this study we seek to predict the near-wake of the UNH-RVAT turbine,
+experimental performance curves for which are shown in
+Figure~\ref{fig:exp_perf}. We modeled the turbine at a tip speed ratio
+$\lambda=1.9$, which corresponds to the maximum power coefficient.
+
+\begin{figure}[ht]
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{figures/exp_perf}
+
+    \caption{Mean rotor power (left) and drag (right) coefficient curves from
+        the tow tank experiments\cite{Bachant2015-RVAT-Re-dep-data}.}
+
+    \label{fig:exp_perf}
+\end{figure}
 

### 2015-08-31 paper.tex
Whitespace stuff
Trying out new title
Add m-dash and abstract

@@
 
-\title{Modeling vertical-axis cross-flow turbine wakes}
+\title{Modeling the near-wake of a vertical-axis cross-flow turbine with 2-D and
+    3-D RANS}
 
@@
 The wake of a vertical-axis cross-flow turbine (CFT) is modeled numerically via
-several techniques.
+blade-resolved $k$--$\omega$ SST and Spalart--Allmaras RANS models in two and
+three dimensions. Results for each case are compared with experimental
+measurements of the turbine shaft power, overall drag, mean velocity, turbulence
+kinetic energy, and momentum transport terms in the near-wake at one diameter
+downstream. It is shown that 2-D simulations overpredict turbine loading and do
+not resolve mean vertical momentum transport, which plays an important role in
+the near-wake's momentum balance. The 3-D simulations fair better at predicting
+performance, with the Spalart--Allmaras model predictions being closest to the
+experiments. The SST model more accurately predicts the turbulence kinetic
+energy while the Spalart--Allmaras model more closely matches the momentum
+transport terms in the near-wake. These results show the potential of RANS as a
+design tool and a way to ``interpolate'' experimental flow field measurements.
 \end{abstract}
@@
 The flow field was modeled using the Reynolds-averaged Navier--Stokes equations,
-employing two different turbulence models--Menter's $k$--$\omega$ SST
+employing two different turbulence models---Menter's $k$--$\omega$ SST
 \cite{Menter1994} and the Spalart--Allmaras (SA) one equation model

### 2015-09-01 paper.tex
Add to intro

@@
 
+Computational fluid dynamics (CFD) presents an attractive method for simulating
+cross-flow turbines since it only requires turbulence modeling, i.e., the
+machine is described from first principles. However, CFD can be computationally
+expensive when done in three dimensions, which may be necessary in some cases.
+There are many examples in the literature of 2-D cross-flow turbine simulations
+with widely varying results.
+
 In this study we set out to model the wake of the University of New Hampshire's
@@
 Reynolds-averaged Navier--Stokes numerical model.
+
+Note that studies in the literature generally focus on predicting the turbine
+loading and the local blade flow rather than the turbine wake. In contrast, we
+seek to model the near-wake flow.
 

### 2015-09-01 paper.tex
Discuss BCs
Add figure placeholder for 2-D blade mesh closeup

@@
 frequency---with and amplitude of 0.19 and the first peak at 0.7 radians---to
-mimic the experimental data. The 2-D mesh is shown in Figure~\ref{fig:mesh}.
+mimic the experimental data. The 2-D mesh overview is shown in
+Figure~\ref{fig:mesh} and the blade mesh is shown in
+Figure~\ref{fig:blade-mesh}.
 
@@
 \begin{figure}[ht]
-\caption{Snapshot of the 2-D computational mesh.}
+    \caption{Overview of the 2-D computational mesh.}
+
     \label{fig:mesh}
+\end{figure}
+
+\begin{figure}[ht]
+    \caption{Detailed view of the 2-D computational mesh near the blades.}
+
+    \label{fig:blade-mesh}
 \end{figure}
@@
 Initial and boundary conditions were set to match those of the tow tank as well
-as possible.
+as possible. The velocity at the bottom and side walls was fixed to 1 m/s to
+match the tow tank case, while the top boundary condition was a slip velocity
+condition. Note that in 2-D the top and bottom boundary conditions are
+``empty,'' which is an OpenFOAM convention.
 

### 2015-09-01 paper.tex
Add comment about PISO-SIMPLE algorithm
Software -> solver
Remove paragraph about free stream nuTilda since these guidelines were not used

@@
 
-\subsection{Software}
+\subsection{Solver}
 
@@
 open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
+\textit{pimpleDyMFoam} uses a hybrid PISO-SIMPLE algorithm for pressure-velocity
+coupling and is compatible with dynamic meshes.
 
@@
 ``empty,'' which is an OpenFOAM convention.
-
-For the SA model, the turbulence transport variable $\tilde{\nu}$, is ideally
-zero in the free stream \cite{Spalart1992}, though others recommend using
-$\tilde{\nu}/\nu \approx 3$--$5$ for fully turbulent simulations
-\cite{Spalart2007}.
 

### 2015-09-01 paper.tex
Cite Crivellini2014; resolves #18
Add Balduzzi2016 citation

@@
 Computational fluid dynamics (CFD) presents an attractive method for simulating
-cross-flow turbines since it only requires turbulence modeling, i.e., the
+cross-flow turbines (CFTs) since it only requires turbulence modeling, i.e., the
 machine is described from first principles. However, CFD can be computationally
@@
 There are many examples in the literature of 2-D cross-flow turbine simulations
-with widely varying results.
+with widely varying results. Balduzzi et al. provides a summary of recent
+computations and an attempt to standardize a methodology for using
+Reynolds-averaged Navier--Stokes (RANS) to correctly simulate performance of a
+2-D CFT \cite{Balduzzi2016}.
 
@@
 Figure~\ref{fig:exp_perf}. We modeled the turbine at a tip speed ratio
-$\lambda=1.9$, which corresponds to the maximum power coefficient.
+$\lambda=1.9$, which corresponds to the maximum power coefficient, and a turbine
+diameter Reynolds number $Re_D \approx 10^6$, which corresponds to a
+$Re$-independent state for both performance and near-wake characteristics, as
+determined from experimental measurements \cite{Bachant2014,
+    Bachant2015-RVAT-Re-dep}.
 
@@
 velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
-low Reynolds number case ($5 \times 10^4$).
+low Reynolds number case ($5 \times 10^4$). Further justification for using the
+SA model for this case comes from Crivellini and D'Alessandro
+\cite{Crivellini2014}, where they successfully modeled the laminar separation
+bubble and subsequent boundary layer transition to turbulence at Reynolds
+numbers similar to ours.
 

### 2015-09-01 getfigs.py
Start getfigs.py

@@ -1,0 +1,7 @@
+#!/usr/bin/env python
+"""
+This script copies figures to this directory.
+"""
+
+import os
+

### 2015-09-01 getfigs.py
Add docs to getfigs.py

@@ -1,7 +1,11 @@
 #!/usr/bin/env python
 """
 This script copies figures to this directory.
+
+Figures that need to be copied:
+  * Mesh overview from 2-D SST RANS case
+  * Blade mesh closeup from 2-D SST RANS case
+  * Vorticity contours from 2-D SA case
 """
 
 import os
-

### 2015-09-01 paper.tex
Add fig and description for 3-D vorticity contours

@@
 
+Visualizing the complicated wake generated by the turbine is presented for the
+3-D Spalart--Allmaras case in Figure~\ref{fig:vorticity-3D}. It can be seen how
+the upstream blade is shedding a large amount of spanwise vorticity due to the
+separated flow. Strong tip vortices are also present, which trace the
+``contracting'' wake flow on the $-y$ side of the turbine associated with the
+induced vertical velocity field.
+
+\begin{figure}
+    \caption{Iso-vorticity contours colored by the streamwise component of
+        vorticity for the 3-D Spalart--Allmaras case.}
+
+    \label{fig:vorticity-3D}
+\end{figure}
+
 Mean velocity profiles at one turbine diameter downstream are shown in

### 2015-09-02 paper.tex
Add mesh figs
Add 2-D vorticity figure and discuss benefits of CFD interpolation

--- 2015-09-01 paper.tex
@@
 inspected in greater detail, i.e., that the experiments will have been
-``interpolated''. This will allow us to see where the dominant flow structures
-originate and help develop low-order wake generator models for use in turbine
-array modeling. For example, Araya et al. modeled the flow through a VAT array
-using leaky Rankine bodies \cite{Araya2014}. Shamsoddin and Porte-Agel modeled a
-2-D CFT using large eddy simulation (LES) with an actuator line approach to
-impart force on the flow field \cite{Shamsoddin2014}.
+``interpolated''. This will give us access to unmeasured quantities, e.g.
+pressure, and allow us to see where the dominant flow structures originate and
+help develop low-order wake generator models for use in turbine array modeling.
+For example, Araya et al. modeled the flow through a VAT array using leaky
+Rankine bodies \cite{Araya2014}. Shamsoddin and Porte-Agel modeled a 2-D CFT
+using large eddy simulation (LES) with an actuator line approach to impart force
+on the flow field \cite{Shamsoddin2014}.
 
@@
 \begin{figure}[ht]    
+    \includegraphics[width=0.8\textwidth]{figures/2d_mesh}
+
     \caption{Overview of the 2-D computational mesh.}
@@
 \begin{figure}[ht]
+    \includegraphics[width=0.7\textwidth]{figures/2D_blade_mesh_closeup}
+
     \caption{Detailed view of the 2-D computational mesh near the blades.}
@@
 
+Visualizing the complicated wake generated by the turbine is presented for the
+2-D and 3-D Spalart--Allmaras cases in Figure~\ref{fig:vorticity-3D} and
+Figure~\ref{fig:vorticity-3D}, respectively. It can be seen how the upstream
+blade---as it turns back into the streamwise direction---is shedding a large
+amount of spanwise vorticity due to the separated flow. In the 3-D case, strong
+tip vortices are also present, which trace the ``contracting'' wake flow on the
+$-y$ side of the turbine associated with the induced vertical velocity field.
+
+\begin{figure}
+    \includegraphics[width=0.65\textwidth]{figures/2D_vorticity_SA_964}
+
+    \caption{Instantaneous vorticity contours (at $t=9.64$ s) computed for the
+        2-D Spalart--Allmaras case.}
+
+    \label{fig:vorticity-2D}
+\end{figure}
+
+\begin{figure}
+    \caption{Iso-vorticity contours colored by the streamwise component of
+        vorticity for the 3-D Spalart--Allmaras case.}
+
+    \label{fig:vorticity-3D}
+\end{figure}
+
 Mean velocity profiles at one turbine diameter downstream are shown in

### 2015-09-02 paper.tex
Add paragraph to intro and make second par flow better

@@
 
-Computational fluid dynamics (CFD) presents an attractive method for simulating
-cross-flow turbines (CFTs) since it only requires turbulence modeling, i.e., the
-machine is described from first principles. However, CFD can be computationally
-expensive when done in three dimensions, which may be necessary in some cases.
-There are many examples in the literature of 2-D cross-flow turbine simulations
-with widely varying results. Balduzzi et al. provides a summary of recent
-computations and an attempt to standardize a methodology for using
-Reynolds-averaged Navier--Stokes (RANS) to correctly simulate performance of a
-2-D CFT \cite{Balduzzi2016}.
+In the pursuit of a fully sustainable energy profile, cross-flow (often
+vertical-axis) turbines (CFTs) still play a role. Despite extensive research and
+development in the 1970s through the early 1990s by groups like Sandia National
+Labs in the US \cite{Sutherland2012} and the National Research Council in Canada
+\cite{Para2002}, the vertical-axis cross-flow Darrieus turbines were all but
+abandoned for large scale commercial terrestrial wind power. Today, however,
+with the development of marine hydrokinetic (MHK) energy devices, CFTs are at
+the forefront, with the first grid-connected commercial device being a
+horizontal-axis CFT manufactured and installed by ORPC in Cobscook Bay, Maine
+\cite{ORPC2012}. CFTs are also being examined for small-scale onshore wind in
+urban areas \cite{Lott2015}, offshore floating wind farms \cite{Paulsen2011,
+    Sandia2012}, and onshore wind farms where power output per unit land area is of
+interest \cite{Dabiri2011}.
+
+Despite the development of many simple engineering models based on blade element
+momentum or vortex modeling, it remains difficult to predict the performance of
+cross-flow turbines in all cases---namely when solidity or blade chord to radius
+ratio is high, a common characteristic of the smaller CFTs. With computing power
+becoming evermore available and affordable, computational fluid dynamics (CFD)
+presents an attractive method for simulating cross-flow turbines since it only
+requires turbulence modeling, i.e., the machine is described from first
+principles. However, CFD can be computationally expensive when done in three
+dimensions, which may be necessary in some cases. There are many examples in the
+literature of 2-D cross-flow turbine simulations with widely varying results.
+Balduzzi et al. provides a summary of recent computations and an attempt to
+standardize a methodology for using Reynolds-averaged Navier--Stokes (RANS) to
+correctly simulate performance of a 2-D CFT \cite{Balduzzi2016}.
 

### 2015-09-03 paper.tex
Add 3-D vorticity figure
Add em dash

@@
 cross-flow turbines in all cases---namely when solidity or blade chord to radius
-ratio is high, a common characteristic of the smaller CFTs. With computing power
-becoming evermore available and affordable, computational fluid dynamics (CFD)
-presents an attractive method for simulating cross-flow turbines since it only
-requires turbulence modeling, i.e., the machine is described from first
+ratio is high---a common characteristic of the smaller CFTs. With computing
+power becoming evermore available and affordable, computational fluid dynamics
+(CFD) presents an attractive method for simulating cross-flow turbines since it
+only requires turbulence modeling, i.e., the machine is described from first
 principles. However, CFD can be computationally expensive when done in three
@@
 \begin{figure}
+    \includegraphics[width=0.8\textwidth]{figures/3D_vorticity_SA_964_10-threshold}
+
     \caption{Iso-vorticity contours colored by the streamwise component of

### 2015-09-03 paper.tex
Add empty figure for CAD drawing

@@
 
-In this study we seek to predict the near-wake of the UNH-RVAT turbine,
-experimental performance curves for which are shown in
-Figure~\ref{fig:exp_perf}. We modeled the turbine at a tip speed ratio
+In this study we seek to predict the near-wake of the UNH-RVAT turbine, shown in
+Figure~\ref{fig:RVAT-CAD}. Experimental performance curves for the UNH-RVAT are
+shown in Figure~\ref{fig:exp_perf}. We modeled the turbine at a tip speed ratio
 $\lambda=1.9$, which corresponds to the maximum power coefficient, and a turbine
@@
     Bachant2015-RVAT-Re-dep}.
+
+\begin{figure}[ht]
+    \centering
+
+    \caption{CAD model of the UNH-RVAT cross-flow turbine}
+
+    \label{fig:RVAT-CAD}
+\end{figure}
 

### 2015-09-03 paper.tex
Add CAD drawings of RVAT
Add subsubsections for 3D wake comparisons
Add note about tip loss corrections

@@
 
-    \caption{CAD model of the UNH-RVAT cross-flow turbine}
+    \includegraphics[clip, trim=0 1in 0 1in, width=0.8\textwidth]{figures/CAD}
+
+    \caption{CAD drawings of the UNH-RVAT cross-flow turbine as designed (left)
+        and cleaned for simulation (right).}
 
@@
 
+\subsubsection{Mean velocity in three dimensions}
+
+% In this section, show mean velocity contours/quivers
+
+\subsubsection{Turbulence kinetic energy contours}
+
+% Compare contours of TKE for 2 simulation cases and experiments
+
 \subsubsection{Momentum recovery}
@@
 
+These results may also help develop new tip loss corrections for blade element
+type models, which currently only exist for axial-flow rotors.
+
 \begin{acknowledgments}

### 2015-09-03 paper.tex
Add time to 3-D vorticity caption
Add centering to 2-D vorticity fig
Make note of 3-D DS vortex

@@
 $-y$ side of the turbine associated with the induced vertical velocity field.
+The 3-D dynamic stall vortex also shows assymmetry about the $x$--$y$ mid-rotor
+plane; once again highlighting the importance of three-dimensional effects on
+wake dynamics.
 
@@
 \begin{figure}
+    \centering
+
     \includegraphics[width=0.8\textwidth]{figures/3D_vorticity_SA_964_10-threshold}
 
-    \caption{Iso-vorticity contours colored by the streamwise component of
-        vorticity for the 3-D Spalart--Allmaras case.}
+    \caption{Iso-vorticity contours (at $t=9.64$ s) colored by the streamwise
+        component of vorticity for the 3-D Spalart--Allmaras case.}
 

### 2015-09-08 makefigs.py
Can make meancontquiv figs from 3-D CFD cases

@@ -30,15 +30,18 @@ sst2dpr = SourceFileLoader("sst2dpr",
 sa2dpr = SourceFileLoader("sa2dpr", 
                   os.path.join(cfd_dirs["2-D"]["SpalartAllmaras"],
                   "modules", "processing.py")).load_module()
-sa3dpr = SourceFileLoader("sa3dpr", 
+sa3dpr = SourceFileLoader("modules.processing", 
                   os.path.join(cfd_dirs["3-D"]["SpalartAllmaras"],
                   "modules", "processing.py")).load_module()
-#sa3dpl = SourceFileLoader("sa3dpl", 
-#                  os.path.join(cfd_dirs["3-D"]["SpalartAllmaras"],
-#                  "modules", "plotting.py")).load_module()
-sst3dpr = SourceFileLoader("sst3dpr", 
+sa3dpl = SourceFileLoader("sa3dpl", 
+                  os.path.join(cfd_dirs["3-D"]["SpalartAllmaras"],
+                  "modules", "plotting.py")).load_module()
+sst3dpr = SourceFileLoader("modules.processing", 
                   os.path.join(cfd_dirs["3-D"]["kOmegaSST"],
                   "modules", "processing.py")).load_module()
+sst3dpl = SourceFileLoader("sst3dpl", 
+                  os.path.join(cfd_dirs["3-D"]["kOmegaSST"],
+                  "modules", "plotting.py")).load_module()
 #exppr = SourceFileLoader("exppl", os.path.join(exp_dir, "Modules", 
 #                         "plotting.py")).load_module()
                          
@@ -324,10 +327,10 @@ if __name__ == "__main__":
     save = True
     savetype = ".pdf"
     
-    plot_exp_perf()
-#    plot_exp_meancontquiv()
-#    plot_cfd_meancontquiv("kOmegaSST")
-#    plot_cfd_meancontquiv("SpalartAllmaras")
+#    plot_exp_perf()
+    plot_exp_meancontquiv()
+    plot_cfd_meancontquiv("kOmegaSST")
+    plot_cfd_meancontquiv("SpalartAllmaras")
 #    plot_cfd_u_profile()
 #    plot_verification()
 #    plot_profiles()

### 2015-09-08 makefigs.py
Add function for plotting experimental k contours

@@ -79,6 +79,15 @@ def plot_exp_meancontquiv():
     os.chdir(paper_dir)
     if save:
         plt.savefig("figures/meancontquiv_exp" + savetype)
+        
+def plot_exp_kcont():
+    os.chdir(exp_dir)
+    import Modules.plotting as plotting_exp
+    wm = plotting_exp.WakeMap(1.0)
+    wm.plot_k()
+    os.chdir(paper_dir)
+    if save:
+        plt.savefig("figures/kcont_exp" + savetype)
 
 def plot_cfd_meancontquiv(case="kOmegaSST"):
     """Plots wake mean velocity contours/quivers from 3-D CFD case."""
@@ -328,12 +337,14 @@ if __name__ == "__main__":
     savetype = ".pdf"
     
 #    plot_exp_perf()
-    plot_exp_meancontquiv()
-    plot_cfd_meancontquiv("kOmegaSST")
-    plot_cfd_meancontquiv("SpalartAllmaras")
+#    plot_exp_meancontquiv()
+#    plot_cfd_meancontquiv("kOmegaSST")
+#    plot_cfd_meancontquiv("SpalartAllmaras")
 #    plot_cfd_u_profile()
 #    plot_verification()
 #    plot_profiles()
 #    make_perf_bar_charts()
 #    make_recovery_bar_chart()
+    plot_exp_kcont()
+
     plt.show()

### 2015-09-08 makefigs.py
Add function for plotting CFD k contours

@@ -100,6 +100,17 @@ def plot_cfd_meancontquiv(case="kOmegaSST"):
     if save:
          plt.savefig("figures/meancontquiv_" + case + savetype)
          
+def plot_cfd_kcont(case="kOmegaSST"):
+    """Plot TKE from 3-D CFD case."""
+    os.chdir(cfd_dirs["3-D"][case])
+    if case == "SpalartAllmaras":
+        sa3dpl.plot_kcont()
+    elif case == "kOmegaSST":
+        sst3dpl.plot_kcont()
+    os.chdir(paper_dir)
+    if save:
+         plt.savefig("figures/kcont_" + case + savetype)
+         
 def plot_cfd_u_profile(case="kOmegaSST", dims="2-D"):
     """Plot mean streamwise velocity profile."""
     os.chdir(cfd_dirs[dims][case])
@@ -346,5 +357,7 @@ if __name__ == "__main__":
 #    make_perf_bar_charts()
 #    make_recovery_bar_chart()
     plot_exp_kcont()
+    plot_cfd_kcont("kOmegaSST")
+#    plot_cfd_kcont("SpalartAllmaras")
 
     plt.show()

### 2015-09-08 paper.tex
Add meancontquiv figure with subfigures

@@
 \usepackage[draft]{todonotes}
+\usepackage{subcaption}
 
@@
 
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering \includegraphics[clip, trim=0 0.1in 0 0.2in,
+        width=0.75\textwidth]{figures/meancontquiv_exp} \caption{Mean velocity
+            field at $x/D=1$ from experiments \cite{Bachant2015-RVAT-Re-dep-data}.}
+        \label{fig:meancontquiv-exp}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{\textwidth}
+        \centering \includegraphics[clip, trim=0 0.2in 0 0.15in,
+        width=0.8\textwidth]{figures/meancontquiv_kOmegaSST} \caption{Mean
+            velocity at $x/D=1$ computed by the 3-D SST model.}
+        \label{fig:meancontquiv-SST}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{\textwidth}
+        \centering \includegraphics[clip, trim=0 0.2in 0 0.15in,
+        width=0.8\textwidth]{figures/meancontquiv_SpalartAllmaras} \caption{Mean
+            velocity at $x/D=1$ computed by the 3-D SA model.}
+        \label{fig:meancontquiv-SA}
+    \end{subfigure}
+
+    \caption{Mean velocity from experiments and 3-D CFD cases. Solid gray lines
+        indicate turbine frontal area and dashed lines indicate experimental
+        measurement plane.}
+
+    \label{fig:mean-velocity}
+\end{figure}
+
 \subsubsection{Turbulence kinetic energy contours}

### 2015-09-08 paper.tex
Add TKE figure with subfigures

@@
 % Compare contours of TKE for 2 simulation cases and experiments
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+
+        \includegraphics[width=0.95\textwidth]{figures/kcont_exp}
+
+        \caption{Turbulence kinetic energy at $x/D=1$ from experiments
+            \cite{Bachant2015-RVAT-Re-dep-data}.}
+
+        \label{fig:kcont-exp}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+
+        \includegraphics[width=0.95\textwidth]{figures/kcont_kOmegaSST}
+
+        \caption{Turbulence kinetic energy at $x/D=1$ computed by the 3-D SST
+            model.}
+
+        \label{fig:kcont-SST}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+
+        \includegraphics[width=0.95\textwidth]{figures/kcont_SpalartAllmaras}
+
+        \caption{Turbulence kinetic energy at $x/D=1$ computed by the 3-D SA
+            model.}
+
+        \label{fig:kcont-SA}
+    \end{subfigure}
+
+    \caption{Turbulence kinetic energy from experiments and 3-D CFD cases. Solid
+        black lines indicate turbine frontal area.}
+
+    \label{fig:kcont}
+\end{figure}
 

### 2015-09-09 makefigs.py
Whitespace

@@ -326,9 +326,7 @@ def make_recovery_bar_chart():
     ax.set_ylabel(r"$\frac{U \, \mathrm{ transport} \times A_c}"
                   "{UU_\infty D^{-1}}$")
     ax.legend(loc="best", ncol=1)
-    fig.tight_layout()
-
-    
+    fig.tight_layout()    
     os.chdir(paper_dir)
     if save:
         fig.savefig("figures/mom_bar_graph"+savetype)
@@ -349,15 +347,15 @@ if __name__ == "__main__":
     
 #    plot_exp_perf()
 #    plot_exp_meancontquiv()
-#    plot_cfd_meancontquiv("kOmegaSST")
-#    plot_cfd_meancontquiv("SpalartAllmaras")
+    plot_cfd_meancontquiv("kOmegaSST")
+    plot_cfd_meancontquiv("SpalartAllmaras")
 #    plot_cfd_u_profile()
 #    plot_verification()
 #    plot_profiles()
 #    make_perf_bar_charts()
 #    make_recovery_bar_chart()
-    plot_exp_kcont()
-    plot_cfd_kcont("kOmegaSST")
+#    plot_exp_kcont()
+#    plot_cfd_kcont("kOmegaSST")
 #    plot_cfd_kcont("SpalartAllmaras")
 
     plt.show()

### 2015-09-09 paper.tex
Fix citation
Some spacing and wrapping

@@
     \begin{subfigure}[b]{\textwidth}
-        \centering \includegraphics[clip, trim=0 0.1in 0 0.2in,
-        width=0.75\textwidth]{figures/meancontquiv_exp} \caption{Mean velocity
-            field at $x/D=1$ from experiments \cite{Bachant2015-RVAT-Re-dep-data}.}
+        \centering
+
+        \includegraphics[clip, trim=0 0.1in 0 0.2in,
+        width=0.75\textwidth]{figures/meancontquiv_exp}
+
+        \caption{Mean velocity field at $x/D=1$ from experiments
+            \cite{Bachant2015-RVAT-Re-dep-data}.}
+
         \label{fig:meancontquiv-exp}
@@
     \begin{subfigure}[b]{\textwidth}
-        \centering \includegraphics[clip, trim=0 0.2in 0 0.15in,
-        width=0.8\textwidth]{figures/meancontquiv_kOmegaSST} \caption{Mean
-            velocity at $x/D=1$ computed by the 3-D SST model.}
+        \centering
+
+        \includegraphics[clip, trim=0 0.2in 0 0.15in,
+        width=0.8\textwidth]{figures/meancontquiv_kOmegaSST}
+
+        \caption{Mean velocity at $x/D=1$ computed by the 3-D SST model.}
+
         \label{fig:meancontquiv-SST}
@@
     \begin{subfigure}[b]{\textwidth}
-        \centering \includegraphics[clip, trim=0 0.2in 0 0.15in,
-        width=0.8\textwidth]{figures/meancontquiv_SpalartAllmaras} \caption{Mean
-            velocity at $x/D=1$ computed by the 3-D SA model.}
+        \centering
+
+        \includegraphics[clip, trim=0 0.2in 0 0.15in,
+        width=0.8\textwidth]{figures/meancontquiv_SpalartAllmaras}
+
+        \caption{Mean velocity at $x/D=1$ computed by the 3-D SA model.}
+
         \label{fig:meancontquiv-SA}
@@
     \caption{Weighted sum normalized momentum recovery terms for each CFD case
-        and experiments\cite{Bachant2015-JoT} at $x/D=1$.}
+        and experiments\cite{Bachant2015-RVAT-Re-dep-data} at $x/D=1$.}
 

### 2015-09-14 paper.tex
A word

@@
 In the pursuit of a fully sustainable energy profile, cross-flow (often
-vertical-axis) turbines (CFTs) still play a role. Despite extensive research and
-development in the 1970s through the early 1990s by groups like Sandia National
-Labs in the US \cite{Sutherland2012} and the National Research Council in Canada
-\cite{Para2002}, the vertical-axis cross-flow Darrieus turbines were all but
-abandoned for large scale commercial terrestrial wind power. Today, however,
-with the development of marine hydrokinetic (MHK) energy devices, CFTs are at
-the forefront, with the first grid-connected commercial device being a
+vertical-axis) turbines (CFTs) can still play a role. Despite extensive research
+and development in the 1970s through the early 1990s by groups like Sandia
+National Labs in the US \cite{Sutherland2012} and the National Research Council
+in Canada \cite{Para2002}, the vertical-axis cross-flow Darrieus turbines were
+all but abandoned for large scale commercial terrestrial wind power. Today,
+however, with the development of marine hydrokinetic (MHK) energy devices, CFTs
+are at the forefront, with the first grid-connected commercial device being a
 horizontal-axis CFT manufactured and installed by ORPC in Cobscook Bay, Maine

### 2015-09-14 paper.tex
A comma

@@
 momentum or vortex modeling, it remains difficult to predict the performance of
-cross-flow turbines in all cases---namely when solidity or blade chord to radius
-ratio is high---a common characteristic of the smaller CFTs. With computing
-power becoming evermore available and affordable, computational fluid dynamics
-(CFD) presents an attractive method for simulating cross-flow turbines since it
-only requires turbulence modeling, i.e., the machine is described from first
-principles. However, CFD can be computationally expensive when done in three
-dimensions, which may be necessary in some cases. There are many examples in the
-literature of 2-D cross-flow turbine simulations with widely varying results.
-Balduzzi et al. provides a summary of recent computations and an attempt to
-standardize a methodology for using Reynolds-averaged Navier--Stokes (RANS) to
-correctly simulate performance of a 2-D CFT \cite{Balduzzi2016}.
+cross-flow turbines in all cases---namely when solidity, or blade chord to
+radius ratio is high---a common characteristic of the smaller CFTs. With
+computing power becoming evermore available and affordable, computational fluid
+dynamics (CFD) presents an attractive method for simulating cross-flow turbines
+since it only requires turbulence modeling, i.e., the machine is described from
+first principles. However, CFD can be computationally expensive when done in
+three dimensions, which may be necessary in some cases. There are many examples
+in the literature of 2-D cross-flow turbine simulations with widely varying
+results. Balduzzi et al. provides a summary of recent computations and an
+attempt to standardize a methodology for using Reynolds-averaged Navier--Stokes
+(RANS) to correctly simulate performance of a 2-D CFT \cite{Balduzzi2016}.
 

### 2015-09-14 paper.tex
A bunch of work on intro

@@
 
-In this study we set out to model the wake of the University of New Hampshire's
-Reference Vertical-Axis (cross-flow) Turbine (UNH-RVAT) using a
-Reynolds-averaged Navier--Stokes numerical model.
-
-Note that studies in the literature generally focus on predicting the turbine
-loading and the local blade flow rather than the turbine wake. In contrast, we
-seek to model the near-wake flow.
-
-It has been shown that the momentum and energy transport processes in the
-near-wake of a high-solidity, low aspect ratio vertical-axis turbine are
-dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
-that a 2-D simulation would not correctly predict wake recovery, but is of
-interest to what extent this is true, since the computational feasibility of 2-D
-simulations is attractive.
-
 Modeling the boundary layer flows on cross-flow turbine blades is very
@@
 transition and separation can have dramatic influence on the blade loading and
-therefore the predicted turbine power output.
+therefore the predicted turbine power output. These challenges cast significant
+doubt on the ability of high-fidelity CFD modeling to replace wind tunnel or
+tank testing of physical models.
+
+\todo[inline]{Put survey of CFD accuracy here}
+
+In this study we set out to model the wake of the University of New Hampshire's
+Reference Vertical-Axis (cross-flow) Turbine (UNH-RVAT) using a
+Reynolds-averaged Navier--Stokes numerical model. Note that studies in the
+literature generally focus on predicting the turbine loading and the local blade
+flow rather than the turbine wake. In contrast, we seek to model the near-wake
+flow.
+
+The wake of this particular turbine is of interest since it has been shown that
+the neark-wake's momentum and energy transport processes are dominated by
+vertical advection \cite{Bachant2015-JoT}. It logically follows that a 2-D
+simulation would not correctly predict wake recovery, but is of interest to what
+extent this is true, since the computational feasibility of 2-D simulations is
+attractive.
 
 It is assumed that if the turbine power output and near-wake predictions from
-the numerical model match those of the experiments, that the flow field can be
+the numerical model match those of the experiments, the flow field can be
 inspected in greater detail, i.e., that the experiments will have been
@@
 help develop low-order wake generator models for use in turbine array modeling.
-For example, Araya et al. modeled the flow through a VAT array using leaky
-Rankine bodies \cite{Araya2014}. Shamsoddin and Porte-Agel modeled a 2-D CFT
-using large eddy simulation (LES) with an actuator line approach to impart force
-on the flow field \cite{Shamsoddin2014}.
+For example, Shamsoddin and Porte-Agel modeled a 2-D CFT using large eddy
+simulation (LES) with an actuator line approach to impart force on the flow
+field \cite{Shamsoddin2014}.
 
-Goude and Agren used a 2-D vortex method to simulate a farm of cross-flow
-turbines \cite{Goude2010}. Durrani et al. used 2-D CFD to model a group of
-cross-flow turbines, but did not compare with experimental results
-\cite{Durrani2011}.
+To date, little computational work has been done to attempt to design arrays of
+CFTs, despite their prospects for closer spacing compared with axial-flow
+turbines (AFTs). For example, Araya et al. modeled the flow through a VAT array
+using ``leaky Rankine body'' potential flow singularities \cite{Araya2014},
+which was able to rank relative performance of array configurations. Goude and
+Agren used a 2-D vortex method to simulate a farm of cross-flow turbines
+\cite{Goude2010}, though this was not validated with experiments. Durrani et al.
+used 2-D CFD to model a group of cross-flow turbines, observing higher power
+output for a staggered configuration, but also did not compare with experimental
+results \cite{Durrani2011}. Li and Calisal \cite{Li2010} used a 3-D vortex line
+method to show mutually improved power output from two adjacent turbines, though
+the simulations over-predicted the effects compared with experiments. Antheaume
+et al. \cite{Antheaume2008} used a blade element approach coupled with a 3-D
+RANS solver to also show how close spacing can improve power output of CFTs.
 
-Questions addressed:
+To summarize, the questions we hope to answer here are:
 
@@
 
-    \item Can 2-D RANS be used for array design?
+    \item Can or should 2-D RANS be used for array design?
 
-    \item Does 3-D RANS tell us more about what is happening, i.e, can it
-    ``interpolate'' the experimental results?
+    \item How accurately can 3-D RANS predict performance?
 
-    \item Does 3-D RANS realize the proper wake recovery proportions? This would
-    imply it could be a very accurate model if used in DES.
+    \item Can 3-D RANS ``interpolate'' the experimental results and provide insight to develop new low-order wake generators to represent CFTs?
+
+    \item Does 3-D RANS realize the correct proportions of wake recovery
+    mechanisms, i.e., are the 3-D blade-resolved results a good ``target'' for
+    those from a low-order model?
 
@@
 
-Goals:
-
-\begin{enumerate}
-
-    \item Evaluate modern computational fluid dynamics techniques for predicting
-    the dynamic loading on cross-flow turbine blades.
-
-    \item If possible, accurately predict the turbine from the perspective as a
-    wake generator, such that the results can be used to develop low-order
-    models.
-
-\end{enumerate}
+\section{Turbine description}
 

### 2015-09-14 paper.tex
More introduction work

@@
 Reference Vertical-Axis (cross-flow) Turbine (UNH-RVAT) using a
-Reynolds-averaged Navier--Stokes numerical model. Note that studies in the
+Reynolds-averaged Navier--Stokes numerical model. Though studies in the
 literature generally focus on predicting the turbine loading and the local blade
-flow rather than the turbine wake. In contrast, we seek to model the near-wake
-flow.
-
-The wake of this particular turbine is of interest since it has been shown that
-the neark-wake's momentum and energy transport processes are dominated by
-vertical advection \cite{Bachant2015-JoT}. It logically follows that a 2-D
-simulation would not correctly predict wake recovery, but is of interest to what
-extent this is true, since the computational feasibility of 2-D simulations is
-attractive.
+boundary layer, we seek to model the larger scale flow produced by the rotor.
+The wake of this particular turbine is of interest since it has been shown
+experimentally that the near-wake's momentum and energy transport processes are
+dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
+that a 2-D simulation would not correctly predict wake recovery, but is of
+interest to what extent this is true, since the computational feasibility of 2-D
+simulations is attractive.
 

### 2015-09-14 paper.tex
Wrap up time step discussion

@@
 Spalart--Allmaras model converged well with decreasing time step, leading to a
-final time step of 0.001 s. The results from the SST model show a local minimum,
-at $\Delta t = 0.002$ s, which was the final time step chosen to run the
-simulations.
-
-\todo[inline]{Find some examples in the literature where time steps are much
-    larger.}
+final time step of 0.001 s. The results from the SST model show a local minimum
+at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
+minimum was chosen as the final time step to run the simulations. Note that the
+SST model's convergence behavior may be due to its specific implementation in
+\textit{OpenFOAM}, and not indicative of the nature of the model equations.
+Verification studies for CFTs with this level of detail in the literature are
+not common, though the final time step is comparable to others
+\cite{Balduzzi2016}.
 

### 2015-09-14 paper.tex
Remove basically empty validation section; content in results
Simulation time --> simulated time

@@
 
-\section{Model validation}
-
-Models were validated against several open datasets.
-
 \section{Results}
@@
 Computations for the 3-D cases were run on 192 processes and took on the order
-of 1,000 CPU hours per second of simulation time. The 2-D simulations were run
-on a single processor and took on the order of one CPU hour per second.
+of 1,000 CPU hours per second of simulated time. The 2-D simulations were run on
+a single processor and took on the order of one CPU hour per second.
 

### 2015-09-14 paper.tex
Discuss profiles a bit

@@
 Mean velocity profiles at one turbine diameter downstream are shown in
-Figure~\ref{fig:profiles}. The 2-D results suffer from a blockage
-mismatch, i.e. keeping the proximity of the walls constant increases the
-blockage ratio.
+Figure~\ref{fig:profiles}. The 2-D results suffer from a blockage mismatch, i.e.
+keeping the proximity of the walls constant increases the blockage ratio. The
+3-D results, however, show good agreement with the experiments.
 
@@
 original velocity will be negligible.
+
+Both Spalart--Allmaras cases do a poor job predicting the turbulence kinetic
+energy in the flow, since it must be resolved as variance in the velocity field.
+The 2-D SST model does a good job predicting the peak in $k$ at $y/R=-1$, though
+is missing the smaller peak at $y/R=-1$. This is once again likely due to
+blockage issues, where local tip speed ratio is decreased, increasing the
+blades' instantaneous angle of attack at this location on the downstream
+passage. In contrast, the 3-D SST model predicts the $+y$ peak in turbulence
+kinetic energy very well, though the $-y$ peak magnitude is overpredicted by
+about 30\%. We also see some smearing of $k$ across the center of the rotor,
+which is likely due to exaggerated levels of the turbulent eddy viscosity.
 

### 2015-09-14 paper.tex
Discuss meancontquiv figs

@@
 
-% In this section, show mean velocity contours/quivers
+In order to visualize the mean velocity field, vector arrows for the mean
+cross-stream and vertical components are superimosed on top of contours of the
+streamwise component at $X/D=1$ in Figure~\ref{fig:mean-velocity}. Both CFD
+models predict the general structure of the mean velocity well, though the SA
+case has slightly larger vertical mean flows, which could be due to stronger tip
+vortex generation, or lower diffusivity compared with the SST model. Additional
+discrepancies between CFD and experiments may be due to the top slip boundary
+condition versus the experiment's free surface.
 

### 2015-09-14 paper.tex
Discuss k contours

@@
 
-% Compare contours of TKE for 2 simulation cases and experiments
+Turbulence kinetic energy contours for the experimental measurements and each
+CFD case at $x/D=1$ are presented in Figure~\ref{fig:kcont}. As seen in the
+profiles in Figure\ref{fig:profiles}, the SA model is resolving very little of
+the flow unsteadiness. In contrast, the SST model does a good job predicting the
+locations and magnitudes of various peaks in $k$. These are generated along the
+top of the turbine via tip vortex shedding, and the $-y$ side of the turbine via
+dynamic stall. We do however see the smearing effect from the dynamic stall
+vortex centered around $z/H=0$, which is likely more of an issue with wake
+evolution rather than wake generation.
+
 \begin{figure}

### 2015-09-14 paper.tex
Discuss recovery terms

@@
 
+We use the RANS models' eddy viscosity to calculate the turbulent transport via
+\begin{equation}
+    \text{Turb. trans.} = \nu_t \nabla^2 \vec{U},
+    \label{eq:turb-trans}
+\end{equation}
+which is a different approach from those taken on the experiments, where Reynolds stresses were measured, but $x$-derivatives were not:
+\begin{equation}
+    \text{Turb. trans. (exp.)} = 
+    -\left( 
+    \frac{\partial}{\partial y} \overline{u^\prime v^\prime}
+    +
+    \frac{\partial}{\partial z} \overline{u^\prime w^\prime}
+    \right).
+\end{equation}
+As such, we should not be surprised if the CFD models predict higher levels of
+turbulent transport than the experiments.
+
+Normalized weighted averages for each recovery term at $x/D=1$ are computed and
+multiplied by the cross-sectional area of the measurement plane, or the channel
+width in the 2-D cases. Results are shown in a bar chart in
+Figure~\ref{fig:recovery}. Consistent with the relatively large Reynolds number
+regime, viscous transport is essentially negligible compared with other
+mechanisms. Cross-stream advection---or the tendency of streamlines to diverge
+and reduce the streamwise momentum---produces a negative effect for all cases,
+though the 3-D SST model predicts significantly lower values. Vertical advection
+is by definition zero for the 2-D cases. The 3-D cases show varying
+results---with the SST model overpredicting and SA underpredicting the vertical
+velocity's effect on replenishing streamwise momentum.
+
+Turbulent transport and streamwise pressure gradient terms show the largest
+discrepancy between results. The 3-D SST case, despite doing a good job
+predicting turbulence kinetic energy, significantly overpredicts the turbulent
+transport term, while other CFD cases are comparable with the experiment. This
+seems to be balanced by a large adverse pressure gradient, which is also present
+to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
+cases create a wake where the pressure gradient is acting to accelerate the flow
+at $x/D=1$. Unfortunately, pressure data were not acquired from the experiment,
+though the 3-D delayed detached eddy simulation (DDES) of Boudreau and Dumas
+\cite{Boudreau2015} concur with the adverse pressure gradient condition.
+
 \begin{figure}[ht]

### 2016-03-19 paper.tex
Add NSF grant to acnkowledgments

@@
 \begin{acknowledgments}
-The authors acknowledge Vincent Neary and the Sandia National Laboratories Water
-Power Program for use of their Red Mesa high performance computing cluster.
+    The authors would acknowledge funding through a National Science Foundation
+    CAREER award (principal investigator Martin Wosnik, NSF 1150797, Energy for
+    Sustainability, program manager Gregory L. Rorrer). The authors also thank
+    Dr. Vincent S. Neary and the Sandia National Laboratories Water Power
+    Program for use of their Red Mesa high performance computing cluster.
 \end{acknowledgments}

### 2016-03-19 paper.tex
Make abstract past-tense

@@
 \begin{abstract}
-The wake of a vertical-axis cross-flow turbine (CFT) is modeled numerically via
-blade-resolved $k$--$\omega$ SST and Spalart--Allmaras RANS models in two and
-three dimensions. Results for each case are compared with experimental
-measurements of the turbine shaft power, overall drag, mean velocity, turbulence
-kinetic energy, and momentum transport terms in the near-wake at one diameter
-downstream. It is shown that 2-D simulations overpredict turbine loading and do
-not resolve mean vertical momentum transport, which plays an important role in
-the near-wake's momentum balance. The 3-D simulations fair better at predicting
-performance, with the Spalart--Allmaras model predictions being closest to the
-experiments. The SST model more accurately predicts the turbulence kinetic
-energy while the Spalart--Allmaras model more closely matches the momentum
-transport terms in the near-wake. These results show the potential of RANS as a
-design tool and a way to ``interpolate'' experimental flow field measurements.
+    The near-wake of a vertical-axis cross-flow turbine (CFT) was modeled
+    numerically via blade-resolved $k$--$\omega$ SST and Spalart--Allmaras RANS
+    models in two and three dimensions. Results for each case are compared with
+    experimental measurements of the turbine shaft power, overall drag, mean
+    velocity, turbulence kinetic energy, and momentum transport terms in the
+    near-wake at one diameter downstream. It was shown that 2-D simulations
+    overpredict turbine loading and do not resolve mean vertical momentum
+    transport, which plays an important role in the near-wake's momentum
+    balance. The 3-D simulations faired better at predicting performance, with
+    the Spalart--Allmaras model predictions being closest to the experiments.
+    The SST model more accurately predicted the turbulence kinetic energy while
+    the Spalart--Allmaras model more closely matched the momentum transport
+    terms in the near-wake. These results show the potential of blade-resolved
+    RANS as a design tool and a way to ``interpolate'' experimental flow field
+    measurements.
 \end{abstract}

### 2016-03-19 paper.tex
Add introduction from thesis CFD chapter

@@
 Despite the development of many simple engineering models based on blade element
-momentum or vortex modeling, it remains difficult to predict the performance of
-cross-flow turbines in all cases---namely when solidity, or blade chord to
-radius ratio is high---a common characteristic of the smaller CFTs. With
-computing power becoming evermore available and affordable, computational fluid
-dynamics (CFD) presents an attractive method for simulating cross-flow turbines
-since it only requires turbulence modeling, i.e., the machine is described from
-first principles. However, CFD can be computationally expensive when done in
-three dimensions, which may be necessary in some cases. There are many examples
-in the literature of 2-D cross-flow turbine simulations with widely varying
-results. Balduzzi et al. provides a summary of recent computations and an
-attempt to standardize a methodology for using Reynolds-averaged Navier--Stokes
-(RANS) to correctly simulate performance of a 2-D CFT \cite{Balduzzi2016}.
+momentum or vortex methods, it remains difficult to predict the performance of
+cross-flow turbines (CFTs, often vertical-axis turbines or VATs) in all
+cases---namely when solidity, or blade chord-to-radius ratio is high---a common
+characteristic of smaller rotors, and those designed for marine hydrokinetic
+(MHK) applications. With computing power becoming evermore available and
+affordable, computational fluid dynamics (CFD) has become an attractive method
+for predicting the performance of cross-flow turbines since it only requires
+that turbulence be modeled, i.e., the physics can be described from theoretical
+true first principles by solving the Navier--Stokes equations using a boundary
+layer resolving body-fitted grid. However, CFD can be computationally expensive
+when done in three dimensions, which may be necessary in some cases.
 
-Modeling the boundary layer flows on cross-flow turbine blades is very
-challenging due to the dynamically changing inflow and angle of attack---which
-often exceeds static stall values and causes dynamic stall. Furthermore, the
-ability to predict the occurrence and interdependence of boundary layer
-transition and separation can have dramatic influence on the blade loading and
-therefore the predicted turbine power output. These challenges cast significant
-doubt on the ability of high-fidelity CFD modeling to replace wind tunnel or
+There are many examples in the literature of 2-D cross-flow turbine simulations
+with widely varying results. Balduzzi et al.~\cite{Balduzzi2016} provides
+a summary of recent efforts and an attempt to standardize a methodology for
+using Reynolds-averaged Navier--Stokes (RANS) to correctly predict performance
+of a 2-D CFT. Howell et al.~\cite{Howell2010}, performed both 2-D and 3-D
+simulations of a high solidity cross-flow turbine using a $k$--$\epsilon$
+renormalization group (RNG) turbulence model. The results from the 2-D
+simulations over-predicted power coefficient, while the 3-D case matched well
+with wind tunnel measurements near the tip speed ratio of maximum power. In
+general, 3-D simulations are less common, but have begun to appear more
+frequently recently---a testament to the progress towards higher computing
+power.
+
+An overview of 3-D blade-resolved cross-flow turbine simulations reported in the
+literature is presented in Table~\ref{tab:cfd-refs}. The $k$--$\omega$ SST RANS
+turbulence model is shown to be a popular choice due to its success in
+predicting flows with adverse pressure gradients and
+separation~\cite{Menter2003}. Higher fidelity methods that resolve the large
+scales of turbulence, such as large-eddy and detached-eddy simulation have also
+been used. Note that for all the studies listed, model validation was either
+done for performance or wake predictions---not both---and in some case omitted
+entirely.
+
+\begin{table}
+    \centering
+    \begin{tabular}{c|c|c|c}
+        Author & Turbulence modeling & Perf. val. & Wake val. \\ 
+        \hline
+        Alaimo et al.~\cite{Alaimo2015} & $k$--$\epsilon$ & N/A & N/A \\ 
+        Marsh et al.~\cite{Marsh2015} & SST & \cite{Rawlings2008} & N/A \\ 
+        Orlandi et al.~\cite{Orlandi2015} & SST & \cite{Akins1989,Mertens2003} & N/A \\ 
+        Lam \& Peng~\cite{Lam2016} & SST RANS \& IDDES\tablefootnote{Improved delayed detached-eddy simulation.} & N/A & \cite{Tescione2014} \\ 
+        Nini et al.~\cite{Nini2014} & Spalart--Allmaras & N/A & \cite{Battisti2011} \\ 
+        Boudreau \& Dumas~\cite{Boudreau2015} & DDES\tablefootnote{Delayed detached-eddy simulation.} & N/A & N/A \\ 
+        Li et al.~\cite{Li2013} & SST \& Smagorinsky--Lilly LES & \cite{McLaren2011} & N/A \\ 
+        Howell et al.~\cite{Howell2010} & $k$--$\epsilon$ RNG\tablefootnote{Renormalization group.} & \cite{Howell2010} & N/A
+    \end{tabular}
+
+    \caption{Selected 3-D blade-resolved cross-flow turbine simulations reported
+        in the literature, turbulence modeling employed, and performance and/or wake
+        studies used for validation. Note the Li et al. study used periodic boundary
+        conditions and is technically considered 2.5-D.}
+
+    \label{tab:cfd-refs}
+\end{table}
+
+Modeling the boundary layer flows on cross-flow turbine blades is essential to
+predicting the blade loading. This flow condition presents a challenge due to
+the dynamically changing inflow velocity and angle of attack---which often
+exceeds static stall values and causes dynamic stall. Furthermore, the ability
+to predict the occurrence and interdependence of boundary layer transition to
+turbulence and separation can have dramatic influence on the blade loading and
+therefore the predicted turbine power output. These challenges present
+significant obstacles to the prospect of using CFD to replace wind tunnel or
 tank testing of physical models.
 
-\todo[inline]{Put survey of CFD accuracy here}
+To date, little computational work has been done to attempt to design arrays of
+CFTs, despite their prospects for closer spacing compared with axial-flow
+turbines (AFTs). For example, Araya et al.~\cite{Araya2014} modeled the flow
+through a VAT array using ``leaky Rankine body'' potential flow singularities,
+which was able to rank relative---though not absolut---performance of array
+configurations. Goude and Agren~\cite{Goude2010} used a 2-D vortex method to
+simulate a farm of cross-flow turbines, though this was not validated with
+experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a
+group of cross-flow turbines, observing higher power output for a staggered
+configuration, but also did not compare with experimental results. Giorgetti
+et al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
+turbine pairs inspired by Dabiri~\cite{Dabiri2011}, but again experimental
+validation was not performed. Li and Calisal \cite{Li2010} used a 3-D vortex
+line method to show mutually improved power output from two adjacent turbines,
+though the simulations over-predicted the effects compared with experiments.
+Antheaume et al.~\cite{Antheaume2008} used a blade element approach
+coupled with a 3-D RANS solver to also show how close spacing can improve power
+output of CFTs.
 
-In this study we set out to model the wake of the University of New Hampshire's
-Reference Vertical-Axis (cross-flow) Turbine (UNH-RVAT) using a
-Reynolds-averaged Navier--Stokes numerical model. Though studies in the
-literature generally focus on predicting the turbine loading and the local blade
-boundary layer, we seek to model the larger scale flow produced by the rotor.
-The wake of this particular turbine is of interest since it has been shown
+In this study we set out to model the performance and near-wake of the high
+solidity University of New Hampshire Reference Vertical Axis Turbine (UNH-RVAT)
+using the Reynolds-averaged Navier--Stokes equations discretized via the finite
+volume method. Though studies in the literature generally focus on predicting
+the turbine loading and the local blade boundary layer, we seek to model both
+this and the larger scale flow produced by the rotor, i.e., the near-wake, which
+is of interest for this particular turbine since it has been shown
 experimentally that the near-wake's momentum and energy transport processes are
 dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
-that a 2-D simulation would not correctly predict wake recovery, but is of
-interest to what extent this is true, since the computational feasibility of 2-D
-simulations is attractive.
+that a 2-D simulation, which omits the vertical dimension, would not correctly
+predict wake recovery and turbine--turbine interaction. However, it is of
+interest to determine how wrong a 2-D model may be, since the computational
+feasibility is attractive.
 
@@
 inspected in greater detail, i.e., that the experiments will have been
-``interpolated''. This will give us access to unmeasured quantities, e.g.
-pressure, and allow us to see where the dominant flow structures originate and
-help develop low-order wake generator models for use in turbine array modeling.
-For example, Shamsoddin and Porte-Agel modeled a 2-D CFT using large eddy
-simulation (LES) with an actuator line approach to impart force on the flow
-field \cite{Shamsoddin2014}.
-
-To date, little computational work has been done to attempt to design arrays of
-CFTs, despite their prospects for closer spacing compared with axial-flow
-turbines (AFTs). For example, Araya et al. modeled the flow through a VAT array
-using ``leaky Rankine body'' potential flow singularities \cite{Araya2014},
-which was able to rank relative performance of array configurations. Goude and
-Agren used a 2-D vortex method to simulate a farm of cross-flow turbines
-\cite{Goude2010}, though this was not validated with experiments. Durrani et al.
-used 2-D CFD to model a group of cross-flow turbines, observing higher power
-output for a staggered configuration, but also did not compare with experimental
-results \cite{Durrani2011}. Li and Calisal \cite{Li2010} used a 3-D vortex line
-method to show mutually improved power output from two adjacent turbines, though
-the simulations over-predicted the effects compared with experiments. Antheaume
-et al. \cite{Antheaume2008} used a blade element approach coupled with a 3-D
-RANS solver to also show how close spacing can improve power output of CFTs.
-
-To summarize, the questions we hope to answer here are:
+``interpolated.'' This will provide access to unmeasured quantities, e.g.,
+pressure, and allow inspection into where the dominant flow structures
+originate. This will ultimately help develop and evaluate low-order wake
+generator models for use in turbine array modeling. In summary, the questions we
+hope to answer here are:
 
 \begin{enumerate}
-
-    \item Can or should 2-D RANS be used for array design?
+    \item Can 2-D RANS be used for individual turbine and/or array design?
 
@@
 
-    \item Can 3-D RANS ``interpolate'' the experimental results and provide insight to develop new low-order wake generators to represent CFTs?
+    \item Can 3-D RANS ``interpolate'' the experimental results and provide
+    insight to develop new low-order wake generators to represent CFTs?
 
@@
     mechanisms, i.e., are the 3-D blade-resolved results a good ``target'' for
-    those from a low-order model?
-
+    those a low-order model should produce?
 \end{enumerate}

### 2016-03-19 paper.tex
Copy stuff from CFD thesis chapter up to results

@@
 
-\section{Numerical models}
+\section{Numerical setup}
 
-The flow field was modeled using the Reynolds-averaged Navier--Stokes equations,
-employing two different turbulence models---Menter's $k$--$\omega$ SST
-\cite{Menter1994} and the Spalart--Allmaras (SA) one equation model
-\cite{Spalart1992}. The SST model was chosen due to its prominence in the
-literature for simulating separating flows, which we assumed to be present in
-the current problem in the form of dynamic stall. The SA model was shown by
-Ferreira et al. \cite{Ferreira2007} to match experimental particle image
+In this study the UNH-RVAT baseline tow tank experiment was simulated using a
+mean rotor tip speed ratio $\lambda=1.9$, which corresponds to the maximum
+measured power coefficient. The tow speed or free stream velocity $U_\infty=1.0$
+m/s gives a turbine diameter Reynolds number $Re_D \approx 10^6$, which
+corresponds to the $Re$-independent state for both performance and near-wake
+characteristics, as determined from previous experimental measurements
+\cite{Bachant2014, Bachant2016-Energies}.
+
+The turbines CAD geometry was prepared or ``cleaned'' for CFD by removing
+details determined to be unnecessary, e.g., screw heads and axial shaft grooves,
+which would complicate the meshing process and contribute very little to the
+overall loading or flow modulation. A drawing of the physically and numerically
+modeled geometries is presented in Figure~\ref{fig:cfd-cad}.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[clip, trim=0 1in 0 1in, width=0.8\textwidth]{BR-CFD_CAD}
+
+    \caption{CAD drawings of the UNH-RVAT cross-flow turbine as designed (left)
+        and cleaned for simulation (right).}
+
+    \label{fig:cfd-cad}
+\end{figure}
+
+To close the RANS equations, two different turbulence models were
+used---Menter's $k$--$\omega$ SST \cite{Menter1994} and the Spalart--Allmaras
+(SA) one equation model \cite{Spalart1992}. Both closures use the eddy-viscosity
+approach---the SA employing a single additional scalar transport equation and
+the SST two. The SST model was chosen due to its prominence in the literature
+for simulating separating flows, which we assumed to be present in the current
+problem in the form of dynamic stall. The SA model was shown by Ferreira
+et al.~\cite{Ferreira2007} to match experimental particle image
 velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
@@
 bubble and subsequent boundary layer transition to turbulence at Reynolds
-numbers similar to ours.
+numbers similar to those investigated here.
 
@@
 
-The computational domain is a rectangular volume 3.66 m long, 3.66 m wide, and
+The computational domain was a rectangular volume 3.66 m long, 3.66 m wide, and
 2.44 m tall (for 3-D), with the turbine located 1.52 m from the inlet, and
-centered vertically with a vertical axis. with a cylindrical sliding mesh
-interface, where the turbine zone rotates at a mean tip speed ratio
-$\lambda=1.9$ with a sinusoidal oscillation at the blade passage
-frequency---with and amplitude of 0.19 and the first peak at 0.7 radians---to
-mimic the experimental data. The 2-D mesh overview is shown in
-Figure~\ref{fig:mesh} and the blade mesh is shown in
+centered vertically with a vertical axis, designed to match the tow tank
+dimensions for comparison with previous experiments. The rotor geometry was
+located at the center of a cylindrical sliding mesh interface, which rotated at
+a mean tip speed ratio $\lambda=1.9$ with a sinusoidal oscillation at the blade
+passage frequency---with and amplitude of 0.19 and the first peak at 1.4
+radians---to mimic the slight deviation from the mean tip speed ratio observed
+in the experiments. The 2-D mesh overview is shown in
+Figure~\ref{fig:2d-br-mesh} and the blade mesh detail is shown in
 Figure~\ref{fig:blade-mesh}.
@@
 
-3-D computer aided design (CAD) files of the turbine are available from
-\cite{Bachant2014-RVAT-CAD}.
+\begin{figure}
+    \centering
 
-\begin{figure}[ht]    
-    \includegraphics[width=0.8\textwidth]{figures/2d_mesh}
+    \includegraphics[width=0.8\textwidth]{BR-CFD_2D_mesh}
 
@@
 
-    \label{fig:mesh}
+    \label{fig:2d-br-mesh}
 \end{figure}
 
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{figures/2D_blade_mesh_closeup}
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.7\textwidth]{BR-CFD_2D_blade_mesh_closeup}
 
@@
 \textit{pimpleDyMFoam} uses a hybrid PISO-SIMPLE algorithm for pressure-velocity
-coupling and is compatible with dynamic meshes.
+coupling and is compatible with dynamic meshes. An Euler scheme was used to
+advance the simulation forward in time. The case files required to replicate the
+simulations are available from \cite{Bachant2016-UNH-RVAT-2D-OpenFOAM-SST,
+    Bachant2016-UNH-RVAT-2D-OpenFOAM-SA, Bachant2016-UNH-RVAT-3D-OpenFOAM-SST,
+    Bachant2016-UNH-RVAT-3D-OpenFOAM-SA}.
 
@@
 Initial and boundary conditions were set to match those of the tow tank as well
-as possible. The velocity at the bottom and side walls was fixed to 1 m/s to
-match the tow tank case, while the top boundary condition was a slip velocity
-condition. Note that in 2-D the top and bottom boundary conditions are
-``empty,'' which is an OpenFOAM convention.
+as possible. The velocity at the inlet, bottom, and side walls was fixed to 1
+m/s to match the tow tank case, while the top boundary condition was a slip
+velocity condition. Pressure was held fixed at the outlet. Note that in 2-D the
+top and bottom boundary conditions are ``empty,'' which is an OpenFOAM
+convention to indicate two-dimensionality.
+
+Since the mesh was only refined next to the blade surfaces in order to resolve
+the boundary layer profile, no wall functions were used. However, wall functions
+were used for the other turbine surfaces, i.e., struts, hub, and shaft, and for
+the tank sidewalls, bottom, and top.
 
@@
 
-The $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified for
-convergence of the turbine mean power coefficient with respect to grid spacing
-and timestep. The grid topology was fixed, but the number of cells per unit
-length were scaled proportionally, maintaining the same background mesh cell
-aspect ratio. Results for this parameter sweep are shown in
-Figure~\ref{fig:verification}, from which the final number of streamwise
-grid points $N_x = 70$ was chosen. This corresponds to a total cell count of
-approximately $5 \times 10^4$ for the 2-D cases and 16 million for the 3-D case.
+Both the $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified
+for convergence of the turbine mean power coefficient with respect to grid
+spacing and time step using 2-D domains. The grid topology was fixed, but the
+number of cells per domain length was scaled proportionally, maintaining the
+same background mesh cell aspect ratio. Results for this parameter sweep are
+shown in Figure~\ref{fig:2d-br-verification}, from which the final number of
+streamwise grid points $N_x = 70$ was chosen. This corresponds to a total cell
+count of approximately $5 \times 10^4$ for the 2-D cases and 16 million for the
+3-D case.
 
-\begin{figure}[ht]
+\begin{figure}
     \centering
 
-    \includegraphics[width=0.9\textwidth]{figures/verification}
+    \includegraphics[width=0.9\textwidth]{BR-CFD_verification}
 
@@
 
-    \label{fig:verification}
+    \label{fig:2d-br-verification}
 \end{figure}
 
-Time step dependence was studied using the $N_x=70$ grid, the results from which
-are shown in Figure~\ref{fig:verification}. It was seen that the
+Time step dependence was evaluated using the 2-D $N_x=70$ grid, the results from
+which are shown in Figure~\ref{fig:2d-br-verification}. It was seen that the
 Spalart--Allmaras model converged well with decreasing time step, leading to a

### 2016-03-19 paper.tex
Add results and conclusions from CFD chapter

@@
 
-Computations for the 3-D cases were run on 192 processes and took on the order
-of 1,000 CPU hours per second of simulated time. The 2-D simulations were run on
-a single processor and took on the order of one CPU hour per second.
+Turbine operation in all cases was simulated for 10 seconds, or approximately
+six rotor revolutions. Computations for the 3-D cases were run on 192 processes
+(24 nodes $\times$ 8 cores each) and took on the order of 1,000 CPU hours per
+second of simulated time. The 2-D simulations were run on a single processor and
+took on the order of one CPU hour per second.
+
+Flow statistics were calculated over the time interval spanning $5$--$10$
+seconds, or approximately three rotor revolutions, from time series downsampled
+to 50 Hz. It is assumed that the downsampling frequency is high enough above the
+blade passage frequency such that differences from the variance (for computing
+turbulence statistics) in the original velocity will be negligible.
 
@@
 
-Predictions for both the mean rotor power and drag coefficients are shown in
-Figure~\ref{fig:perf-comp}. In general, the 2-D CFD cases both overpredict
-turbine loading, which is likely due to their increased blockage ratio,
-unresolved blade end effects, and lack of blade support struts.
+Mean performance quantities---power and overall rotor drag coefficients---were
+calculated for the interval after two full rotor revolutions, beyond which
+values had reached an approximately periodic state. Predictions for both the
+mean rotor power and drag coefficients are shown in
+Figure~\ref{fig:br-cfd-perf-bar-chart}. In general, the 2-D CFD cases both
+significantly overpredict turbine loading and therefore mechanical power output,
+which is likely due to their increased blockage ratio, unresolved blade end
+effects, and lack of blade support struts.
 
-The 3-D simulations fair better at predicting the experimental measurements,
-with the Spalart--Allmaras model performing relatively better. The apparent
-overprediction of rotor drag coefficient could be an effect of the experimental
-procedure, where the ``tare drag'' from the turbine mounting structure was
-measured without a turbine installed, then subtracted in post-processing.
-Technically the local flow field will have changed, and hence the tare drag will
-change as well.
+The 3-D simulations fair better at predicting the experimental measurements. The
+Spalart--Allmaras model's mean power coefficient was 0.24 while the SST's was
+0.33, which represent a 6\% under- and 30\% overprediction, respectively,
+compared with the experiments. The apparent overprediction of rotor drag
+coefficient could be an effect of the experimental procedure, where the ``tare
+drag'' from the turbine mounting structure was measured without a turbine
+installed, then subtracted in post-processing. Technically the local flow field
+will have changed, and hence the tare drag will change as well.
 
-\begin{figure}[ht]
+\begin{figure}
     \centering
 
-    \includegraphics[width=0.9\textwidth]{figures/perf_bar_chart}
+    \includegraphics[width=0.95\textwidth]{BR-CFD_perf_bar_chart}
 
@@
 
-    \label{fig:perf-comp}
+    \label{fig:br-cfd-perf-bar-chart}
 \end{figure}
@@
 
-Visualizing the complicated wake generated by the turbine is presented for the
-2-D and 3-D Spalart--Allmaras cases in Figure~\ref{fig:vorticity-3D} and
-Figure~\ref{fig:vorticity-3D}, respectively. It can be seen how the upstream
-blade---as it turns back into the streamwise direction---is shedding a large
-amount of spanwise vorticity due to the separated flow. In the 3-D case, strong
-tip vortices are also present, which trace the ``contracting'' wake flow on the
-$-y$ side of the turbine associated with the induced vertical velocity field.
-The 3-D dynamic stall vortex also shows assymmetry about the $x$--$y$ mid-rotor
-plane; once again highlighting the importance of three-dimensional effects on
-wake dynamics.
+Visualizations of the complex vorticity field generated by the turbine are
+presented for the 2-D and 3-D Spalart--Allmaras cases in
+Figure~\ref{fig:br-vorticity-3d} and Figure~\ref{fig:br-vorticity-3d},
+respectively. It can be seen how the upstream blade---as it turns back into the
+streamwise direction---is shedding a large amount of spanwise vorticity due to
+the separated flow. In the 3-D case, strong tip vortices are also present, which
+trace the ``contracting'' wake flow on the $-y$ side of the turbine associated
+with the induced vertical velocity field. The 3-D dynamic stall vortex also
+shows asymmetry about the $x$--$y$ mid-rotor plane; once again highlighting the
+importance of three-dimensional effects on wake dynamics.
 
 \begin{figure}
-    \includegraphics[width=0.65\textwidth]{figures/2D_vorticity_SA_964}
+    \centering
+
+    \includegraphics[width=0.75\textwidth]{BR-CFD_2D_vorticity_SA_964}
 
@@
 
-    \label{fig:vorticity-2D}
+    \label{fig:br-vorticity-2d}
 \end{figure}
@@
 
-    \includegraphics[width=0.8\textwidth]{figures/3D_vorticity_SA_964_10-threshold}
+    \includegraphics[width=0.75\textwidth]{BR-CFD_3D_vorticity_SA_964_10-threshold}
 
@@
 
-    \label{fig:vorticity-3D}
+    \label{fig:br-vorticity-3d}
 \end{figure}
@@
 Mean velocity profiles at one turbine diameter downstream are shown in
-Figure~\ref{fig:profiles}. The 2-D results suffer from a blockage mismatch, i.e.
-keeping the proximity of the walls constant increases the blockage ratio. The
-3-D results, however, show good agreement with the experiments.
+Figure~\ref{fig:br-cfd-profiles}. The 2-D results suffer from a blockage
+mismatch, i.e., keeping the proximity of the walls constant increases the
+blockage ratio. The 3-D results, however, show good agreement with the
+experiments.
 
@@
 
-    \includegraphics[width=0.95\textwidth]{figures/profiles}
+    \includegraphics[width=0.95\textwidth]{BR-CFD_profiles}
 
@@
 
-    \label{fig:profiles}
+    \label{fig:br-cfd-profiles}
 \end{figure}
 
-Turbulence kinetic energy profiles are also shown in Figure~\ref{fig:profiles}.
-The turbulence kinetic energy is calculated as
+Turbulence kinetic energy profiles are also shown in
+Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy is calculated as
 \begin{equation}
@@
 where $U^\prime = U - \overline{U}$ and $k_{\mathrm{RA}}$ is the kinetic energy
-calculated by the turbulence model, which is zero for the SA model. Note that
-the statistics of the Reynolds-averaged velocity are calculated from a time
-series that has been downsampled to 50 Hz. It is assumed that this is far enough
-from the blade passage frequency that differences from the variance in the
-original velocity will be negligible.
+calculated by the turbulence model, which is zero for the SA model.
 
@@
 cross-stream and vertical components are superimosed on top of contours of the
-streamwise component at $X/D=1$ in Figure~\ref{fig:mean-velocity}. Both CFD
-models predict the general structure of the mean velocity well, though the SA
-case has slightly larger vertical mean flows, which could be due to stronger tip
-vortex generation, or lower diffusivity compared with the SST model. Additional
-discrepancies between CFD and experiments may be due to the top slip boundary
-condition versus the experiment's free surface.
+streamwise component at $X/D=1$ in Figure~\ref{fig:br-cfd-mean-velocity}. Both
+CFD models predict the general structure of the mean velocity well, though the
+SA case has slightly larger vertical mean flows, which could be due to stronger
+tip vortex generation, or lower diffusivity compared with the SST model.
+Additional discrepancies between CFD and experiments may be due to the top slip
+boundary condition versus the experiment's free surface.
 
@@
 
-        \includegraphics[clip, trim=0 0.1in 0 0.2in,
-        width=0.75\textwidth]{figures/meancontquiv_exp}
+        \includegraphics[clip, trim=0 0.1in 0.3in 0.2in,
+        width=0.72\textwidth]{RVAT-Re-dep_meancontquiv_10}
 
         \caption{Mean velocity field at $x/D=1$ from experiments
-            \cite{Bachant2015-RVAT-Re-dep-data}.}
+            \cite{Bachant2016-RVAT-Re-dep}.}
 
-        \label{fig:meancontquiv-exp}
+        \label{fig:br-cfd-meancontquiv-exp}
     \end{subfigure}
@@
         \includegraphics[clip, trim=0 0.2in 0 0.15in,
-        width=0.8\textwidth]{figures/meancontquiv_kOmegaSST}
+        width=0.8\textwidth]{BR-CFD_meancontquiv_kOmegaSST}
 
@@
         \includegraphics[clip, trim=0 0.2in 0 0.15in,
-        width=0.8\textwidth]{figures/meancontquiv_SpalartAllmaras}
+        width=0.8\textwidth]{BR-CFD_meancontquiv_SpalartAllmaras}
 
@@
 
-    \label{fig:mean-velocity}
+    \label{fig:br-cfd-mean-velocity}
 \end{figure}
@@
 Turbulence kinetic energy contours for the experimental measurements and each
-CFD case at $x/D=1$ are presented in Figure~\ref{fig:kcont}. As seen in the
-profiles in Figure\ref{fig:profiles}, the SA model is resolving very little of
-the flow unsteadiness. In contrast, the SST model does a good job predicting the
-locations and magnitudes of various peaks in $k$. These are generated along the
-top of the turbine via tip vortex shedding, and the $-y$ side of the turbine via
-dynamic stall. We do however see the smearing effect from the dynamic stall
-vortex centered around $z/H=0$, which is likely more of an issue with wake
-evolution rather than wake generation.
+CFD case at $x/D=1$ are presented in Figure~\ref{fig:br-cfd-kcont}. As seen in
+the profiles in Figure~\ref{fig:br-cfd-profiles}, the SA model is resolving very
+little of the flow unsteadiness. In contrast, the SST model does a good job
+predicting the locations and magnitudes of various peaks in $k$. These are
+generated along the top of the turbine via tip vortex shedding, and the $-y$
+side of the turbine via dynamic stall. We do however see the smearing effect
+from the dynamic stall vortex centered around $z/H=0$, which is likely more of
+an issue with wake evolution rather than wake generation.
 
@@
 
-        \includegraphics[width=0.95\textwidth]{figures/kcont_exp}
+        \includegraphics[width=0.95\textwidth]{RVAT-Re-dep_k_contours_10}
 
         \caption{Turbulence kinetic energy at $x/D=1$ from experiments
-            \cite{Bachant2015-RVAT-Re-dep-data}.}
+            \cite{Bachant2016-RVAT-Re-dep}.}
 
@@
 
-        \includegraphics[width=0.95\textwidth]{figures/kcont_kOmegaSST}
+        \includegraphics[width=0.95\textwidth]{BR-CFD_kcont_kOmegaSST}
 
@@
 
-        \includegraphics[width=0.95\textwidth]{figures/kcont_SpalartAllmaras}
+        \includegraphics[width=0.95\textwidth]{BR-CFD_kcont_SpalartAllmaras}
 
@@
 
-    \label{fig:kcont}
+    \label{fig:br-cfd-kcont}
 \end{figure}
@@
 $\partial U / \partial x$---following Bachant and
-Wosnik\cite{Bachant2015-JoT}---and compute each term at $X/D = 1$ to compare
+Wosnik~\cite{Bachant2015-JoT}---and compute each term at $X/D = 1$ to compare
 with the experimental results.
@@
 \end{equation}
-which is a different approach from those taken on the experiments, where Reynolds stresses were measured, but $x$-derivatives were not:
+which is a different approach from those taken on the experiments, where
+Reynolds stresses were measured, but $x$-derivatives were not:
 \begin{equation}
@@
 width in the 2-D cases. Results are shown in a bar chart in
-Figure~\ref{fig:recovery}. Consistent with the relatively large Reynolds number
-regime, viscous transport is essentially negligible compared with other
+Figure~\ref{fig:br-cfd-recovery}. Consistent with the relatively large Reynolds
+number regime, viscous transport is essentially negligible compared with other
 mechanisms. Cross-stream advection---or the tendency of streamlines to diverge
@@
 to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
-cases create a wake where the pressure gradient is acting to accelerate the flow
-at $x/D=1$. Unfortunately, pressure data were not acquired from the experiment,
-though the 3-D delayed detached eddy simulation (DDES) of Boudreau and Dumas
-\cite{Boudreau2015} concur with the adverse pressure gradient condition.
+cases create a wake where the pressure gradient is acting to slightly accelerate
+the flow at $x/D=1$. Unfortunately, pressure data were not acquired from the
+experiment, though the 3-D delayed detached eddy simulation (DDES) of Boudreau
+and Dumas \cite{Boudreau2015} concur with the adverse pressure gradient
+condition.
 
-\begin{figure}[ht]
+\begin{figure}
     \centering
 
-    \includegraphics[width=0.9\textwidth]{figures/mom_bar_graph}
+    \includegraphics[width=0.9\textwidth]{BR-CFD_mom_bar_graph}
 
     \caption{Weighted sum normalized momentum recovery terms for each CFD case
-        and experiments\cite{Bachant2015-RVAT-Re-dep-data} at $x/D=1$.}
+        and experiments\cite{Bachant2016-RVAT-Re-dep} at $x/D=1$.}
 
-    \label{fig:recovery}
+    \label{fig:br-cfd-recovery}
 \end{figure}
@@
 A cross-flow turbine was modeled using the $k$--$\omega$ SST and
-Spalart--Allmaras Reynolds-averaged Navier--Stokes turbulence models to test
+Spalart--Allmaras (SA) Reynolds-averaged Navier--Stokes turbulence models to test
 their abilities to predict turbine performance and near-wake dynamics. It was
@@
 
-These results may also help develop new tip loss corrections for blade element
-type models, which currently only exist for axial-flow rotors.
+The 3-D blade-resolved RANS simulations predicted turbine performance and
+near-wake quite well, where the SA model results were closest to the
+experimentally measured performance. The SA model's effectiveness gives hope for
+the prospect of replacing some physical modeling with blade-resolved CFD.
+However, the  overprediction of power coefficient by the SST model highlights
+the level of technological and economic risk involved with CFD. That is to say,
+the SST model, despite being considered relatively trustworthy in these flow
+conditions, would overestimate power output by 30\%.
+
+Both the SST and SA models did a good job predicting the mean velocity field at
+one turbine diameter downstream, but the SST model was more effective at
+predicting turbulence kinetic energy, since it is solved for in the turbulence
+model equations. Streamwise momentum recovery terms were computed from the CFD
+results over an entire cross-section of the domain at $x/D=1$. Values for
+transport due to the mean pressure gradient and turbulent fluctuations varied a
+lot between the two turbulence models. Both models, however, we able to at least
+qualitatively resolve the vertical velocity field, which will be crucial to
+predicting the performance of closely spaced CFTs.
+
+The effectiveness at predicting the mean velocity field gives credibility to the
+prospect of using the computed flow field to interpolate the experimental
+results, such that the CFD results can be used as a target for a low-order wake
+generator or force parameterization. These results may also help develop new tip
+loss corrections for blade element type models, which currently only exist for
+axial-flow rotors, since they provide access to the pressure and shear forces
+over the entire blade surface, which were not measured experimentally.
+Ultimately, however, the computational cost of 3-D simulations---about 10,000
+CPU hours per operating case---may be too expensive to be used for CFT
+engineering work, especially considering the uncertainty involved compared with
+physical model studies. Since the 3-D wake results appeared to be only weakly
+asymmetrical in the $x$--$y$ plane about $z/H=0$, it may be possible to reduce
+computing power by only modeling the top half of the rotor and using a symmetry
+boundary condition at the mid-plane. However, 3-D blade-resolved RANS will
+likely not be practical for turbine array simulations for quite some time, and
+until they are, low-order models will need to be developed and improved for this
+purpose.
 

### 2016-03-19 paper.tex
Fix fig names so document can compile

@@
 \draft % marks overfull lines with a black rule on the right
+\usepackage{graphicx}
+\graphicspath{{figures/}}
+
+% Footnotes in tables
+\usepackage{tablefootnote}
 
@@
 
-    \includegraphics[clip, trim=0 1in 0 1in, width=0.8\textwidth]{BR-CFD_CAD}
+    \includegraphics[clip, trim=0 1in 0 1in, width=0.8\textwidth]{CAD}
 
@@
 
-    \includegraphics[width=0.8\textwidth]{BR-CFD_2D_mesh}
+    \includegraphics[width=0.8\textwidth]{2D_mesh}
 
@@
 
-    \includegraphics[width=0.7\textwidth]{BR-CFD_2D_blade_mesh_closeup}
+    \includegraphics[width=0.7\textwidth]{2D_blade_mesh_closeup}
 
@@
 
-    \includegraphics[width=0.9\textwidth]{BR-CFD_verification}
+    \includegraphics[width=0.9\textwidth]{verification}
 
@@
 
-    \includegraphics[width=0.95\textwidth]{BR-CFD_perf_bar_chart}
+    \includegraphics[width=0.95\textwidth]{perf_bar_chart}
 
@@
 
-    \includegraphics[width=0.75\textwidth]{BR-CFD_2D_vorticity_SA_964}
+    \includegraphics[width=0.75\textwidth]{2D_vorticity_SA_964}
 
@@
 
-    \includegraphics[width=0.75\textwidth]{BR-CFD_3D_vorticity_SA_964_10-threshold}
+    \includegraphics[width=0.75\textwidth]{3D_vorticity_SA_964_10-threshold}
 
@@
 
-    \includegraphics[width=0.95\textwidth]{BR-CFD_profiles}
+    \includegraphics[width=0.95\textwidth]{profiles}
 
@@
         \includegraphics[clip, trim=0 0.1in 0.3in 0.2in,
-        width=0.72\textwidth]{RVAT-Re-dep_meancontquiv_10}
+        width=0.72\textwidth]{meancontquiv_exp}
 
@@
         \includegraphics[clip, trim=0 0.2in 0 0.15in,
-        width=0.8\textwidth]{BR-CFD_meancontquiv_kOmegaSST}
+        width=0.8\textwidth]{meancontquiv_kOmegaSST}
 
@@
         \includegraphics[clip, trim=0 0.2in 0 0.15in,
-        width=0.8\textwidth]{BR-CFD_meancontquiv_SpalartAllmaras}
+        width=0.8\textwidth]{meancontquiv_SpalartAllmaras}
 
@@
 
-        \includegraphics[width=0.95\textwidth]{RVAT-Re-dep_k_contours_10}
+        \includegraphics[width=0.95\textwidth]{kcont_exp}
 
@@
 
-        \includegraphics[width=0.95\textwidth]{BR-CFD_kcont_kOmegaSST}
+        \includegraphics[width=0.95\textwidth]{kcont_kOmegaSST}
 
@@
 
-        \includegraphics[width=0.95\textwidth]{BR-CFD_kcont_SpalartAllmaras}
+        \includegraphics[width=0.95\textwidth]{kcont_SpalartAllmaras}
 
@@
 
-    \includegraphics[width=0.9\textwidth]{BR-CFD_mom_bar_graph}
+    \includegraphics[width=0.9\textwidth]{mom_bar_graph}
 

### 2016-03-19 paper.tex
Remove 'would'

@@
 \begin{acknowledgments}
-    The authors would acknowledge funding through a National Science Foundation
-    CAREER award (principal investigator Martin Wosnik, NSF 1150797, Energy for
+    The authors acknowledge funding through a National Science Foundation CAREER
+    award (principal investigator Martin Wosnik, NSF 1150797, Energy for
     Sustainability, program manager Gregory L. Rorrer). The authors also thank

### 2016-03-19 paper.tex
Indent equations
Indent equation and make past tense
Fix ref

@@
     \caption{Mean rotor power (left) and drag (right) coefficient curves from
-        the tow tank experiments\cite{Bachant2015-RVAT-Re-dep-data}.}
+        the tow tank experiments\cite{Bachant2016-RVAT-Re-dep}.}
 
@@
 Turbulence kinetic energy profiles are also shown in
-Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy is calculated as
+Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy was calculated as
 \begin{equation}

### 2016-03-19 paper.tex
Fix ref

@@
 determined from experimental measurements \cite{Bachant2014,
-    Bachant2015-RVAT-Re-dep}.
+    Bachant2016-RVAT-Re-dep}.
 

### 2016-03-19 paper.tex
Move exp C_P curves into numerical setup section

@@
 
-\section{Turbine description}
+\section{Numerical setup}
 
-In this study we seek to predict the near-wake of the UNH-RVAT turbine, shown in
-Figure~\ref{fig:RVAT-CAD}. Experimental performance curves for the UNH-RVAT are
-shown in Figure~\ref{fig:exp_perf}. We modeled the turbine at a tip speed ratio
-$\lambda=1.9$, which corresponds to the maximum power coefficient, and a turbine
-diameter Reynolds number $Re_D \approx 10^6$, which corresponds to a
-$Re$-independent state for both performance and near-wake characteristics, as
-determined from experimental measurements \cite{Bachant2014,
-    Bachant2016-RVAT-Re-dep}.
+In this study the UNH-RVAT baseline tow tank experiment was simulated using a
+mean rotor tip speed ratio $\lambda=1.9$, which corresponds to the maximum
+measured power coefficient; cf. Figure~\ref{fig:exp-perf}. The tow speed or free
+stream velocity $U_\infty=1.0$ m/s gives a turbine diameter Reynolds number
+$Re_D \approx 10^6$, which corresponds to the $Re$-independent state for both
+performance and near-wake characteristics, as determined from previous
+experimental measurements \cite{Bachant2014, Bachant2016-Energies}.
 
-\begin{figure}[ht]
-    \centering
+The turbines CAD geometry was prepared or ``cleaned'' for CFD by removing
+details determined to be unnecessary, e.g., screw heads and axial shaft grooves,
+which would complicate the meshing process and contribute very little to the
+overall loading or flow modulation. A drawing of the physically and numerically
+modeled geometries is presented in Figure~\ref{fig:cfd-cad}.
 
-    \includegraphics[clip, trim=0 1in 0 1in, width=0.8\textwidth]{figures/CAD}
-
-    \caption{CAD drawings of the UNH-RVAT cross-flow turbine as designed (left)
-        and cleaned for simulation (right).}
-
-    \label{fig:RVAT-CAD}
-\end{figure}
-
-\begin{figure}[ht]
+\begin{figure}
     \centering
@@
 
-    \label{fig:exp_perf}
+    \label{fig:exp-perf}
 \end{figure}
-
-\section{Numerical setup}
-
-In this study the UNH-RVAT baseline tow tank experiment was simulated using a
-mean rotor tip speed ratio $\lambda=1.9$, which corresponds to the maximum
-measured power coefficient. The tow speed or free stream velocity $U_\infty=1.0$
-m/s gives a turbine diameter Reynolds number $Re_D \approx 10^6$, which
-corresponds to the $Re$-independent state for both performance and near-wake
-characteristics, as determined from previous experimental measurements
-\cite{Bachant2014, Bachant2016-Energies}.
-
-The turbines CAD geometry was prepared or ``cleaned'' for CFD by removing
-details determined to be unnecessary, e.g., screw heads and axial shaft grooves,
-which would complicate the meshing process and contribute very little to the
-overall loading or flow modulation. A drawing of the physically and numerically
-modeled geometries is presented in Figure~\ref{fig:cfd-cad}.
 

### 2016-03-19 paper.tex
Remove table footnotes

@@
 \graphicspath{{figures/}}
-
-% Footnotes in tables
-\usepackage{tablefootnote}
 
@@
         Alaimo et al.~\cite{Alaimo2015} & $k$--$\epsilon$ & N/A & N/A \\ 
-        Marsh et al.~\cite{Marsh2015} & SST & \cite{Rawlings2008} & N/A \\ 
-        Orlandi et al.~\cite{Orlandi2015} & SST & \cite{Akins1989,Mertens2003} & N/A \\ 
-        Lam \& Peng~\cite{Lam2016} & SST RANS \& IDDES\tablefootnote{Improved delayed detached-eddy simulation.} & N/A & \cite{Tescione2014} \\ 
-        Nini et al.~\cite{Nini2014} & Spalart--Allmaras & N/A & \cite{Battisti2011} \\ 
-        Boudreau \& Dumas~\cite{Boudreau2015} & DDES\tablefootnote{Delayed detached-eddy simulation.} & N/A & N/A \\ 
-        Li et al.~\cite{Li2013} & SST \& Smagorinsky--Lilly LES & \cite{McLaren2011} & N/A \\ 
-        Howell et al.~\cite{Howell2010} & $k$--$\epsilon$ RNG\tablefootnote{Renormalization group.} & \cite{Howell2010} & N/A
+        Marsh et al.~\cite{Marsh2015} & SST & Reference~\cite{Rawlings2008} & N/A \\ 
+        Orlandi et al.~\cite{Orlandi2015} & SST & References~\cite{Akins1989,Mertens2003} & N/A \\ 
+        Lam \& Peng~\cite{Lam2016} & SST RANS \& IDDES & N/A & Reference~\cite{Tescione2014} \\ 
+        Nini et al.~\cite{Nini2014} & Spalart--Allmaras & N/A & Reference~\cite{Battisti2011} \\ 
+        Boudreau \& Dumas~\cite{Boudreau2015} & DDES & N/A & N/A \\ 
+        Li et al.~\cite{Li2013} & SST \& Smagorinsky--Lilly LES & Reference~\cite{McLaren2011} & N/A \\ 
+        Howell et al.~\cite{Howell2010} & $k$--$\epsilon$ RNG & Reference~\cite{Howell2010} & N/A
     \end{tabular}

### 2016-04-09 scripts/zip.py
Add zip script

@@ -1,0 +1,20 @@
+#!/usr/bin/env python
+"""This script zips up all the necessary source files to compile the paper."""
+
+from zipfile import ZipFile
+from subprocess import check_output
+import os
+
+# Describe version via git
+version = check_output(["git", "describe"]).strip().decode()
+
+figures = ["figures/" + f for f in os.listdir("figures") \
+           if f.endswith(".pdf") or f.endswith(".png")]
+
+files = ["paper.tex", "library.bib"]
+
+with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
+    for fig in figures:
+        f.write(fig)
+    for file in files:
+        f.write(file)

### 2016-04-09 scripts/zip.py
Zip bbl file

@@ -11,7 +11,7 @@ version = check_output(["git", "describe"]).strip().decode()
 figures = ["figures/" + f for f in os.listdir("figures") \
            if f.endswith(".pdf") or f.endswith(".png")]
 
-files = ["paper.tex", "library.bib"]
+files = ["paper.tex", "library.bib", "paper.bbl"]
 
 with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
     for fig in figures:

### 2016-04-09 scripts/zip.py
Do not zip up library.bib

@@ -11,7 +11,7 @@ version = check_output(["git", "describe"]).strip().decode()
 figures = ["figures/" + f for f in os.listdir("figures") \
            if f.endswith(".pdf") or f.endswith(".png")]
 
-files = ["paper.tex", "library.bib", "paper.bbl"]
+files = ["paper.tex", "paper.bbl"]
 
 with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
     for fig in figures:

### 2016-04-09 paper.tex
Disable todo list and add preprint note
Add journal submission date
Fix typo

@@
 % Multiple \preprint commands are allowed.
-\preprint{}
+\preprint{Preprint submitted to Journal of Renewable and Sustainable Energy
+19 March 2016}
 
@@
     transport, which plays an important role in the near-wake's momentum
-    balance. The 3-D simulations faired better at predicting performance, with
+    balance. The 3-D simulations fared better at predicting performance, with
     the Spalart--Allmaras model predictions being closest to the experiments.
@@
 \maketitle %\maketitle must follow title, authors, abstract and \pacs
-\listoftodos
+% \listoftodos
 

### 2016-05-06 paper.tex
Text wrapping
Add edits from thesis chapter; resolves #38 resolves #37 resolves #34 resolves #36

@@
 momentum or vortex methods, it remains difficult to predict the performance of
-cross-flow turbines (CFTs, often vertical-axis turbines or VATs) in all
-cases---namely when solidity, or blade chord-to-radius ratio is high---a common
-characteristic of smaller rotors, and those designed for marine hydrokinetic
-(MHK) applications. With computing power becoming evermore available and
-affordable, computational fluid dynamics (CFD) has become an attractive method
-for predicting the performance of cross-flow turbines since it only requires
-that turbulence be modeled, i.e., the physics can be described from theoretical
-true first principles by solving the Navier--Stokes equations using a boundary
-layer resolving body-fitted grid. However, CFD can be computationally expensive
-when done in three dimensions, which may be necessary in some cases.
+cross-flow turbines for all cases---namely when solidity, or blade
+chord-to-radius ratio is high---a common characteristic of smaller rotors, and
+those designed for marine hydrokinetic (MHK) applications. With computing power
+becoming evermore available and affordable, CFD based on the Reynolds-averaged
+Navier--Stokes (RANS) equations has become an attractive method for predicting
+the performance of cross-flow turbines. Since a direct numerical simulation
+(DNS) of all scales at realistic Reynolds numbers is not currently feasible,
+turbulence be modeled. However, with an appropriate turbulence model,
+blade-resolved RANS presents a more physically realistic first principles based
+approach versus simpler models based on momentum theory or potential flow.
+However, CFD can be computationally expensive when done in three dimensions,
+which may be necessary in some cases.
 
@@
 separation~\cite{Menter2003}. Higher fidelity methods that resolve the large
-scales of turbulence, such as large-eddy and detached-eddy simulation have also
-been used. Note that for all the studies listed, model validation was either
-done for performance or wake predictions---not both---and in some case omitted
+scales of turbulence, such as large eddy and detached eddy simulation have also
+been used. Note that for all the studies listed, model validation was performed
+for either performance or wake predictions---not both---and in some case omitted
 entirely.
@@
 through a VAT array using ``leaky Rankine body'' potential flow singularities,
-which was able to rank relative---though not absolut---performance of array
+which was able to rank relative---though not absolute---performance of array
 configurations. Goude and Agren~\cite{Goude2010} used a 2-D vortex method to
 simulate a farm of cross-flow turbines, though this was not validated with
-experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a
-group of cross-flow turbines, observing higher power output for a staggered
-configuration, but also did not compare with experimental results. Giorgetti
-et al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
+experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a group of
+cross-flow turbines, observing higher power output for a staggered
+configuration, but also did not compare with experimental results. Giorgetti et
+al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
 turbine pairs inspired by Dabiri~\cite{Dabiri2011}, but again experimental
@@
 line method to show mutually improved power output from two adjacent turbines,
-though the simulations over-predicted the effects compared with experiments.
-Antheaume et al.~\cite{Antheaume2008} used a blade element approach
-coupled with a 3-D RANS solver to also show how close spacing can improve power
-output of CFTs.
+though the simulations over-predicted the effects by approximately 5\% compared
+with experiments. Antheaume et al.~\cite{Antheaume2008} used a blade element
+approach coupled with a 3-D RANS solver to also show how close spacing can
+improve power output of CFTs.
 
 In this study we set out to model the performance and near-wake of the high
-solidity University of New Hampshire Reference Vertical Axis Turbine (UNH-RVAT)
-using the Reynolds-averaged Navier--Stokes equations discretized via the finite
-volume method. Though studies in the literature generally focus on predicting
-the turbine loading and the local blade boundary layer, we seek to model both
-this and the larger scale flow produced by the rotor, i.e., the near-wake, which
-is of interest for this particular turbine since it has been shown
-experimentally that the near-wake's momentum and energy transport processes are
-dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
-that a 2-D simulation, which omits the vertical dimension, would not correctly
-predict wake recovery and turbine--turbine interaction. However, it is of
-interest to determine how wrong a 2-D model may be, since the computational
-feasibility is attractive.
+solidity University of New Hampshire Reference Vertical-Axis Turbine (UNH-RVAT)
+using the Reynolds-averaged Navier--Stokes equations using the open-source
+finite volume CFD package OpenFOAM, version 2.3.x. Though studies in the
+literature generally focus on predicting the turbine loading and the local blade
+boundary layer, we seek to model both this and the larger scale flow produced by
+the rotor, i.e., the near-wake, which is of interest for this particular turbine
+since it has been shown experimentally that the near-wake's momentum and energy
+transport processes are dominated by vertical advection \cite{Bachant2015-JoT}.
+It logically follows that a 2-D simulation, which omits the vertical dimension,
+would not correctly predict wake recovery and turbine--turbine interaction.
+However, it is of interest to determine how wrong a 2-D model may be, since the
+lower computational cost of 2-D simulations is attractive.
 
-It is assumed that if the turbine power output and near-wake predictions from
-the numerical model match those of the experiments, the flow field can be
-inspected in greater detail, i.e., that the experiments will have been
-``interpolated.'' This will provide access to unmeasured quantities, e.g.,
-pressure, and allow inspection into where the dominant flow structures
-originate. This will ultimately help develop and evaluate low-order wake
-generator models for use in turbine array modeling. In summary, the questions we
-hope to answer here are:
+We seek to validate 2-D and 3-D blade-resolved RANS models against the UNH-RVAT
+mechanical power and near-wake measurements. If the blade loading and velocity
+in the near-wake match well enough, the flow field can potentially be inspected
+in greater detail, i.e., we may be able to make observations of quantities not
+measured experimentally, e.g., pressure, and gain greater insight into where the
+dominant flow structures originate. This will ultimately help develop and
+evaluate low-order wake generator models for use in turbine array modeling. In
+summary, the questions we hope to answer here are:
 
@@
 
-    \item Can 3-D RANS ``interpolate'' the experimental results and provide
-    insight to develop new low-order wake generators to represent CFTs?
+    \item Do the flow fields predicted by 3-D RANS match the experimental wake
+    measurements well enough to provide insight for developing new low-order
+    wake generators to represent CFTs?
 
     \item Does 3-D RANS realize the correct proportions of wake recovery
-    mechanisms, i.e., are the 3-D blade-resolved results a good ``target'' for
-    those a low-order model should produce?
+    mechanisms, i.e., are the 3-D blade-resolved results a good reference case
+    or ``target'' for those a low-order model should produce?
 \end{enumerate}
@@
 
-To close the RANS equations, two different turbulence models were
-used---Menter's $k$--$\omega$ SST \cite{Menter1994} and the Spalart--Allmaras
-(SA) one equation model \cite{Spalart1992}. Both closures use the eddy-viscosity
-approach---the SA employing a single additional scalar transport equation and
-the SST two. The SST model was chosen due to its prominence in the literature
-for simulating separating flows, which we assumed to be present in the current
-problem in the form of dynamic stall. The SA model was shown by Ferreira
-et al.~\cite{Ferreira2007} to match experimental particle image
-velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
-low Reynolds number case ($5 \times 10^4$). Further justification for using the
-SA model for this case comes from Crivellini and D'Alessandro
-\cite{Crivellini2014}, where they successfully modeled the laminar separation
-bubble and subsequent boundary layer transition to turbulence at Reynolds
-numbers similar to those investigated here.
+To close the RANS equations, two different turbulence models were used: Menter's
+$k$--$\omega$ SST \cite{Menter1994} and the Spalart--Allmaras (SA) one equation
+model \cite{Spalart1992}. Both closures use the eddy-viscosity approach---the SA
+employing a single additional scalar transport equation for an
+eddy-viscosity-like quantity $\tilde{\nu}$ and the SST solving equations for the
+turbulence kinetic energy $k$ and specific dissipation $\omega$. The SST model
+was chosen due to its prominence in the literature for simulating separating
+flows, which we assumed to be present in the current problem in the form of
+dynamic stall. The SA model was shown by Ferreira et al.~\cite{Ferreira2007} to
+match experimental particle image velocimetry (PIV) results for a CFT in dynamic
+stall, though this was a somewhat low Reynolds number case ($Re_c \approx 6
+\times 10^4$). Further justification for using the SA model for this case comes
+from Crivellini and D'Alessandro \cite{Crivellini2014}, where they successfully
+modeled the laminar separation bubble and subsequent boundary layer transition
+to turbulence at Reynolds numbers similar to those investigated here.
 
@@
 
-Meshes were generated using \textit{OpenFOAM}'s \textit{blockMesh} and
-\textit{snappyHexMesh} utilities. Mesh topology consists of a background
-hexahedral mesh, which is refined in all three directions by a factor of 2 in a
-rectangular region containing the turbine and near-wake (0.9 m upstream, 1.3 m
-downstream, $\pm 0.9$ m cross-stream, and $\pm 0.8$ m vertically). Cells
-adjacent to the turbine shaft and struts are refined by a factor of 4, while
-cells adjacent to the blades are refined by a factor of 6. To capture the
-boundary layer, 20 layers are added next to the blades with an expansion ratio
-of 1.2. Overall mesh refinement is controlled by a single parameter---the number
-of cells in the streamwise direction, $N_x$.
+Meshes were generated using OpenFOAM's blockMesh and snappyHexMesh utilities.
+Mesh topology consists of a background hexahedral mesh, which is refined in all
+three directions by a factor of 2 in a rectangular region containing the turbine
+and near-wake (0.9 m upstream, 1.3 m downstream, $\pm 0.9$ m cross-stream, and
+$\pm 0.8$ m vertically). Cells adjacent to the turbine shaft and struts are
+refined by a factor of 4, while cells adjacent to the blades are refined by a
+factor of 6. To capture the boundary layer, 20 layers were added next to the
+blades with an expansion ratio of 1.2. Overall mesh refinement is controlled by
+a single parameter---the number of cells in the streamwise direction, $N_x$.
 
@@
 
-Simulations were run using the \textit{pimpleDyMFoam} solver from the
-open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
-\textit{pimpleDyMFoam} uses a hybrid PISO-SIMPLE algorithm for pressure-velocity
-coupling and is compatible with dynamic meshes. An Euler scheme was used to
-advance the simulation forward in time. The case files required to replicate the
-simulations are available from \cite{Bachant2016-UNH-RVAT-2D-OpenFOAM-SST,
-    Bachant2016-UNH-RVAT-2D-OpenFOAM-SA, Bachant2016-UNH-RVAT-3D-OpenFOAM-SST,
-    Bachant2016-UNH-RVAT-3D-OpenFOAM-SA}.
+Simulations were run using OpenFOAM's pimpleDyMFoam solver, which uses a hybrid
+PISO-SIMPLE algorithm for pressure-velocity coupling and is compatible with
+dynamic meshes. An Euler scheme was used to advance the simulation forward in
+time. The case files required to replicate the simulations are available from
+\cite{Bachant2016-UNH-RVAT-2D-OpenFOAM-SST, Bachant2016-UNH-RVAT-2D-OpenFOAM-SA,
+Bachant2016-UNH-RVAT-3D-OpenFOAM-SST, Bachant2016-UNH-RVAT-3D-OpenFOAM-SA}.
 
@@
 m/s to match the tow tank case, while the top boundary condition was a slip
-velocity condition. Pressure was held fixed at the outlet. Note that in 2-D the
-top and bottom boundary conditions are ``empty,'' which is an OpenFOAM
-convention to indicate two-dimensionality.
+velocity condition. Pressure was held fixed at the outlet while a zero-gradient
+boundary condition was applied at the inlet (a typical
+velocity-inlet/pressure-outlet case). Note that in 2-D the top and bottom
+boundary conditions are ``empty,'' which is an OpenFOAM convention to indicate
+two-dimensionality.
 
@@
 shown in Figure~\ref{fig:2d-br-verification}, from which the final number of
-streamwise grid points $N_x = 70$ was chosen. This corresponds to a total cell
-count of approximately $5 \times 10^4$ for the 2-D cases and 16 million for the
-3-D case.
+streamwise grid points $N_x = 70$ was chosen, which resulted in a nondimensional
+wall distance $y^+ = u^* y / \nu \sim 1$, where $u^*$ is the friction velocity,
+defined as $u^*=\sqrt{\tau_w / \rho}$, where $\tau_w$ is the wall shear stress.
+This mesh resolution gave a total cell count of approximately $5 \times 10^4$
+for the 2-D cases and 16 million for the 3-D case. Sensitivity to domain length
+was assessed as well, showing a 2\% increase in mean power coefficient for the
+2-D SST case with the domain extended an additional two rotor diameters
+downstream.
 
@@
 SST model's convergence behavior may be due to its specific implementation in
-\textit{OpenFOAM}, and not indicative of the nature of the model equations.
-Verification studies for CFTs with this level of detail in the literature are
-not common, though the final time step is comparable to others
-\cite{Balduzzi2016}.
+OpenFOAM, and not indicative of the nature of the model equations. Verification
+studies for CFTs with this level of detail in the literature are not common,
+though the final time step is comparable to others \cite{Balduzzi2016}.
 
@@
 six rotor revolutions. Computations for the 3-D cases were run on 192 processes
-(24 nodes $\times$ 8 cores each) and took on the order of 1,000 CPU hours per
+(24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
+performance computing cluster) and took on the order of 1,000 CPU hours per
 second of simulated time. The 2-D simulations were run on a single processor and
@@
 
-Flow statistics were calculated over the time interval spanning $5$--$10$
-seconds, or approximately three rotor revolutions, from time series downsampled
-to 50 Hz. It is assumed that the downsampling frequency is high enough above the
-blade passage frequency such that differences from the variance (for computing
+Performance statistics were computed after the first revolution onward and flow
+statistics were calculated over the time interval spanning $5$--$10$ seconds, or
+approximately three rotor revolutions, from time series downsampled to 50 Hz. It
+is assumed that the downsampling frequency is high enough above the blade
+passage frequency such that differences from the variance (for computing
 turbulence statistics) in the original velocity will be negligible.
@@
 
-Mean performance quantities---power and overall rotor drag coefficients---were
-calculated for the interval after two full rotor revolutions, beyond which
-values had reached an approximately periodic state. Predictions for both the
-mean rotor power and drag coefficients are shown in
+Predictions for both the mean rotor power and drag coefficients are shown in
 Figure~\ref{fig:br-cfd-perf-bar-chart}. In general, the 2-D CFD cases both
 significantly overpredict turbine loading and therefore mechanical power output,
-which is likely due to their increased blockage ratio, unresolved blade end
-effects, and lack of blade support struts.
+which is due to their increased blockage ratio, unresolved blade end effects,
+and lack of blade support struts.
 
-The 3-D simulations fair better at predicting the experimental measurements. The
+The 3-D simulations matched the experimental measurements more closely. The
 Spalart--Allmaras model's mean power coefficient was 0.24 while the SST's was
 0.33, which represent a 6\% under- and 30\% overprediction, respectively,
-compared with the experiments. The apparent overprediction of rotor drag
-coefficient could be an effect of the experimental procedure, where the ``tare
-drag'' from the turbine mounting structure was measured without a turbine
-installed, then subtracted in post-processing. Technically the local flow field
-will have changed, and hence the tare drag will change as well.
+compared with the experiments. Some of the apparent overprediction of rotor drag
+coefficient could also be an effect of the experimental procedure, where the
+``tare drag'' from the turbine mounting structure was measured without a turbine
+installed, then subtracted in post-processing. With the turbine installed, flow
+past the mounting frame will be higher due to blockage, meaning the tare drag
+would be underestimated, and the CFD results for $C_D$ should be slightly higher
+than the experimental measurements.
 
@@
 presented for the 2-D and 3-D Spalart--Allmaras cases in
-Figure~\ref{fig:br-vorticity-3d} and Figure~\ref{fig:br-vorticity-3d},
+Figure~\ref{fig:br-vorticity-2d} and Figure~\ref{fig:br-vorticity-3d},
 respectively. It can be seen how the upstream blade---as it turns back into the
@@
 Turbulence kinetic energy profiles are also shown in
-Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy was calculated as
+Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy for the unsteady
+RANS models was calculated as
 \begin{equation}
@@
 \end{equation}
-where $U^\prime = U - \overline{U}$ and $k_{\mathrm{RA}}$ is the kinetic energy
-calculated by the turbulence model, which is zero for the SA model.
+where $U^\prime = U - \overline{U}$ (the resolved velocity fluctuations) and
+$k_{\mathrm{RA}}$ is the kinetic energy calculated by the turbulence model,
+which is zero for the SA model.
 
-Both Spalart--Allmaras cases do a poor job predicting the turbulence kinetic
+Both Spalart--Allmaras cases did a poor job predicting the turbulence kinetic
 energy in the flow, since it must be resolved as variance in the velocity field.
-The 2-D SST model does a good job predicting the peak in $k$ at $y/R=-1$, though
+The 2-D SST model did a good job predicting the peak in $k$ at $y/R=-1$, though
 is missing the smaller peak at $y/R=-1$. This is once again likely due to
@@
 In order to visualize the mean velocity field, vector arrows for the mean
-cross-stream and vertical components are superimosed on top of contours of the
+cross-stream and vertical components are superimposed on top of contours of the
 streamwise component at $X/D=1$ in Figure~\ref{fig:br-cfd-mean-velocity}. Both
 CFD models predict the general structure of the mean velocity well, though the
-SA case has slightly larger vertical mean flows, which could be due to stronger
-tip vortex generation, or lower diffusivity compared with the SST model.
-Additional discrepancies between CFD and experiments may be due to the top slip
-boundary condition versus the experiment's free surface.
+SA case has a slightly larger vertical mean flow component, which could be due
+to stronger tip vortex generation, or lower diffusivity compared with the SST
+model.
 
@@
 CFD case at $x/D=1$ are presented in Figure~\ref{fig:br-cfd-kcont}. As seen in
-the profiles in Figure~\ref{fig:br-cfd-profiles}, the SA model is resolving very
-little of the flow unsteadiness. In contrast, the SST model does a good job
-predicting the locations and magnitudes of various peaks in $k$. These are
-generated along the top of the turbine via tip vortex shedding, and the $-y$
-side of the turbine via dynamic stall. We do however see the smearing effect
-from the dynamic stall vortex centered around $z/H=0$, which is likely more of
-an issue with wake evolution rather than wake generation.
+the profiles in Figure~\ref{fig:br-cfd-profiles}, the SA model was not able to
+resolve the majority of the unsteadiness in the flow. In contrast, the SST model
+did a good job predicting the locations and magnitudes of various peaks in $k$.
+These are generated along the top of the turbine via tip vortex shedding, and
+the $-y$ side of the turbine via dynamic stall. We do however see the smearing
+effect from the dynamic stall vortex centered around $z/H=0$, which is likely
+more of an issue with wake evolution rather than wake generation.
 
@@
 the flow at $x/D=1$. Unfortunately, pressure data were not acquired from the
-experiment, though the 3-D delayed detached eddy simulation (DDES) of Boudreau
-and Dumas \cite{Boudreau2015} concur with the adverse pressure gradient
-condition.
+experiment, though results from the 3-D delayed detached eddy simulation (DDES)
+of Boudreau and Dumas \cite{Boudreau2015} concur with the adverse pressure
+gradient condition.
 
@@
 A cross-flow turbine was modeled using the $k$--$\omega$ SST and
-Spalart--Allmaras (SA) Reynolds-averaged Navier--Stokes turbulence models to test
-their abilities to predict turbine performance and near-wake dynamics. It was
-observed that when modeled in 2-D, the performance is over-predicted compared to
-the data from the tow tank experiments, which was expected due to omission of
-blade end effects, support strut drag, and increased blockage. Vertical (or
-axial) wake dynamics were unresolved in the 2-D model, despite being identified
-as a significant contributor to streamwise wake recovery, which casts doubt on
-the 2-D model's applicability as a tool to study array spacing effects.
+Spalart--Allmaras (SA) Reynolds-averaged Navier--Stokes turbulence models to
+test their abilities to predict turbine performance and near-wake dynamics. It
+was observed that when modeled in 2-D, the performance is over-predicted
+compared to the data from the tow tank experiments, which was expected due to
+omission of blade end effects, support strut drag, and increased blockage.
+Vertical (or axial) wake dynamics were unresolved in the 2-D model, despite
+being identified as a significant contributor to streamwise wake recovery, which
+casts doubt on the 2-D model's applicability as a tool to study array spacing
+effects.
 
@@
 The effectiveness at predicting the mean velocity field gives credibility to the
-prospect of using the computed flow field to interpolate the experimental
+prospect of using the computed flow field to extrapolate the experimental
 results, such that the CFD results can be used as a target for a low-order wake
@@
 over the entire blade surface, which were not measured experimentally.
-Ultimately, however, the computational cost of 3-D simulations---about 10,000
-CPU hours per operating case---may be too expensive to be used for CFT
-engineering work, especially considering the uncertainty involved compared with
-physical model studies. Since the 3-D wake results appeared to be only weakly
-asymmetrical in the $x$--$y$ plane about $z/H=0$, it may be possible to reduce
-computing power by only modeling the top half of the rotor and using a symmetry
-boundary condition at the mid-plane. However, 3-D blade-resolved RANS will
-likely not be practical for turbine array simulations for quite some time, and
-until they are, low-order models will need to be developed and improved for this
-purpose.
+Ultimately, however, the computational cost of 3-D simulations---about 1,000 CPU
+hours per simulated second (or roughly 10,000 CPU hours per operating
+point)---may be too expensive to be used for CFT engineering work, especially
+considering the uncertainty involved compared with physical model studies. Since
+the 3-D wake results appeared to be only weakly asymmetrical in the $x$--$y$
+plane about $z/H=0$, it may be possible to reduce computing power by only
+modeling the top half of the rotor and using a symmetry boundary condition at
+the mid-plane. However, 3-D blade-resolved RANS will likely not be practical for
+turbine array simulations for quite some time, and until they are, low-order
+models will need to be developed and improved for this purpose.
 

### 2016-05-06 paper.tex
Interpolate --> extrapolate
Use gmail since UNH address changing
Add NS type to table

@@
 \author{Peter Bachant}
-\email[]{pxl3@unh.edu}
+\email[]{pete.bachant@gmail.com}
 %\homepage[]{Your web page}
@@
     terms in the near-wake. These results show the potential of blade-resolved
-    RANS as a design tool and a way to ``interpolate'' experimental flow field
+    RANS as a design tool and a way to ``extrapolate'' experimental flow field
     measurements.
@@
         \hline
-        Alaimo et al.~\cite{Alaimo2015} & $k$--$\epsilon$ & N/A & N/A \\
-        Marsh et al.~\cite{Marsh2015} & SST & Reference~\cite{Rawlings2008} & N/A \\
-        Orlandi et al.~\cite{Orlandi2015} & SST & References~\cite{Akins1989,Mertens2003} & N/A \\
+        Alaimo et al.~\cite{Alaimo2015} & $k$--$\epsilon$ RANS & N/A & N/A \\
+        Marsh et al.~\cite{Marsh2015} & SST RANS & Reference~\cite{Rawlings2008} & N/A \\
+        Orlandi et al.~\cite{Orlandi2015} & SST RANS & References~\cite{Akins1989,Mertens2003} & N/A \\
         Lam \& Peng~\cite{Lam2016} & SST RANS \& IDDES & N/A & Reference~\cite{Tescione2014} \\
-        Nini et al.~\cite{Nini2014} & Spalart--Allmaras & N/A & Reference~\cite{Battisti2011} \\
-        Boudreau \& Dumas~\cite{Boudreau2015} & DDES & N/A & N/A \\
-        Li et al.~\cite{Li2013} & SST \& Smagorinsky--Lilly LES & Reference~\cite{McLaren2011} & N/A \\
-        Howell et al.~\cite{Howell2010} & $k$--$\epsilon$ RNG & Reference~\cite{Howell2010} & N/A
+        Nini et al.~\cite{Nini2014} & Spalart--Allmaras RANS & N/A & Reference~\cite{Battisti2011} \\
+        Boudreau \& Dumas~\cite{Boudreau2015} & Spalart--Allmaras DDES & N/A & N/A \\
+        Li et al.~\cite{Li2013} & SST RANS \& Smagorinsky--Lilly LES & Reference~\cite{McLaren2011} & N/A \\
+        Howell et al.~\cite{Howell2010} & $k$--$\epsilon$ RNG RANS & Reference~\cite{Howell2010} & N/A
     \end{tabular}

### 2016-09-25 scripts/diff.sh
Update build infrastructure

@@ -1,0 +1,12 @@
+#!/usr/bin/env bash
+# This script creates a diff version comparing the recent with whatever revision
+# is supplied as the first argument. This can be any valid Git identifier, e.g.,
+# a commit hash, tag, etc.
+
+CURRENT=$(git describe --always --dirty)
+rev=$1
+git show $rev:paper.tex > archive/paper-$CURRENT-$rev.tex
+latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$CURRENT-$rev.tex paper.tex > paper-$CURRENT-$rev-diff.tex
+latexmk paper-$CURRENT-$rev-diff.tex -pdf -f
+mv paper-$CURRENT-$rev-diff.pdf archive
+rm paper-$CURRENT-$rev*

### 2016-09-25 reviews/reviewer2-response.md
Start review responses

# Response to reviewer 2

We thank the reviewer for his or her thoughtful comments, which have helped us
improve the paper.

>In general more work needs to be performed on
• Discussion of mesh factors that may affect the simulation results

>• Nomenclature, and the lack of definition of many of the variables will cause confusion for many readers

>• Improvement of figures with nomenclature, labels, axis etc.

>• More explanation of why the various transport properties are investigated and their applicability

>Additionally the title could be better, maybe add a Computational Fluid Dynamics or simulation somewhere?

### 2016-09-25 reviews/reviewer3-response.md
Start review responses

# Response to reviewer 3

We appreciate the comments given here, and addressing them has improved the manuscript significantly.

>1. For convergence study, Fig 5 gives the results for different choice of time step. however, for SST model, the Cp do not convergent with the time step. On the other hand, the choice of time step is always coupled with the choice of space step. Generally speaking, the time step cannot be too small for a certain space discretion condition. Therefore, It is suggested that the author provide this relationship between space and time discretion.

>2. No the convergence study for 3d model in this paper.

>3. In Performance Prediction session, the author gives a figure(Fig.6 ) to show the comparison of 3d , 2d and experimental data. But fig.6 only gives the results for one case for different methods. The 2d method seems over-predict the Cp and Cd. more results of different cases should be given to verify this idea.

>4. Shaft and Arm/Strut effects are important to the turbine, and have been discussed by previous publications by using RANS such as
>
Behrouzi F, Malik A M A, Nakisa M, et al. Arm Effect on Performance of Vertical Axis Current Turbine Using RANS Simulation. Applied Mechanics & Materials, 2014, 554(554):531-535.
>
and by using experimental test such as
Li Y, Calisal S M. Three-dimensional effects and arm effects on modeling a vertical axis tidal current turbine. Renewable Energy, 2010, 35(10):2325-2334.
>
However, the authors have not well explained it in the manuscript. Please extend this part.

### 2016-09-25 reviews/reviewer3-response.md
Clarify convergence study strategy

@@
 Therefore, It is suggested that the author provide this relationship between space and time discretion.
+
+It is true that in general spatial and temporal discretization are coupled.
+However, using the PIMPLE solver in OpenFOAM allows for stable solutions even
+when the CFL condition is not satisfied in every cell. In our case, we used a
+fixed time step, such that it would not be changed by the solver based on the
+Courant number. For example, the temporal convergence data shown were acquired
+using the "final" spatial resolution and vice versa. This did take some
+iterative process, but ultimately we have one independent variable for each
+plot. A note has been added to the manuscript to clarify this point.
 

### 2016-09-25 paper.tex
Clarify wording on BC description
Clarify convergence study strategy

@@
 
-Initial and boundary conditions were set to match those of the tow tank as well
-as possible. The velocity at the inlet, bottom, and side walls was fixed to 1
-m/s to match the tow tank case, while the top boundary condition was a slip
+Initial and boundary conditions were set to approximate those of the tow tank as
+closely as possible. The velocity at the inlet, bottom, and side walls was fixed
+to 1 m/s to match the tow tank case, while the top boundary condition was a slip
 velocity condition. Pressure was held fixed at the outlet while a zero-gradient
@@
 studies for CFTs with this level of detail in the literature are not common,
-though the final time step is comparable to others \cite{Balduzzi2016}.
+though the final time step is comparable to others \cite{Balduzzi2016}. Note
+that while in general there may be a coupling between necessary spatial and
+temporal resolution, these values were arrived at iteratively, i.e., the spatial
+dependence was varied with the chosen time step, and vice versa, such that there
+was one independent variable for each parameter sweep.
 

### 2016-09-25 reviews/reviewer3-response.md
Note that a 3-D grid convergence study was not done

@@
 
+This is true, we did not perform a true convergence study for the 3-D model. We
+simply extrapolated the 2-D grid into three dimensions. While this may not be
+ideal, this strategy may be useful to practitioners looking to perform similar
+simulations, but do not have the computational resources to perform multiple
+high fidelity 3-D simulations. The paper has been adjusted to clarify this
+point.
+
 >3.

### 2016-09-25 paper.tex
Note that a 3-D grid convergence study was not done

@@
 
+Sensitivity to spatial resolution and time step was not assessed for the 3-D
+case, as the computational cost would have been prohibitive. However, the mesh
+resolution in the $x$--$y$ plane and time step were chosen based on the 2-D
+results. Selecting a grid and time step in this way presents an alternative that
+may be useful to practitioners who do not have access to the computationl
+resources necessary to perform multiple high fidelity 3-D simulations.
+
 \section{Results}

### 2016-09-25 reviews/reviewer3-response.md
Clarify that we were only looking at optimal TSR

@@
 
+For this study we were mainly looking to model the near-wake at peak performance
+(determined experimentally), rather than the entire performance envelope. The
+performance predictions are therefore an indication of the loading on the
+turbine and the level to which it imparts force onto the flow. Notes have been
+added to the manuscript to clarify this point.
+
 >4. Shaft and Arm/Strut effects are important to the turbine, and have been discussed by previous publications by using RANS such as

### 2016-09-25 paper.tex
Clarify that we were only looking at optimal TSR

@@
 
-We seek to validate 2-D and 3-D blade-resolved RANS models against the UNH-RVAT
-mechanical power and near-wake measurements. If the blade loading and velocity
-in the near-wake match well enough, the flow field can potentially be inspected
-in greater detail, i.e., we may be able to make observations of quantities not
-measured experimentally, e.g., pressure, and gain greater insight into where the
-dominant flow structures originate. This will ultimately help develop and
-evaluate low-order wake generator models for use in turbine array modeling. In
-summary, the questions we hope to answer here are:
+We sought to validate 2-D and 3-D blade-resolved RANS models against the
+UNH-RVAT mechanical power and near-wake measurements taken at the experimentally
+determined optimal tip speed ratio $\lambda_0 = 1.9$. If the blade loading and
+velocity in the near-wake match well enough, the flow field can potentially be
+inspected in greater detail, i.e., we may be able to make observations of
+quantities not measured experimentally, e.g., pressure, and gain greater insight
+into where the dominant flow structures originate. This will ultimately help
+develop and evaluate low-order wake generator models for use in turbine array
+modeling. In summary, the questions we hope to answer here are:
 
@@
 
-Predictions for both the mean rotor power and drag coefficients are shown in
+Predictions for both the mean rotor power and drag coefficients at the optimal
+tip speed ratio $\lambda_0 = 1.9$ are shown in
 Figure~\ref{fig:br-cfd-perf-bar-chart}. In general, the 2-D CFD cases both
@@
 
-    \caption{Power (left) and drag (right) coefficient predictions from
-        experiments and each numerical model.}
+    \caption{Power (left) and drag (right) coefficient predictions at $\lambda =
+    1.9$, from experiments and each numerical model.}
 

### 2016-09-25 paper.tex
Add note about strut drag; resolves #43

@@
 which is due to their increased blockage ratio, unresolved blade end effects,
-and lack of blade support struts.
+and lack of blade support struts. Effects of blade support strut drag were
+investigated experimentally by Rawlings~\cite{Rawlings2008}, showing that for a
+rotor with six foil-shaped struts, parasitic losses would cause a 4 percentage
+point reduction in $C_P$ at $\lambda=2$. Since the UNH-RVAT has only three
+foil-shaped struts, it therefore follows that the overprediction by the 2-D
+simulations is mostly attributable to blockage and end effects, rather than
+support strut drag.
 

### 2016-10-01 paper.tex
Address reviewer 2 comments in abstract

@@
     numerically via blade-resolved $k$--$\omega$ SST and Spalart--Allmaras RANS
-    models in two and three dimensions. Results for each case are compared with
-    experimental measurements of the turbine shaft power, overall drag, mean
-    velocity, turbulence kinetic energy, and momentum transport terms in the
-    near-wake at one diameter downstream. It was shown that 2-D simulations
-    overpredict turbine loading and do not resolve mean vertical momentum
-    transport, which plays an important role in the near-wake's momentum
-    balance. The 3-D simulations fared better at predicting performance, with
-    the Spalart--Allmaras model predictions being closest to the experiments.
-    The SST model more accurately predicted the turbulence kinetic energy while
-    the Spalart--Allmaras model more closely matched the momentum transport
-    terms in the near-wake. These results show the potential of blade-resolved
-    RANS as a design tool and a way to ``extrapolate'' experimental flow field
-    measurements.
+    models in two and three dimensions. Results for each case were compared with
+    experimental measurements of the turbine shaft power, overall streamwise
+    rotor drag, mean velocity, turbulence kinetic energy, and momentum transport
+    terms in the near-wake at one diameter downstream. It was shown that 2-D
+    simulations overpredict turbine loading and do not resolve mean vertical
+    momentum transport, which plays an important role in the near-wake's
+    momentum balance. The 3-D simulations fared better at predicting
+    performance, with the Spalart--Allmaras model predictions being closest to
+    the experiments. The SST model more accurately predicted the turbulence
+    kinetic energy while the Spalart--Allmaras model more closely matched the
+    momentum transport terms in the near-wake. These results show the potential
+    of blade-resolved RANS as a design tool and a way to ``extrapolate''
+    experimental flow field measurements.
 \end{abstract}

### 2016-10-01 paper.tex
Address reviewer 2 comments on page 2

@@
 National Labs in the US \cite{Sutherland2012} and the National Research Council
-in Canada \cite{Para2002}, the vertical-axis cross-flow Darrieus turbines were
-all but abandoned for large scale commercial terrestrial wind power. Today,
-however, with the development of marine hydrokinetic (MHK) energy devices, CFTs
-are at the forefront, with the first grid-connected commercial device being a
+in Canada \cite{Para2002}, vertical-axis cross-flow Darrieus turbines were all
+but abandoned for large scale commercial terrestrial wind power. Today however,
+with the development of marine hydrokinetic (MHK) energy devices, CFTs are at
+the forefront, with the first grid-connected commercial device being a
 horizontal-axis CFT manufactured and installed by ORPC in Cobscook Bay, Maine
@@
 momentum or vortex methods, it remains difficult to predict the performance of
-cross-flow turbines for all cases---namely when solidity, or blade
-chord-to-radius ratio is high---a common characteristic of smaller rotors, and
-those designed for marine hydrokinetic (MHK) applications. With computing power
-becoming evermore available and affordable, CFD based on the Reynolds-averaged
-Navier--Stokes (RANS) equations has become an attractive method for predicting
-the performance of cross-flow turbines. Since a direct numerical simulation
-(DNS) of all scales at realistic Reynolds numbers is not currently feasible,
-turbulence be modeled. However, with an appropriate turbulence model,
-blade-resolved RANS presents a more physically realistic first principles based
-approach versus simpler models based on momentum theory or potential flow.
-However, CFD can be computationally expensive when done in three dimensions,
-which may be necessary in some cases.
+cross-flow turbines when solidity, or blade chord-to-radius ratio is high---a
+common characteristic of smaller rotors, and those designed for marine
+hydrokinetic (MHK) applications, since water's higher density increases rotor
+loading by roughly an order of magnitude for the same frontal
+area~\cite{Bachant2016-thesis}. With computing power becoming evermore available
+and affordable, CFD based on the Reynolds-averaged Navier--Stokes (RANS)
+equations has become an attractive method for predicting the performance of
+cross-flow turbines. Since a direct numerical simulation (DNS) of all scales at
+realistic Reynolds numbers is not currently feasible, turbulence be modeled.
+However, with an appropriate turbulence model, blade-resolved RANS presents a
+more physically realistic first principles based approach versus simpler models
+based on momentum theory or potential flow. However, CFD can be computationally
+expensive when done in three dimensions, but may be necessary to predict
+performance for low aspect ratio or variable radius rotors, and to predict the
+unique flow field produced by CFTs, and therefore wake recovery and interaction,
+which is discussed in detail later.
 

### 2016-10-01 paper.tex
Rev 2 comments up through p3

@@
 There are many examples in the literature of 2-D cross-flow turbine simulations
-with widely varying results. Balduzzi et al.~\cite{Balduzzi2016} provides
-a summary of recent efforts and an attempt to standardize a methodology for
-using Reynolds-averaged Navier--Stokes (RANS) to correctly predict performance
-of a 2-D CFT. Howell et al.~\cite{Howell2010}, performed both 2-D and 3-D
-simulations of a high solidity cross-flow turbine using a $k$--$\epsilon$
-renormalization group (RNG) turbulence model. The results from the 2-D
-simulations over-predicted power coefficient, while the 3-D case matched well
-with wind tunnel measurements near the tip speed ratio of maximum power. In
-general, 3-D simulations are less common, but have begun to appear more
-frequently recently---a testament to the progress towards higher computing
-power.
+with widely varying results. Balduzzi et al.~\cite{Balduzzi2016} provides a
+summary of recent efforts and an attempt to standardize a methodology for using
+Reynolds-averaged Navier--Stokes (RANS) to correctly predict performance of a
+2-D CFT. Howell et al.~\cite{Howell2010}, performed both 2-D and 3-D simulations
+of a high solidity cross-flow turbine using a $k$--$\epsilon$ renormalization
+group (RNG) turbulence model. The results from the 2-D simulations
+over-predicted power coefficient since the rotor had a low aspect ratio, while
+the 3-D case matched well with wind tunnel measurements near the tip speed ratio
+of maximum power. In general, 3-D simulations are less common, but have begun to
+appear more frequently recently---a testament to the progress towards higher
+computing power.
 
@@
 scales of turbulence, such as large eddy and detached eddy simulation have also
-been used. Note that for all the studies listed, model validation was performed
-for either performance or wake predictions---not both---and in some case omitted
-entirely.
+been used, though the high computational expense of these techniques may
+preclude their use in industrial applications. Note that for all the studies
+listed, model validation was performed for either performance or wake
+predictions---not both---and in some case omitted entirely. In general, the 3-D
+RANS results cannot replicate a turbine's entire performance envelope, i.e.,
+$C_P$--$\lambda$ curve, but are successful for some tip speed ratios.
 
@@
 Modeling the boundary layer flows on cross-flow turbine blades is essential to
-predicting the blade loading. This flow condition presents a challenge due to
-the dynamically changing inflow velocity and angle of attack---which often
-exceeds static stall values and causes dynamic stall. Furthermore, the ability
-to predict the occurrence and interdependence of boundary layer transition to
-turbulence and separation can have dramatic influence on the blade loading and
-therefore the predicted turbine power output. These challenges present
-significant obstacles to the prospect of using CFD to replace wind tunnel or
-tank testing of physical models.
+predicting the blade loading. This flow condition presents a challenge since the
+blades encounter a constantly changing relative velocity and angle of
+attack---which often exceeds static stall values and causes dynamic stall.
+Furthermore, the ability to predict the occurrence and interdependence of
+boundary layer transition to turbulence and separation can have dramatic
+influence on the blade loading and therefore the predicted turbine power output.
+These challenges present significant obstacles to the prospect of using CFD to
+replace wind tunnel or tank testing of physical models.
 

### 2016-10-01 paper.tex
Remove irrelevant paragraph per rev 2

@@
 replace wind tunnel or tank testing of physical models.
-
-To date, little computational work has been done to attempt to design arrays of
-CFTs, despite their prospects for closer spacing compared with axial-flow
-turbines (AFTs). For example, Araya et al.~\cite{Araya2014} modeled the flow
-through a VAT array using ``leaky Rankine body'' potential flow singularities,
-which was able to rank relative---though not absolute---performance of array
-configurations. Goude and Agren~\cite{Goude2010} used a 2-D vortex method to
-simulate a farm of cross-flow turbines, though this was not validated with
-experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a group of
-cross-flow turbines, observing higher power output for a staggered
-configuration, but also did not compare with experimental results. Giorgetti et
-al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
-turbine pairs inspired by Dabiri~\cite{Dabiri2011}, but again experimental
-validation was not performed. Li and Calisal \cite{Li2010} used a 3-D vortex
-line method to show mutually improved power output from two adjacent turbines,
-though the simulations over-predicted the effects by approximately 5\% compared
-with experiments. Antheaume et al.~\cite{Antheaume2008} used a blade element
-approach coupled with a 3-D RANS solver to also show how close spacing can
-improve power output of CFTs.
 

### 2016-10-01 paper.tex
Add apostrophe

@@
 
-The turbines CAD geometry was prepared or ``cleaned'' for CFD by removing
+The turbine's CAD geometry was prepared or ``cleaned'' for CFD by removing
 details determined to be unnecessary, e.g., screw heads and axial shaft grooves,

### 2016-10-02 paper.tex
Add ref for rev 2

@@
 flows, which we assumed to be present in the current problem in the form of
-dynamic stall. The SA model was shown by Ferreira et al.~\cite{Ferreira2007} to
-match experimental particle image velocimetry (PIV) results for a CFT in dynamic
-stall, though this was a somewhat low Reynolds number case ($Re_c \approx 6
-\times 10^4$). Further justification for using the SA model for this case comes
-from Crivellini and D'Alessandro \cite{Crivellini2014}, where they successfully
-modeled the laminar separation bubble and subsequent boundary layer transition
-to turbulence at Reynolds numbers similar to those investigated here.
+dynamic stall~\cite{Para2002}. The SA model was shown by Ferreira et
+al.~\cite{Ferreira2007} to match experimental particle image velocimetry (PIV)
+results for a CFT in dynamic stall, though this was a somewhat low Reynolds
+number case ($Re_c \approx 6 \times 10^4$). Further justification for using the
+SA model for this case comes from Crivellini and D'Alessandro
+\cite{Crivellini2014}, where they successfully modeled the laminar separation
+bubble and subsequent boundary layer transition to turbulence at Reynolds
+numbers similar to those investigated here.
 

### 2016-10-02 paper.tex
Clarify computational domain size for rev 2

@@
 
-The computational domain was a rectangular volume 3.66 m long, 3.66 m wide, and
-2.44 m tall (for 3-D), with the turbine located 1.52 m from the inlet, and
-centered vertically with a vertical axis, designed to match the tow tank
-dimensions for comparison with previous experiments. The rotor geometry was
-located at the center of a cylindrical sliding mesh interface, which rotated at
-a mean tip speed ratio $\lambda=1.9$ with a sinusoidal oscillation at the blade
-passage frequency---with and amplitude of 0.19 and the first peak at 1.4
-radians---to mimic the slight deviation from the mean tip speed ratio observed
-in the experiments. The 2-D mesh overview is shown in
-Figure~\ref{fig:2d-br-mesh} and the blade mesh detail is shown in
-Figure~\ref{fig:blade-mesh}.
+The computational domain was designed to match the tow tank dimensions for
+comparison with previous experiments. Thus, it was modeled as a rectangular
+volume 3.66 m long, 3.66 m wide, and 2.44 m tall (for the 3-D cases), with the
+turbine located 1.52 m from the inlet, centered within the tank depth with its
+axis oriented vertically. The rotor geometry was located at the center of a
+cylindrical sliding mesh interface, which rotated at a mean tip speed ratio
+$\lambda=1.9$ with a sinusoidal oscillation at the blade passage
+frequency---with and amplitude of 0.19 and the first peak at 1.4 radians---to
+mimic the slight deviation from the mean tip speed ratio observed in the
+experiments. The 2-D mesh overview is shown in Figure~\ref{fig:2d-br-mesh} and
+the blade mesh detail is shown in Figure~\ref{fig:blade-mesh}.
 

### 2016-10-02 paper.tex
Explain justification for using wall functions on struts, etc.
Address possibility of mesh resolution effects per rev 2

@@
 were used for the other turbine surfaces, i.e., struts, hub, and shaft, and for
-the tank sidewalls, bottom, and top.
+the tank sidewalls, bottom, and top. This strategy was adopted to reduce the
+overall cell count; since the struts should be operating at nearly zero angle of
+attack, it was assumed that separation would not be a signifigant contributor to
+their parasitic drag.
 
@@
 about 30\%. We also see some smearing of $k$ across the center of the rotor,
-which is likely due to exaggerated levels of the turbulent eddy viscosity.
+which is likely due to exaggerated levels of the turbulent eddy viscosity. This
+could also be an effect of mesh resolution, though for this study only base mesh
+size sensitivity was assessed, rather than local refinement, as this would have
+expanded the parameter space significantly.
 

### 2016-10-02 paper.tex
Clarify what x/D means

@@
 cross-stream and vertical components are superimposed on top of contours of the
-streamwise component at $X/D=1$ in Figure~\ref{fig:br-cfd-mean-velocity}. Both
-CFD models predict the general structure of the mean velocity well, though the
-SA case has a slightly larger vertical mean flow component, which could be due
-to stronger tip vortex generation, or lower diffusivity compared with the SST
-model.
+streamwise component at one rotor diameter downstream ($x/D=1$) of the axis in
+Figure~\ref{fig:br-cfd-mean-velocity}. Both CFD models predict the general
+structure of the mean velocity well, though the SA case has a slightly larger
+vertical mean flow component, which could be due to stronger tip vortex
+generation, or lower diffusivity compared with the SST model.
 

### 2016-10-02 paper.tex
Explain how experimental measurements were taken

@@
 
-        \caption{Mean velocity field at $x/D=1$ from experiments
-            \cite{Bachant2016-RVAT-Re-dep}.}
+        \caption{Mean velocity field at $x/D=1$ from acoustic Doppler
+        velocimetry measurements in a towing
+        tank~\cite{Bachant2016-RVAT-Re-dep}.}
 
@@
 
-        \caption{Turbulence kinetic energy at $x/D=1$ from experiments
-            \cite{Bachant2016-RVAT-Re-dep}.}
+        \caption{Turbulence kinetic energy at $x/D=1$ from acoustic Doppler
+        velocimetry measurements in a towing
+        tank~\cite{Bachant2016-RVAT-Re-dep}.}
 

### 2016-10-02 paper.tex
Remove 'the' per rev 2

@@
 test their abilities to predict turbine performance and near-wake dynamics. It
-was observed that when modeled in 2-D, the performance is over-predicted
-compared to the data from the tow tank experiments, which was expected due to
-omission of blade end effects, support strut drag, and increased blockage.
-Vertical (or axial) wake dynamics were unresolved in the 2-D model, despite
-being identified as a significant contributor to streamwise wake recovery, which
-casts doubt on the 2-D model's applicability as a tool to study array spacing
-effects.
+was observed that when modeled in 2-D, performance is over-predicted compared to
+the data from the tow tank experiments, which was expected due to omission of
+blade end effects, support strut drag, and increased blockage. Vertical (or
+axial) wake dynamics were unresolved in the 2-D model, despite being identified
+as a significant contributor to streamwise wake recovery, which casts doubt on
+the 2-D model's applicability as a tool to study array spacing effects.
 

### 2016-10-02 paper.tex
Do not hyphenate 'overpredict'

@@
 group (RNG) turbulence model. The results from the 2-D simulations
-over-predicted power coefficient since the rotor had a low aspect ratio, while
+overpredicted power coefficient since the rotor had a low aspect ratio, while
 the 3-D case matched well with wind tunnel measurements near the tip speed ratio
@@
 test their abilities to predict turbine performance and near-wake dynamics. It
-was observed that when modeled in 2-D, performance is over-predicted compared to
+was observed that when modeled in 2-D, performance is overpredicted compared to
 the data from the tow tank experiments, which was expected due to omission of

### 2016-10-02 reviews/reviewer2-response.md
Address reviewer 2 comments in abstract
Discuss effect of slip BC
Explain why viscous terms appear to be missing

@@
 >Additionally the title could be better, maybe add a Computational Fluid Dynamics or simulation somewhere?
+
+## Responses to PDF comments
+
+>Effect of this [slip boundary condition] on the results?
+
+It didn't seem to matter much. The mean velocity fields appear to be nearly
+symmetrical about the $x$--$y$ plane, but nonetheless, the slip BC is at least
+closer to how a free surface would behave than a no slip BC.
+
+>Why blank? [viscous transport on in Figure 12]
+
+These terms aren't zero, but are extremely small.

### 2016-10-02 reviews/reviewer2-response.md
Remove extraneous word
Comment on near-blade mesh resolution

@@
 
->Why blank? [viscous transport on in Figure 12]
+>Is this mesh [Figure 4] too coarse for adequate resolution of the flow?
+
+It could be. For our mesh dependence study we held the topology fixed and scaled
+the mesh proportionally. The mesh should be fine enough to resolve the boundary
+layer since $y^+ \sim 1$, but it is a fair point that maybe separated flow
+structures may be dissipated in an exaggerated way. However, exploring the mesh
+topology parameter space would have been prohibitive.
+
+>Why blank? [viscous transport in Figure 12]
 

### 2016-10-02 paper.tex
Clarify N_x parameter sweep

@@
 Meshes were generated using OpenFOAM's blockMesh and snappyHexMesh utilities.
-Mesh topology consists of a background hexahedral mesh, which is refined in all
+Mesh  consists of a background hexahedral mesh, which is refined in all
 three directions by a factor of 2 in a rectangular region containing the turbine
@@
 spacing and time step using 2-D domains. The grid topology was fixed, but the
-number of cells per domain length was scaled proportionally, maintaining the
-same background mesh cell aspect ratio. Results for this parameter sweep are
-shown in Figure~\ref{fig:2d-br-verification}, from which the final number of
-streamwise grid points $N_x = 70$ was chosen, which resulted in a nondimensional
-wall distance $y^+ = u^* y / \nu \sim 1$, where $u^*$ is the friction velocity,
-defined as $u^*=\sqrt{\tau_w / \rho}$, where $\tau_w$ is the wall shear stress.
-This mesh resolution gave a total cell count of approximately $5 \times 10^4$
-for the 2-D cases and 16 million for the 3-D case. Sensitivity to domain length
-was assessed as well, showing a 2\% increase in mean power coefficient for the
-2-D SST case with the domain extended an additional two rotor diameters
-downstream.
+number of cells per domain length $N_x$ was scaled proportionally, maintaining
+the same background mesh cell aspect ratio. Results for the $N_x$ parameter
+sweep are shown in Figure~\ref{fig:2d-br-verification}, from which the final
+number of streamwise grid points $N_x = 70$ was chosen, which resulted in a
+nondimensional wall distance $y^+ = u^* y / \nu \sim 1$, where $u^*$ is the
+friction velocity, defined as $u^*=\sqrt{\tau_w / \rho}$, where $\tau_w$ is the
+wall shear stress. This mesh resolution gave a total cell count of approximately
+$5 \times 10^4$ for the 2-D cases and 16 million for the 3-D case. Sensitivity
+to domain length was assessed as well, showing a 2\% increase in mean power
+coefficient for the 2-D SST case with the domain extended an additional two
+rotor diameters downstream.
 

### 2016-10-02 paper.tex
Define C_P and aspect ratio

@@
 of a high solidity cross-flow turbine using a $k$--$\epsilon$ renormalization
-group (RNG) turbulence model. The results from the 2-D simulations
-overpredicted power coefficient since the rotor had a low aspect ratio, while
-the 3-D case matched well with wind tunnel measurements near the tip speed ratio
-of maximum power. In general, 3-D simulations are less common, but have begun to
-appear more frequently recently---a testament to the progress towards higher
-computing power.
+group (RNG) turbulence model. The results from the 2-D simulations overpredicted
+power coefficient
+\begin{equation}
+    C_P = \frac{P}{\frac{1}{2} \rho A U_\infty^3},
+    \label{eq:cp}
+\end{equation}
+where $P$ is the mechanical shaft power, $\rho$ is the fluid density, $A$ is the
+rotor frontal area, and $U_\infty$ is the free stream velocity. This was likely
+because the rotor had a low aspect ratio (or ratio of height to rotor diameter
+$H/D$), while the 3-D case matched well with wind tunnel measurements near the
+tip speed ratio of maximum power. In general, 3-D simulations are less common,
+but have begun to appear more frequently recently---a testament to the progress
+towards higher computing power.
 

### 2016-10-02 paper.tex
Define rotor drag coefficient

@@
 
-Predictions for both the mean rotor power and drag coefficients at the optimal
+Predictions for both the mean rotor power coefficient $C_P$ and drag coefficient
+\begin{equation}
+    C_D = \frac{F_D}{\frac{1}{2} \rho A U_\infty^2},
+    \label{eq:cd}
+\end{equation}
+where $F_D$ is the total streamwise force acting on the rotor, at the optimal
 tip speed ratio $\lambda_0 = 1.9$ are shown in

### 2016-10-02 paper.tex
Clarify which drag coefficient in figure br-cfd-perf-bar-chart
Explain why we care about TKE

@@
 
-    \caption{Power (left) and drag (right) coefficient predictions at $\lambda =
-    1.9$, from experiments and each numerical model.}
+    \caption{Power (left) and rotor drag (right) coefficient predictions at
+    $\lambda = 1.9$, from experiments and each numerical model.}
 
@@
 
-Turbulence kinetic energy profiles are also shown in
-Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy for the unsteady
-RANS models was calculated as
+Since this is a relatively high Reynolds number flow, it was of interest to
+evaluate turbulence predictions, since these will play an important role in
+overall flow dynamics. Validation was performed against the UNH-RVAT's
+experimental dataset, which was acquired via acoustic Doppler velocimetry. The
+turbulence kinetic energy for the unsteady RANS models was calculated as
 \begin{equation}
@@
 $k_{\mathrm{RA}}$ is the kinetic energy calculated by the turbulence model,
-which is zero for the SA model.
+which is zero for the SA model. Profiles from the simulation results are shown
+in Figure~\ref{fig:br-cfd-profiles}.
 

### 2016-10-02 reviews/reviewer2-response.md
Discuss meshing of walls
Discuss differences in solver time

@@
 
+>So the SA and SST took the same time?
+
+Since the chosen SA time step was about half that of the SST, the SA model did
+take longer to solve, though they were still the same order of magnitude. We
+decided it wouldn't be useful to readers to give very specific metrics of solver
+time since there could be many intervening variables, e.g., hardware.
+
+>Did you prove that the meshing of the outer walls didn't affect the results.
+
+We did not, but if we inspect the velocity fields, which don't extend out all
+the way to the walls, gradients are very small around the outer region,
+indicating the effects should be negligible.
+
 >Why blank? [viscous transport in Figure 12]

### 2016-10-02 paper.tex
Clarify simulated second
Define z and H

@@
 second of simulated time. The 2-D simulations were run on a single processor and
-took on the order of one CPU hour per second.
+took on the order of one CPU hour per simulated second.
 
@@
 
-Mean velocity profiles at one turbine diameter downstream are shown in
+Mean velocity profiles at the mid $x$--$y$ plane (or the nondimensional vertical
+coordinate $z/H=0$) one turbine diameter downstream are shown in
 Figure~\ref{fig:br-cfd-profiles}. The 2-D results suffer from a blockage
@@
     \caption{Mean velocity (left) and turbulence kinetic energy (right) profiles
-        at $x/D=1$ from 2-D simulations, 3-D simulations ($z/H=0$), and experiments
-        \cite{Bachant2015-JoT}.}
+    at $x/D=1$ from 2-D simulations, 3-D simulations (at the mid $x$--$y$ plane,
+    or $z/H=0$), and experiments \cite{Bachant2015-JoT}.}
 

### 2016-10-02 paper.tex
Define y and R

@@
 energy in the flow, since it must be resolved as variance in the velocity field.
-The 2-D SST model did a good job predicting the peak in $k$ at $y/R=-1$, though
-is missing the smaller peak at $y/R=-1$. This is once again likely due to
-blockage issues, where local tip speed ratio is decreased, increasing the
-blades' instantaneous angle of attack at this location on the downstream
-passage. In contrast, the 3-D SST model predicts the $+y$ peak in turbulence
-kinetic energy very well, though the $-y$ peak magnitude is overpredicted by
-about 30\%. We also see some smearing of $k$ across the center of the rotor,
-which is likely due to exaggerated levels of the turbulent eddy viscosity. This
-could also be an effect of mesh resolution, though for this study only base mesh
-size sensitivity was assessed, rather than local refinement, as this would have
-expanded the parameter space significantly.
+The 2-D SST model did a good job predicting the peak in $k$ at $y/R=-1$, where
+$y$ is the cross-stream coordinate and $R$ is the rotor radius, though is
+missing the smaller peak at $y/R=-1$. This is once again likely due to blockage
+issues, where local tip speed ratio is decreased, increasing the blades'
+instantaneous angle of attack at this location on the downstream passage. In
+contrast, the 3-D SST model predicts the $+y$ peak in turbulence kinetic energy
+very well, though the $-y$ peak magnitude is overpredicted by about 30\%. We
+also see some smearing of $k$ across the center of the rotor, which is likely
+due to exaggerated levels of the turbulent eddy viscosity. This could also be an
+effect of mesh resolution, though for this study only base mesh size sensitivity
+was assessed, rather than local refinement, as this would have expanded the
+parameter space significantly.
 

### 2016-10-02 paper.tex
Clarify why we care about wake recovery

@@
 
-To get an overall idea of the wake recovery predicted by each model, we
-rearrange the streamwise component of the Navier--Stokes equation to isolate
-$\partial U / \partial x$---following Bachant and
-Wosnik~\cite{Bachant2015-JoT}---and compute each term at $X/D = 1$ to compare
-with the experimental results.
+To get an overall idea of the wake recovery predicted by each model, and
+therefore its potential effectivness in evaluating array spacing, the streamwise
+component of the Navier--Stokes equation was rearranged to isolate $\partial U /
+\partial x$---following Bachant and Wosnik~\cite{Bachant2015-JoT}---and compute
+each term at $X/D = 1$ to compare with the experimental results.
 

### 2016-10-02 paper.tex
Clarify the vertical component of the mean velocity field

@@
 results---with the SST model overpredicting and SA underpredicting the vertical
-velocity's effect on replenishing streamwise momentum.
+velocity's effect on replenishing streamwise momentum. Nevertheless, both 3-D
+models were able to qualitatively resolve this mean vertical advection caused by
+blade tip vortex shedding, which will be crucial to predicting overall wake
+recovery and the performance of closely spaced turbines.
 
@@
 transport due to the mean pressure gradient and turbulent fluctuations varied a
-lot between the two turbulence models. Both models, however, we able to at least
-qualitatively resolve the vertical velocity field, which will be crucial to
-predicting the performance of closely spaced CFTs.
+lot between the two turbulence models. Both models, however, were able to at
+least qualitatively resolve the vertical component of the near-wake's mean
+velocity field, which will be crucial to predicting the performance of closely
+spaced CFTs.
 

### 2016-10-02 reviews/reviewer2-response.md
Discuss periodicity of rotor performance
Discuss DMS model tip corrections

@@
 
+>why approximately [rotor performance periodicity]?
+
+Since this is an unsteady simulation, there are interactions between blade wakes
+that are not perfectly periodic, so there are some minor differences between
+rotor revolutions. Note that this is consistent with the experimental results.
+
 >Why blank? [viscous transport in Figure 12]
@@
 These terms aren't zero, but are extremely small.
+
+>Not sure if this is true - DMS models have tip correction factors - see reference #2?
+
+We could not find any instances in this reference for DMS model tip correction
+factors for VAWTs. However, if the reviewer has a specific correction in mind we
+would be happy to note this in the manuscript.

### 2016-10-02 scripts/make-perf-ts-fig.py
Add script to make performance time series

@@ -1,0 +1,30 @@
+#!/usr/bin/env python
+"""Create performance coefficient time series plot."""
+
+from __future__ import division, print_function
+from pxl.styleplot import set_sns
+import matplotlib.pyplot as plt
+import pandas as pd
+import sys
+
+
+set_sns(context="paper")
+
+fpath = "UNH-RVAT-3D-OpenFOAM/processed/perf.csv"
+
+df = pd.read_csv(fpath)
+
+fig, ax = plt.subplots()
+
+ax.plot(df.theta_deg, df.cp)
+
+ax.set_xlabel(r"$\theta$ (degrees)")
+ax.set_ylabel("$C_P$")
+ax.set_ylim((0, 0.8))
+
+fig.tight_layout()
+
+fig.savefig("figures/cp-time-series.pdf")
+
+if "--show" in sys.argv:
+    plt.show()

### 2016-10-02 scripts/make-perf-ts-fig.py
Plot revolution averaged C_P for clarity

@@ -6,7 +6,7 @@ from pxl.styleplot import set_sns
 import matplotlib.pyplot as plt
 import pandas as pd
 import sys
-
+import numpy as np
 
 set_sns(context="paper")
 
@@ -14,13 +14,24 @@ fpath = "UNH-RVAT-3D-OpenFOAM/processed/perf.csv"
 
 df = pd.read_csv(fpath)
 
+# Create revolution-averaged data
+nrevs = int(df.theta_deg.max() / 360)
+rev = []
+cp_rev_ave = []
+for n in range(nrevs):
+    rev.append(n + 1)
+    cp_rev_ave.append(
+        df.cp[np.logical_and(df.theta_deg >= n*360,
+                             df.theta_deg < (n + 1)*360)].mean()
+    )
+
 fig, ax = plt.subplots()
 
-ax.plot(df.theta_deg, df.cp)
+ax.plot(rev, cp_rev_ave, marker="o")
 
-ax.set_xlabel(r"$\theta$ (degrees)")
+ax.set_xlabel(r"Rotor revolution")
 ax.set_ylabel("$C_P$")
-ax.set_ylim((0, 0.8))
+# ax.set_ylim((0, 0.8))
 
 fig.tight_layout()
 

### 2016-10-02 reviews/reviewer2-response.md
Discuss difference plots
Discuss number of complete revs

@@
 would be happy to note this in the manuscript.
+
+>Can you show differences between these graphically, i.e subtract B from A and C from A to show the change - this would make it a lot easier to see differences
+
+We did try computing "difference" plots, but these were a bit noisy and were
+difficult to interpret.
+
+>Wouldnt this have to be a number of complete revolutions?
+
+It virtually is--6.03 revolutions technically, but the additional 0.03
+revolutions should not have any appreciable effect on the statistics.

### 2016-10-02 paper.tex
Clarify which formulas indicate which recovery terms

@@
 therefore its potential effectivness in evaluating array spacing, the streamwise
-component of the Navier--Stokes equation was rearranged to isolate $\partial U /
+component of the Navier--Stokes equation was rearranged to isolate the
+streamwise partial derivative of the streamwise mean velocity $\partial U /
 \partial x$---following Bachant and Wosnik~\cite{Bachant2015-JoT}---and compute
-each term at $X/D = 1$ to compare with the experimental results.
+each term at $x/D = 1$ to compare with the experimental results.
 
@@
 number regime, viscous transport is essentially negligible compared with other
-mechanisms. Cross-stream advection---or the tendency of streamlines to diverge
-and reduce the streamwise momentum---produces a negative effect for all cases,
-though the 3-D SST model predicts significantly lower values. Vertical advection
-is by definition zero for the 2-D cases. The 3-D cases show varying
-results---with the SST model overpredicting and SA underpredicting the vertical
-velocity's effect on replenishing streamwise momentum. Nevertheless, both 3-D
-models were able to qualitatively resolve this mean vertical advection caused by
-blade tip vortex shedding, which will be crucial to predicting overall wake
-recovery and the performance of closely spaced turbines.
+mechanisms. Cross-stream advection $-V \frac{\partial U}{\partial y}$; or the
+tendency of streamlines to diverge and reduce the streamwise momentum---produces
+a negative effect for all cases, though the 3-D SST model predicts significantly
+lower values. Mean vertical advection $-W \frac{\partial U}{\partial z}$ is by
+definition zero for the 2-D cases. The 3-D cases show varying results---with the
+SST model overpredicting and SA underpredicting the vertical velocity's effect
+on replenishing streamwise momentum. Nevertheless, both 3-D models were able to
+qualitatively resolve this mean vertical advection caused by blade tip vortex
+shedding, which will be crucial to predicting overall wake recovery and the
+performance of closely spaced turbines.
 
-Turbulent transport and streamwise pressure gradient terms show the largest
-discrepancy between results. The 3-D SST case, despite doing a good job
-predicting turbulence kinetic energy, significantly overpredicts the turbulent
-transport term, while other CFD cases are comparable with the experiment. This
-seems to be balanced by a large adverse pressure gradient, which is also present
-to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
-cases create a wake where the pressure gradient is acting to slightly accelerate
-the flow at $x/D=1$. Unfortunately, pressure data were not acquired from the
-experiment, though results from the 3-D delayed detached eddy simulation (DDES)
-of Boudreau and Dumas \cite{Boudreau2015} concur with the adverse pressure
-gradient condition.
+Turbulent transport and mean streamwise pressure gradient ($-\partial P \partial
+x$) terms show the largest discrepancy between results. The 3-D SST case,
+despite doing a good job predicting turbulence kinetic energy, significantly
+overpredicts the turbulent transport term, while other CFD cases are comparable
+with the experiment. This seems to be balanced by a large adverse pressure
+gradient, which is also present to a smaller degree in the 3-D SA case.
+Interestingly, in contrast, both 2-D CFD cases create a wake where the pressure
+gradient is acting to slightly accelerate the flow at $x/D=1$. Unfortunately,
+pressure data were not acquired from the experiment, though results from the 3-D
+delayed detached eddy simulation (DDES) of Boudreau and Dumas
+\cite{Boudreau2015} concur with the adverse pressure gradient condition.
 

### 2016-10-02 reviews/reviewer2-response.md
Discuss PIV/CFD validation
Update rev 2 response

@@
 
-We thank the reviewer for his or her thoughtful comments, which have helped us
-improve the paper.
+We thank the reviewer for his or her very thorough and thoughtful comments,
+which have helped us improve the paper.
 
@@
 closer to how a free surface would behave than a no slip BC.
+
+>Any studies proving this - PIV works?
+
+To our knowledge, there aren't any studies directly showing that if CFD results
+match well at one point in the domain, they will therefore match elsewhere in
+the domain, but if the reviewer knows of any, we'd be happy to reference them.
 

### 2016-10-09 paper.tex
Discuss strut drag modeling effects

@@
 overall loading or flow modulation. A drawing of the physically and numerically
-modeled geometries is presented in Figure~\ref{fig:cfd-cad}.
+modeled geometries is presented in Figure~\ref{fig:cfd-cad}, and geometric
+parameters are listed in Table~\ref{tab:turb-geom}.
 
@@
 \end{figure}
+
+\begin{table}
+    \centering
+    \begin{tabular}{r|l}
+        Parameter & Value \\
+        \hline
+        Diameter $D$ & 1.0 m \\
+        Height or blade span $H$ & 1.0 m \\
+        Blade chord length $c$ & 0.14 m \\
+        Blade profile & NACA 0020 \\
+        Blade mounting point & $c/2$ and $H/2$ \\
+        Strut chord length & 0.14 m \\
+        Strut profile & NACA 0020
+    \end{tabular}
+
+    \caption{Geometric parameters for the UNH-RVAT rotor modeled in this study.}
+
+    \label{tab:turb-geom}
+\end{table}
 
@@
 Meshes were generated using OpenFOAM's blockMesh and snappyHexMesh utilities.
-Mesh  consists of a background hexahedral mesh, which is refined in all
+The mesh consisted of a background hexahedral mesh, which was refined in all
 three directions by a factor of 2 in a rectangular region containing the turbine
 and near-wake (0.9 m upstream, 1.3 m downstream, $\pm 0.9$ m cross-stream, and
-$\pm 0.8$ m vertically). Cells adjacent to the turbine shaft and struts are
-refined by a factor of 4, while cells adjacent to the blades are refined by a
-factor of 6. To capture the boundary layer, 20 layers were added next to the
-blades with an expansion ratio of 1.2. Overall mesh refinement is controlled by
-a single parameter---the number of cells in the streamwise direction, $N_x$.
+$\pm 0.8$ m vertically). Cells adjacent to the turbine shaft and struts were
+refined by a factor of 4, while cells adjacent to the blades were refined by a
+factor of 6. To capture the boundary layer, i.e., where flow physics most affect
+rotor torque---20 layers were added next to the blades with an expansion ratio
+of 1.2. Overall mesh refinement is controlled by a single parameter---the number
+of cells in the streamwise direction, $N_x$.
 
@@
 final time step of 0.001 s. The results from the SST model show a local minimum
-at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
-minimum was chosen as the final time step to run the simulations. Note that the
-SST model's convergence behavior may be due to its specific implementation in
-OpenFOAM, and not indicative of the nature of the model equations. Verification
-studies for CFTs with this level of detail in the literature are not common,
-though the final time step is comparable to others \cite{Balduzzi2016}. Note
-that while in general there may be a coupling between necessary spatial and
-temporal resolution, these values were arrived at iteratively, i.e., the spatial
-dependence was varied with the chosen time step, and vice versa, such that there
-was one independent variable for each parameter sweep.
+at $\Delta t = 0.002$ s, with some divergence for smaller time steps, the reason
+for which is unknown. The local minimum was chosen as the final time step to run
+the simulations. Note that the SST model's convergence behavior may be due to
+its specific implementation in OpenFOAM, and not indicative of the nature of the
+model equations. Note that while in general there may be a coupling between
+necessary spatial and temporal resolution, these values were arrived at
+iteratively, i.e., the spatial dependence was varied with the chosen time step,
+and vice versa, such that there was one independent variable for each parameter
+sweep. Reassuringly, the chosen time step and average $y^+$ adjacent to the
+blades are comparable to others reported in the literature
+\cite{Maitre2013,Balduzzi2016}.
 
@@
 Turbine operation in all cases was simulated for 10 seconds, or approximately
-six rotor revolutions. Computations for the 3-D cases were run on 192 processes
-(24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
+six rotor revolutions, to allow the revolution-averaged performance coefficient
+to converge to within $\sim 1$ percentage point; cf.
+Figure~\ref{fig:cp-convergence}. Computations for the 3-D cases were run on 192
+processes (24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
 performance computing cluster) and took on the order of 1,000 CPU hours per
@@
 took on the order of one CPU hour per simulated second.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.75\textwidth]{cp-time-series}
+
+    \caption{Revolution-averaged power coefficient versus rotor revolution
+    simulated with the 3-D $k$--$\omega$ SST model.}
+
+    \label{fig:cp-convergence}
+\end{figure}
 
@@
 0.33, which represent a 6\% under- and 30\% overprediction, respectively,
-compared with the experiments. Some of the apparent overprediction of rotor drag
-coefficient could also be an effect of the experimental procedure, where the
-``tare drag'' from the turbine mounting structure was measured without a turbine
-installed, then subtracted in post-processing. With the turbine installed, flow
-past the mounting frame will be higher due to blockage, meaning the tare drag
-would be underestimated, and the CFD results for $C_D$ should be slightly higher
-than the experimental measurements.
+compared with the experiments. Strut drag modeling discrepancies due to the
+struts' lower mesh resolution compared to the blades, along with the use of wall
+functions, could have played a minor role in degrading overall performance
+predictions, and should be investigated further.
+
+Some of the apparent overprediction of rotor drag coefficient could also be an
+effect of the experimental procedure, where the ``tare drag'' from the turbine
+mounting structure was measured without a turbine installed, then subtracted in
+post-processing. With the turbine installed, flow past the mounting frame will
+be higher due to blockage, meaning the tare drag would be underestimated, and
+the CFD results for $C_D$ should be slightly higher than the experimental
+measurements.
 
@@
 effect from the dynamic stall vortex centered around $z/H=0$, which is likely
-more of an issue with wake evolution rather than wake generation.
+more of an issue with wake evolution rather than wake generation. The effects of
+mesh resolution on the breakdown and diffusion of the dynamic stall vortex may
+also have an influence here.
 

### 2016-10-09 paper.tex
High nut causes high turb trans

@@
 despite doing a good job predicting turbulence kinetic energy, significantly
-overpredicts the turbulent transport term, while other CFD cases are comparable
-with the experiment. This seems to be balanced by a large adverse pressure
-gradient, which is also present to a smaller degree in the 3-D SA case.
-Interestingly, in contrast, both 2-D CFD cases create a wake where the pressure
-gradient is acting to slightly accelerate the flow at $x/D=1$. Unfortunately,
-pressure data were not acquired from the experiment, though results from the 3-D
-delayed detached eddy simulation (DDES) of Boudreau and Dumas
-\cite{Boudreau2015} concur with the adverse pressure gradient condition.
+overpredicts the turbulent transport term, thanks to its high levels of
+turbulent viscosity, while other CFD cases are comparable with the experiment.
+This seems to be balanced by a large adverse pressure gradient, which is also
+present to a smaller degree in the 3-D SA case. Interestingly, in contrast, both
+2-D CFD cases create a wake where the pressure gradient is acting to slightly
+accelerate the flow at $x/D=1$. Unfortunately, pressure data were not acquired
+from the experiment, though results from the 3-D delayed detached eddy
+simulation (DDES) of Boudreau and Dumas \cite{Boudreau2015} concur with the
+adverse pressure gradient condition.
 

### 2016-10-09 paper.tex
Discuss reason for assymmetric DS vortex

@@
 with the induced vertical velocity field. The 3-D dynamic stall vortex also
-shows asymmetry about the $x$--$y$ mid-rotor plane; once again highlighting the
-importance of three-dimensional effects on wake dynamics.
+shows slight asymmetry about the $x$--$y$ mid-rotor plane, which was likely
+caused by the slip boundary condition for velocity at the top surface.
 

### 2016-10-09 reviews/reviewer2-response.md
Add to rev 2 response

@@
 
+Specific comments in the PDF have been addressed in the manuscript and in the
+subsection below.
+
 >• Nomenclature, and the lack of definition of many of the variables will cause confusion for many readers
+
+This is a very important point, and we have added many nomenclature definitions
+throughout the text.
 
@@
 
+These transport properties indicate which terms in the governing equations most
+affect wake evolution and its recovery, which will determine rotor spacing in an
+array. A comment has been added to the manuscript to clarify this point.
+
 >Additionally the title could be better, maybe add a Computational Fluid Dynamics or simulation somewhere?
+
+Maybe change "modeling" to "simulating"? It is somewhat implied that "RANS"
+modeling means it is a computational study, but we understand the reviewer's
+point.
 
@@
 revolutions should not have any appreciable effect on the statistics.
+
+>Is this relevant to the paper [the performance curves]?
+
+This is a valid point. If the reviewer thinks the paper would read okay without
+this figure, we would be willing to remove it.
+
+>Why so long [the computational domain]?
+
+We did not perform a mesh dependence study for domain length---only for the
+resolution. Since this is a RANS model, the large scale turbulent structures do
+not advect downstream like they would in an LES model, as the 3-D vorticity
+contours show.
+
+>Or the mesh resolution at 1 diameter downstream [on the difference between the
+3-D SA and SST models' mean velocity]?
+
+This shouldn't be a mesh resolution effect between models since the same mesh
+was used for each.

### 2016-10-09 scripts/diff.sh
Fix up diff script and add Makefile target

@@ -5,8 +5,11 @@
 
 CURRENT=$(git describe --always --dirty)
 rev=$1
-git show $rev:paper.tex > archive/paper-$CURRENT-$rev.tex
-latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$CURRENT-$rev.tex paper.tex > paper-$CURRENT-$rev-diff.tex
-latexmk paper-$CURRENT-$rev-diff.tex -pdf -f
-mv paper-$CURRENT-$rev-diff.pdf archive
-rm paper-$CURRENT-$rev*
+if [ "$rev" = "" ]; then
+    rev=$(git tag | tail -n1)
+fi
+git show $rev:paper.tex > archive/paper-$rev-$CURRENT.tex
+latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$rev-$CURRENT.tex paper.tex > paper-$rev-$CURRENT-diff.tex
+latexmk paper-$rev-$CURRENT-diff.tex -pdf -f
+mv paper-$rev-$CURRENT-diff.pdf archive
+rm paper-$rev-$CURRENT*

### 2016-10-09 scripts/diff.sh
Use diff script from VAT-ALM paper

@@ -1,15 +1,18 @@
 #!/usr/bin/env bash
-# This script creates a diff version comparing the recent with whatever revision
-# is supplied as the first argument. This can be any valid Git identifier, e.g.,
-# a commit hash, tag, etc.
+# This script creates a diff version comparing the current version with the
+# most recent tagged version
 
+PREVIOUS=$(git tag | tail -n1)
 CURRENT=$(git describe --always --dirty)
-rev=$1
-if [ "$rev" = "" ]; then
-    rev=$(git tag | tail -n1)
+
+# If current version is a tag, use one tag prior
+if [ $CURRENT = $PREVIOUS ]
+then
+    PREVIOUS=$(git tag | tail -n2 | head -n1)
 fi
-git show $rev:paper.tex > archive/paper-$rev-$CURRENT.tex
-latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$rev-$CURRENT.tex paper.tex > paper-$rev-$CURRENT-diff.tex
-latexmk paper-$rev-$CURRENT-diff.tex -pdf -f
-mv paper-$rev-$CURRENT-diff.pdf archive
-rm paper-$rev-$CURRENT*
+
+git show $PREVIOUS:paper.tex > archive/paper-$PREVIOUS.tex
+latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$PREVIOUS.tex paper.tex > paper-$PREVIOUS-$CURRENT-diff.tex
+latexmk paper-$PREVIOUS-$CURRENT-diff.tex -pdf -f
+mv paper-$PREVIOUS-$CURRENT-diff.pdf archive
+rm paper-$PREVIOUS-$CURRENT*

### 2016-10-09 reviews/reviewer3-response.md
Add note about strut drag; resolves #43
Update rev 3 response

@@
 
-We appreciate the comments given here, and addressing them has improved the manuscript significantly.
+Firstly, we'd like to thank the reviewer for his or her thoughtful review. We
+appreciate the comments given here, and addressing them has improved the
+manuscript significantly.
 
@@
 However, the authors have not well explained it in the manuscript. Please extend this part.
+
+In the manuscript we have cited a figure from Rawlings (2008, the source of
+experimental data used by Li and Calisal for validation), where support strut
+drag only accounts for a 4 percentage point decrease in $C_P$, which is much
+smaller than the discrepancy observed in the 2-D simulations.

### 2016-10-09 paper.tex
Add labels and more description to 2-D mesh; resolves #44 and resolves #42
Put revised date in header
Add rotor geometry table per rev 2

--- 2016-10-09 paper.tex
@@
 Meshes were generated using OpenFOAM's blockMesh and snappyHexMesh utilities.
-The mesh consisted of a background hexahedral mesh, which was refined in all
+Mesh  consists of a background hexahedral mesh, which is refined in all
 three directions by a factor of 2 in a rectangular region containing the turbine
 and near-wake (0.9 m upstream, 1.3 m downstream, $\pm 0.9$ m cross-stream, and
-$\pm 0.8$ m vertically). Cells adjacent to the turbine shaft and struts were
-refined by a factor of 4, while cells adjacent to the blades were refined by a
-factor of 6. To capture the boundary layer, i.e., where flow physics most affect
-rotor torque---20 layers were added next to the blades with an expansion ratio
-of 1.2. Overall mesh refinement is controlled by a single parameter---the number
-of cells in the streamwise direction, $N_x$.
+$\pm 0.8$ m vertically). Cells adjacent to the turbine shaft and struts are
+refined by a factor of 4, while cells adjacent to the blades are refined by a
+factor of 6. To capture the boundary layer, 20 layers were added next to the
+blades with an expansion ratio of 1.2. Overall mesh refinement is controlled by
+a single parameter---the number of cells in the streamwise direction, $N_x$.
 
@@
 final time step of 0.001 s. The results from the SST model show a local minimum
-at $\Delta t = 0.002$ s, with some divergence for smaller time steps, the reason
-for which is unknown. The local minimum was chosen as the final time step to run
-the simulations. Note that the SST model's convergence behavior may be due to
-its specific implementation in OpenFOAM, and not indicative of the nature of the
-model equations. Note that while in general there may be a coupling between
-necessary spatial and temporal resolution, these values were arrived at
-iteratively, i.e., the spatial dependence was varied with the chosen time step,
-and vice versa, such that there was one independent variable for each parameter
-sweep. Reassuringly, the chosen time step and average $y^+$ adjacent to the
-blades are comparable to others reported in the literature
-\cite{Maitre2013,Balduzzi2016}.
+at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
+minimum was chosen as the final time step to run the simulations. Note that the
+SST model's convergence behavior may be due to its specific implementation in
+OpenFOAM, and not indicative of the nature of the model equations. Verification
+studies for CFTs with this level of detail in the literature are not common,
+though the final time step is comparable to others \cite{Balduzzi2016}. Note
+that while in general there may be a coupling between necessary spatial and
+temporal resolution, these values were arrived at iteratively, i.e., the spatial
+dependence was varied with the chosen time step, and vice versa, such that there
+was one independent variable for each parameter sweep.
 
@@
 Turbine operation in all cases was simulated for 10 seconds, or approximately
-six rotor revolutions, to allow the revolution-averaged performance coefficient
-to converge to within $\sim 1$ percentage point; cf.
-Figure~\ref{fig:cp-convergence}. Computations for the 3-D cases were run on 192
-processes (24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
+six rotor revolutions. Computations for the 3-D cases were run on 192 processes
+(24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
 performance computing cluster) and took on the order of 1,000 CPU hours per
@@
 took on the order of one CPU hour per simulated second.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.75\textwidth]{cp-time-series}
-
-    \caption{Revolution-averaged power coefficient versus rotor revolution
-    simulated with the 3-D $k$--$\omega$ SST model.}
-
-    \label{fig:cp-convergence}
-\end{figure}
 
@@
 0.33, which represent a 6\% under- and 30\% overprediction, respectively,
-compared with the experiments. Strut drag modeling discrepancies due to the
-struts' lower mesh resolution compared to the blades, along with the use of wall
-functions, could have played a minor role in degrading overall performance
-predictions, and should be investigated further.
-
-Some of the apparent overprediction of rotor drag coefficient could also be an
-effect of the experimental procedure, where the ``tare drag'' from the turbine
-mounting structure was measured without a turbine installed, then subtracted in
-post-processing. With the turbine installed, flow past the mounting frame will
-be higher due to blockage, meaning the tare drag would be underestimated, and
-the CFD results for $C_D$ should be slightly higher than the experimental
-measurements.
+compared with the experiments. Some of the apparent overprediction of rotor drag
+coefficient could also be an effect of the experimental procedure, where the
+``tare drag'' from the turbine mounting structure was measured without a turbine
+installed, then subtracted in post-processing. With the turbine installed, flow
+past the mounting frame will be higher due to blockage, meaning the tare drag
+would be underestimated, and the CFD results for $C_D$ should be slightly higher
+than the experimental measurements.
 
@@
 effect from the dynamic stall vortex centered around $z/H=0$, which is likely
-more of an issue with wake evolution rather than wake generation. The effects of
-mesh resolution on the breakdown and diffusion of the dynamic stall vortex may
-also have an influence here.
+more of an issue with wake evolution rather than wake generation.
 

### 2016-10-09 reviews/reviewer2-response.md
Add labels and more description to 2-D mesh; resolves #44 and resolves #42
Add responses to rev 2

--- 2016-10-02 reviews/reviewer2-response.md
@@
 revolutions should not have any appreciable effect on the statistics.
+
+>Is this relevant to the paper [the performance curves]?
+
+This is a valid point. If the reviewer thinks the paper would read okay without
+this figure, we would be willing to remove it.
+
+>Why so long [the computational domain]?
+
+We did not perform a mesh dependence study for domain length---only for the
+resolution. Since this is a RANS model, the large scale turbulent structures do
+not advect downstream like they would in an LES model, as the 3-D vorticity
+contours show.

### 2016-10-09 paper.tex
Discuss mesh resolution per rev 2
SST divergence for smaller time steps is unknown

@@
 final time step of 0.001 s. The results from the SST model show a local minimum
-at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
-minimum was chosen as the final time step to run the simulations. Note that the
-SST model's convergence behavior may be due to its specific implementation in
-OpenFOAM, and not indicative of the nature of the model equations. Verification
-studies for CFTs with this level of detail in the literature are not common,
-though the final time step is comparable to others \cite{Balduzzi2016}. Note
-that while in general there may be a coupling between necessary spatial and
-temporal resolution, these values were arrived at iteratively, i.e., the spatial
-dependence was varied with the chosen time step, and vice versa, such that there
-was one independent variable for each parameter sweep.
+at $\Delta t = 0.002$ s, with some divergence for smaller time steps, the reason
+for which is unknown. The local minimum was chosen as the final time step to run
+the simulations. Note that the SST model's convergence behavior may be due to
+its specific implementation in OpenFOAM, and not indicative of the nature of the
+model equations. Verification studies for CFTs with this level of detail in the
+literature are not common, though the final time step is comparable to others
+\cite{Balduzzi2016}. Note that while in general there may be a coupling between
+necessary spatial and temporal resolution, these values were arrived at
+iteratively, i.e., the spatial dependence was varied with the chosen time step,
+and vice versa, such that there was one independent variable for each parameter
+sweep.
 

### 2016-10-09 paper.tex
Add citation for another verification study
Explain length of simulations

@@
 its specific implementation in OpenFOAM, and not indicative of the nature of the
-model equations. Verification studies for CFTs with this level of detail in the
-literature are not common, though the final time step is comparable to others
-\cite{Balduzzi2016}. Note that while in general there may be a coupling between
+model equations. Note that while in general there may be a coupling between
 necessary spatial and temporal resolution, these values were arrived at
@@
 and vice versa, such that there was one independent variable for each parameter
-sweep.
+sweep. Reassuringly, the chosen time step and average $y^+$ adjacent to the
+blades are comparable to others reported in the literature
+\cite{Maitre2013,Balduzzi2016}.
 
@@
 Turbine operation in all cases was simulated for 10 seconds, or approximately
-six rotor revolutions. Computations for the 3-D cases were run on 192 processes
-(24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
-performance computing cluster) and took on the order of 1,000 CPU hours per
-second of simulated time. The 2-D simulations were run on a single processor and
-took on the order of one CPU hour per simulated second.
+six rotor revolutions, to allow the revolution-averaged performance coefficient
+to converge to within $\sim 1$ percentage point. Computations for the 3-D cases
+were run on 192 processes (24 nodes $\times$ 8 cores each on Sandia National
+Labs' Red Mesa high performance computing cluster) and took on the order of
+1,000 CPU hours per second of simulated time. The 2-D simulations were run on a
+single processor and took on the order of one CPU hour per simulated second.
 

### 2016-10-09 paper.tex
Add C_P convergence plot

@@
 six rotor revolutions, to allow the revolution-averaged performance coefficient
-to converge to within $\sim 1$ percentage point. Computations for the 3-D cases
-were run on 192 processes (24 nodes $\times$ 8 cores each on Sandia National
-Labs' Red Mesa high performance computing cluster) and took on the order of
-1,000 CPU hours per second of simulated time. The 2-D simulations were run on a
-single processor and took on the order of one CPU hour per simulated second.
+to converge to within $\sim 1$ percentage point; cf.
+Figure~\ref{fig:cp-convergence}. Computations for the 3-D cases were run on 192
+processes (24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
+performance computing cluster) and took on the order of 1,000 CPU hours per
+second of simulated time. The 2-D simulations were run on a single processor and
+took on the order of one CPU hour per simulated second.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.75\textwidth]{cp-time-series}
+
+    \caption{Revolution-averaged power coefficient versus rotor revolution
+    simulated with the 3-D $k$--$\omega$ SST model.}
+
+    \label{fig:cp-convergence}
+\end{figure}
 

### 2016-10-09 reviews/reviewer2-response.md
Discuss mesh resolution per rev 2
Add to rev 2 response

@@
 >Additionally the title could be better, maybe add a Computational Fluid Dynamics or simulation somewhere?
+
+Maybe change "modeling" to "simulating"? It is somewhat implied that "RANS"
+modeling means it is a computational study, but we understand the reviewer's
+point.
 
@@
 contours show.
+
+>Or the mesh resolution at 1 diameter downstream [on the difference between the
+3-D SA and SST models' mean velocity]?
+
+This shouldn't be a mesh resolution effect between models since the same mesh
+was used for each.
